

Statistical analyses for 2024 Cornell ponds floating solar experiment


Analyses (each in it's own chunk):
1. Pond temp and DO over the summer
2. Testing for stratification using pond temperature data
3. Pond dissolved CO2 and CH4 data throughout the summer
4. Diffusive CH4 fluxes
5. Sediment fluxes
6. Microbial data

Load necessary libraries
```{r}
library(readxl)
library(lubridate)
library(tidyverse)
library(MBA)
library(reshape2)
library(ggplot2)
library(ggtext)
library(patchwork)
library(lme4)
library(emmeans)
```



1. Start with the basic temp/do plots
```{r}
wc_temp_do <- read_excel("C:/Users/nray1/Dropbox/Cornell/Floating Solar Project/Intensive CH4 Day/Manuscript/Response to Review 2/fs_methane_intensive_2024 - 2026_07_05 version.xlsx", 
    sheet = "ysi_dat")
View(wc_temp_do)

str(wc_temp_do)
wc_temp_do$Date = as.Date(ymd(wc_temp_do$Date))
wc_temp_do$Pond = as.factor(wc_temp_do$Pond)
str(wc_temp_do)

#make the color palette
temp_colors = c("#b2182b", "#ef8a62", "#fddbc7", "#f7f7f7", "#d1e5f0", "#67a9cf", "#2166ac")

##make the temperature plots
min(wc_temp_do$Temp_deg_C)
max(wc_temp_do$Temp_deg_C)

#start with p123
#get the data ready
p123 = filter(wc_temp_do, Pond == "123")
p123$Date = as.Date(ymd(p123$Date))
p123_plot = p123 %>%
  mutate(depth = - `Depth_m`)
p123_plot$Date = as.Date(p123_plot$Date)
p123_temp_plot_dat = p123_plot[,c("doy", "depth", "Temp_deg_C")]
p123_temp = mba.surf(p123_temp_plot_dat[c("doy", "depth", "Temp_deg_C")], no.X = 300, no.Y = 300, extend = T)
dimnames(p123_temp$xyz.est$z) = list(p123_temp$xyz.est$x, p123_temp$xyz.est$y)
p123_temp = melt(p123_temp$xyz.est$z, varnames = c('doy', 'depth'), value.name = "Temp_deg_C")

#graph it
p123_temp_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p123_temp, aes(x = doy, y = depth, fill = Temp_deg_C))+
  scale_fill_gradientn("Water Temp<br>(\u00b0 C)",colours = rev(temp_colors), limits = c(13.5,30))+
  geom_point(data = p123_temp_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 123 - with FPV")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p123_temp_plot

#repeat for pond 124
p124 = filter(wc_temp_do, Pond == "124")
p124$Date = as.Date(ymd(p124$Date))
p124_plot = p124 %>%
  mutate(depth = - `Depth_m`)
p124_plot$Date = as.Date(p124_plot$Date)
p124_temp_plot_dat = p124_plot[,c("doy", "depth", "Temp_deg_C")]
p124_temp = mba.surf(p124_temp_plot_dat[c("doy", "depth", "Temp_deg_C")], no.X = 300, no.Y = 300, extend = T)
dimnames(p124_temp$xyz.est$z) = list(p124_temp$xyz.est$x, p124_temp$xyz.est$y)
p124_temp = melt(p124_temp$xyz.est$z, varnames = c('doy', 'depth'), value.name = "Temp_deg_C")
p124_temp_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p124_temp, aes(x = doy, y = depth, fill = Temp_deg_C))+
  scale_fill_gradientn("Water Temp<br>(\u00b0 C)",colours = rev(temp_colors), limits = c(13.5,30))+
  geom_point(data = p124_temp_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 124 - with FPV")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p124_temp_plot

#pond 125
p125 = filter(wc_temp_do, Pond == "125")
p125$Date = as.Date(ymd(p125$Date))
p125_plot = p125 %>%
  mutate(depth = - `Depth_m`)
p125_plot$Date = as.Date(p125_plot$Date)
p125_temp_plot_dat = p125_plot[,c("doy", "depth", "Temp_deg_C")]
p125_temp = mba.surf(p125_temp_plot_dat[c("doy", "depth", "Temp_deg_C")], no.X = 300, no.Y = 300, extend = T)
dimnames(p125_temp$xyz.est$z) = list(p125_temp$xyz.est$x, p125_temp$xyz.est$y)
p125_temp = melt(p125_temp$xyz.est$z, varnames = c('doy', 'depth'), value.name = "Temp_deg_C")
p125_temp_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p125_temp, aes(x = doy, y = depth, fill = Temp_deg_C))+
  scale_fill_gradientn("Water Temp<br>(\u00b0 C)",colours = rev(temp_colors), limits = c(13.5,30))+
  geom_point(data = p125_temp_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 125 - with FPV")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p125_temp_plot

#pond 128
p128 = filter(wc_temp_do, Pond == "128")
p128$Date = as.Date(ymd(p128$Date))
p128_plot = p128 %>%
  mutate(depth = - `Depth_m`)
p128_plot$Date = as.Date(p128_plot$Date)
p128_temp_plot_dat = p128_plot[,c("doy", "depth", "Temp_deg_C")]
p128_temp = mba.surf(p128_temp_plot_dat[c("doy", "depth", "Temp_deg_C")], no.X = 300, no.Y = 300, extend = T)
dimnames(p128_temp$xyz.est$z) = list(p128_temp$xyz.est$x, p128_temp$xyz.est$y)
p128_temp = melt(p128_temp$xyz.est$z, varnames = c('doy', 'depth'), value.name = "Temp_deg_C")
p128_temp_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p128_temp, aes(x = doy, y = depth, fill = Temp_deg_C))+
  scale_fill_gradientn("Water Temp<br>(\u00b0 C)",colours = rev(temp_colors), limits = c(13.5,30))+
  geom_point(data = p128_temp_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 128 - open water")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p128_temp_plot

#pond 131
p131 = filter(wc_temp_do, Pond == "131")
p131$Date = as.Date(ymd(p131$Date))
p131_plot = p131 %>%
  mutate(depth = - `Depth_m`)
p131_plot$Date = as.Date(p131_plot$Date)
p131_temp_plot_dat = p131_plot[,c("doy", "depth", "Temp_deg_C")]
p131_temp = mba.surf(p131_temp_plot_dat[c("doy", "depth", "Temp_deg_C")], no.X = 300, no.Y = 300, extend = T)
dimnames(p131_temp$xyz.est$z) = list(p131_temp$xyz.est$x, p131_temp$xyz.est$y)
p131_temp = melt(p131_temp$xyz.est$z, varnames = c('doy', 'depth'), value.name = "Temp_deg_C")
p131_temp_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p131_temp, aes(x = doy, y = depth, fill = Temp_deg_C))+
  scale_fill_gradientn("Water Temp<br>(\u00b0 C)",colours = rev(temp_colors), limits = c(13.5,30))+
  geom_point(data = p131_temp_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 131 - open water")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p131_temp_plot

#pond 132
p132 = filter(wc_temp_do, Pond == "132")
p132$Date = as.Date(ymd(p132$Date))
p132_plot = p132 %>%
  mutate(depth = - `Depth_m`)
p132_plot$Date = as.Date(p132_plot$Date)
p132_temp_plot_dat = p132_plot[,c("doy", "depth", "Temp_deg_C")]
p132_temp = mba.surf(p132_temp_plot_dat[c("doy", "depth", "Temp_deg_C")], no.X = 300, no.Y = 300, extend = T)
dimnames(p132_temp$xyz.est$z) = list(p132_temp$xyz.est$x, p132_temp$xyz.est$y)
p132_temp = melt(p132_temp$xyz.est$z, varnames = c('doy', 'depth'), value.name = "Temp_deg_C")
p132_temp_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p132_temp, aes(x = doy, y = depth, fill = Temp_deg_C))+
  scale_fill_gradientn("Water Temp<br>(\u00b0 C)",colours = rev(temp_colors), limits = c(13.5,30))+
  geom_point(data = p132_temp_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 132 - open water")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p132_temp_plot

#put the plots together
temp_profile_plots = p123_temp_plot + p124_temp_plot + p125_temp_plot + p128_temp_plot + p131_temp_plot + p132_temp_plot +
  plot_layout(nrow = 2, ncol = 3,
              guides = "collect")
temp_profile_plots

temp_profile_plots[[1]] = temp_profile_plots[[1]] + theme(axis.title.x = element_blank(), axis.text.x = element_blank())
temp_profile_plots[[2]] = temp_profile_plots[[2]] + theme(axis.title.x = element_blank(), axis.text.x = element_blank(),
                                                                            axis.title.y = element_blank(), axis.text.y = element_blank())
temp_profile_plots[[3]] = temp_profile_plots[[3]] + theme(axis.title.x = element_blank(), axis.text.x = element_blank(),
                                                                            axis.title.y = element_blank(), axis.text.y = element_blank())
temp_profile_plots[[5]] = temp_profile_plots[[5]] + theme(axis.title.y = element_blank(), axis.text.y = element_blank())
temp_profile_plots[[6]] = temp_profile_plots[[6]] + theme(axis.title.y = element_blank(), axis.text.y = element_blank())
temp_profile_plots

#save it as figure 1
ggsave(filename = "figure_1_temp_profiles.png", plot = temp_profile_plots, width = 6.5, height = 4.3, units = "in")

#############
#repeat for dissolved oxygen
min(wc_temp_do$DO_PercSat)
max(wc_temp_do$DO_PercSat)

#pond123
p123_do_plot_dat = p123_plot[,c("doy", "depth", "DO_PercSat")]
p123_do = mba.surf(p123_do_plot_dat[c("doy", "depth", "DO_PercSat")], no.X = 300, no.Y = 300, extend = T)
dimnames(p123_do$xyz.est$z) = list(p123_do$xyz.est$x, p123_do$xyz.est$y)
p123_do = melt(p123_do$xyz.est$z, varnames = c('doy', 'depth'), value.name = "DO_PercSat")
p123_do_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p123_do, aes(x = doy, y = depth, fill = DO_PercSat))+
  scale_fill_gradientn("Oxygen<br>Saturation (%)",colours = rev(temp_colors), limits = c(0,200))+
  geom_point(data = p123_do_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 123 - with FPV")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p123_do_plot

#pond124
p124_do_plot_dat = p124_plot[,c("doy", "depth", "DO_PercSat")]
p124_do = mba.surf(p124_do_plot_dat[c("doy", "depth", "DO_PercSat")], no.X = 300, no.Y = 300, extend = T)
dimnames(p124_do$xyz.est$z) = list(p124_do$xyz.est$x, p124_do$xyz.est$y)
p124_do = melt(p124_do$xyz.est$z, varnames = c('doy', 'depth'), value.name = "DO_PercSat")
p124_do_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p124_do, aes(x = doy, y = depth, fill = DO_PercSat))+
  scale_fill_gradientn("Oxygen<br>Saturation (%)",colours = rev(temp_colors), limits = c(0,200))+
  geom_point(data = p124_do_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 124 - with FPV")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p124_do_plot

#pond125
p125_do_plot_dat = p125_plot[,c("doy", "depth", "DO_PercSat")]
p125_do = mba.surf(p125_do_plot_dat[c("doy", "depth", "DO_PercSat")], no.X = 300, no.Y = 300, extend = T)
dimnames(p125_do$xyz.est$z) = list(p125_do$xyz.est$x, p125_do$xyz.est$y)
p125_do = melt(p125_do$xyz.est$z, varnames = c('doy', 'depth'), value.name = "DO_PercSat")
p125_do_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p125_do, aes(x = doy, y = depth, fill = DO_PercSat))+
  scale_fill_gradientn("Oxygen<br>Saturation (%)",colours = rev(temp_colors), limits = c(0,200))+
  geom_point(data = p125_do_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 125 - with FPV")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p125_do_plot

#pond128
p128_do_plot_dat = p128_plot[,c("doy", "depth", "DO_PercSat")]
p128_do = mba.surf(p128_do_plot_dat[c("doy", "depth", "DO_PercSat")], no.X = 300, no.Y = 300, extend = T)
dimnames(p128_do$xyz.est$z) = list(p128_do$xyz.est$x, p128_do$xyz.est$y)
p128_do = melt(p128_do$xyz.est$z, varnames = c('doy', 'depth'), value.name = "DO_PercSat")
p128_do_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p128_do, aes(x = doy, y = depth, fill = DO_PercSat))+
  scale_fill_gradientn("Oxygen<br>Saturation (%)",colours = rev(temp_colors), limits = c(0,200))+
  geom_point(data = p128_do_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 128 - open water")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p128_do_plot

#pond131
p131_do_plot_dat = p131_plot[,c("doy", "depth", "DO_PercSat")]
p131_do = mba.surf(p131_do_plot_dat[c("doy", "depth", "DO_PercSat")], no.X = 300, no.Y = 300, extend = T)
dimnames(p131_do$xyz.est$z) = list(p131_do$xyz.est$x, p131_do$xyz.est$y)
p131_do = melt(p131_do$xyz.est$z, varnames = c('doy', 'depth'), value.name = "DO_PercSat")
p131_do$DO_PercSat[p131_do$DO_PercSat < 0] = 0
p131_do_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p131_do, aes(x = doy, y = depth, fill = DO_PercSat))+
  scale_fill_gradientn("Oxygen<br>Saturation (%)",colours = rev(temp_colors), limits = c(0,200))+
  geom_point(data = p131_do_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 131 - open water")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p131_do_plot

#pond132
p132_do_plot_dat = p132_plot[,c("doy", "depth", "DO_PercSat")]
p132_do = mba.surf(p132_do_plot_dat[c("doy", "depth", "DO_PercSat")], no.X = 300, no.Y = 300, extend = T)
dimnames(p132_do$xyz.est$z) = list(p132_do$xyz.est$x, p132_do$xyz.est$y)
p132_do = melt(p132_do$xyz.est$z, varnames = c('doy', 'depth'), value.name = "DO_PercSat")
p132_do$DO_PercSat[p132_do$DO_PercSat < 0] = 0
p132_do_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p132_do, aes(x = doy, y = depth, fill = DO_PercSat))+
  scale_fill_gradientn("Oxygen<br>Saturation (%)",colours = rev(temp_colors), limits = c(0,200))+
  geom_point(data = p132_do_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 132 - open water")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p132_do_plot

do_profile_plots = p123_do_plot + p124_do_plot + p125_do_plot + p128_do_plot + p131_do_plot + p132_do_plot +
  plot_layout(nrow = 2, ncol = 3,
              guides = "collect")
do_profile_plots

do_profile_plots[[1]] = do_profile_plots[[1]] + theme(axis.title.x = element_blank(), axis.text.x = element_blank())
do_profile_plots[[2]] = do_profile_plots[[2]] + theme(axis.title.x = element_blank(), axis.text.x = element_blank(),
                                                                            axis.title.y = element_blank(), axis.text.y = element_blank())
do_profile_plots[[3]] = do_profile_plots[[3]] + theme(axis.title.x = element_blank(), axis.text.x = element_blank(),
                                                                            axis.title.y = element_blank(), axis.text.y = element_blank())
do_profile_plots[[5]] = do_profile_plots[[5]] + theme(axis.title.y = element_blank(), axis.text.y = element_blank())
do_profile_plots[[6]] = do_profile_plots[[6]] + theme(axis.title.y = element_blank(), axis.text.y = element_blank())
do_profile_plots

#save it as figure 2
ggsave(filename = "figure_2_dissolved_oxygen.png", plot = do_profile_plots, width = 6.5, height = 4.3, units = "in")

########
#Do a quick comparison of average temp and DO by treatment

#get the daily averages by pond
temp_do_av_dat = wc_temp_do %>%
  group_by(doy, Pond, treatment) %>%
  dplyr::summarize(mean_temp = mean(Temp_deg_C, na.rm = TRUE),
                   mean_do = mean(DO_PercSat, na.rm = TRUE),
                   mean_do_mg_L = mean(DO_mg_L, na.rm = TRUE))

#and then do the comparisons
pond_temp_model = lmer(mean_temp ~ treatment + doy + (1|Pond), data = temp_do_av_dat)
summary(pond_temp_model)
lsmeans(pond_temp_model, pairwise ~ treatment)

pond_do_model = lmer(mean_do ~ treatment + doy + (1|Pond), data = temp_do_av_dat)
summary(pond_do_model)
lsmeans(pond_do_model, pairwise ~ treatment)

pond_domgL_model = lmer(mean_do_mg_L ~ treatment + doy + (1|Pond), data = temp_do_av_dat)
summary(pond_domgL_model)
lsmeans(pond_domgL_model, pairwise ~ treatment)

#get the means and SDs
temp_do_means = temp_do_av_dat %>%
  group_by(treatment) %>%
  dplyr::summarize(sd_temp = sd(mean_temp, na.rm = TRUE),
                   max_temp = max(mean_temp),
                   min_temp = min(mean_temp),
                   mean_temp = mean(mean_temp),
                   sd_do = sd(mean_do),
                   max_do = max(mean_do),
                   min_do = min(mean_do),
                   mean_do = mean(mean_do),
                   sd_do_mg_L = sd(mean_do_mg_L),
                   max_do_mg_L = max(mean_do_mg_L),
                   min_do_mg_L = min(mean_do_mg_L),
                   mean_do_mg_L = mean(mean_do_mg_L)
                   )
temp_do_means

#did FPV ponds ever have DO > 100%?
fpv_do = subset(wc_temp_do, treatment == "fpv")
max(fpv_do$DO_PercSat)

##############
## and graphs of DO in mg L-1
min(wc_temp_do$DO_mg_L) #0.11
max(wc_temp_do$DO_mg_L) #13.15

#pond123-125
p123_domg_plot_dat = p123_plot[,c("doy", "depth", "DO_mg_L")]
p123_do = mba.surf(p123_domg_plot_dat[c("doy", "depth", "DO_mg_L")], no.X = 300, no.Y = 300, extend = T)
dimnames(p123_do$xyz.est$z) = list(p123_do$xyz.est$x, p123_do$xyz.est$y)
p123_do = melt(p123_do$xyz.est$z, varnames = c('doy', 'depth'), value.name = "DO_mg_L")
min(p123_do$DO_mg_L)

p123_domg_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p123_do, aes(x = doy, y = depth, fill = DO_mg_L))+
  scale_fill_gradientn("Dissolved Oxygen<br>Concentration (mg L<sup>-1</sup>)",colours = rev(temp_colors), limits = c(0,14))+
  geom_point(data = p123_domg_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 123 - with FPV")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p123_domg_plot

p124_domg_plot_dat = p124_plot[,c("doy", "depth", "DO_mg_L")]
p124_do = mba.surf(p124_domg_plot_dat[c("doy", "depth", "DO_mg_L")], no.X = 300, no.Y = 300, extend = T)
dimnames(p124_do$xyz.est$z) = list(p124_do$xyz.est$x, p124_do$xyz.est$y)
p124_do = melt(p124_do$xyz.est$z, varnames = c('doy', 'depth'), value.name = "DO_mg_L")
min(p124_do$DO_mg_L)

p124_domg_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p124_do, aes(x = doy, y = depth, fill = DO_mg_L))+
  scale_fill_gradientn("Dissolved Oxygen<br>Concentration (mg L<sup>-1</sup>)",colours = rev(temp_colors), limits = c(0,14))+
  geom_point(data = p124_domg_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 124 - with FPV")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p124_domg_plot

p125_domg_plot_dat = p125_plot[,c("doy", "depth", "DO_mg_L")]
p125_do = mba.surf(p125_domg_plot_dat[c("doy", "depth", "DO_mg_L")], no.X = 300, no.Y = 300, extend = T)
dimnames(p125_do$xyz.est$z) = list(p125_do$xyz.est$x, p125_do$xyz.est$y)
p125_do = melt(p125_do$xyz.est$z, varnames = c('doy', 'depth'), value.name = "DO_mg_L")
min(p125_do$DO_mg_L)

p125_domg_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p125_do, aes(x = doy, y = depth, fill = DO_mg_L))+
  scale_fill_gradientn("Dissolved Oxygen<br>Concentration (mg L<sup>-1</sup>)",colours = rev(temp_colors), limits = c(0,14))+
  geom_point(data = p125_domg_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 125 - with FPV")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p125_domg_plot

#p128, 131, and 132
p128_domg_plot_dat = p128_plot[,c("doy", "depth", "DO_mg_L")]
p128_do = mba.surf(p128_domg_plot_dat[c("doy", "depth", "DO_mg_L")], no.X = 300, no.Y = 300, extend = T)
dimnames(p128_do$xyz.est$z) = list(p128_do$xyz.est$x, p128_do$xyz.est$y)
p128_do = melt(p128_do$xyz.est$z, varnames = c('doy', 'depth'), value.name = "DO_mg_L")
min(p128_do$DO_mg_L)

p128_domg_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p128_do, aes(x = doy, y = depth, fill = DO_mg_L))+
  scale_fill_gradientn("Dissolved Oxygen<br>Concentration (mg L<sup>-1</sup>)",colours = rev(temp_colors), limits = c(0,14))+
  geom_point(data = p128_domg_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 128 - open water")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p128_domg_plot

p131_domg_plot_dat = p131_plot[,c("doy", "depth", "DO_mg_L")]
p131_do = mba.surf(p131_domg_plot_dat[c("doy", "depth", "DO_mg_L")], no.X = 300, no.Y = 300, extend = T)
dimnames(p131_do$xyz.est$z) = list(p131_do$xyz.est$x, p131_do$xyz.est$y)
p131_do = melt(p131_do$xyz.est$z, varnames = c('doy', 'depth'), value.name = "DO_mg_L")
min(p131_do$DO_mg_L)
p131_do$DO_mg_L[p131_do$DO_mg_L < 0] = 0
min(p131_do$DO_mg_L)

p131_domg_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p131_do, aes(x = doy, y = depth, fill = DO_mg_L))+
  scale_fill_gradientn("Dissolved Oxygen<br>Concentration (mg L<sup>-1</sup>)",colours = rev(temp_colors), limits = c(0,14))+
  geom_point(data = p131_domg_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 131 - open water")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p131_domg_plot

p132_domg_plot_dat = p132_plot[,c("doy", "depth", "DO_mg_L")]
p132_do = mba.surf(p132_domg_plot_dat[c("doy", "depth", "DO_mg_L")], no.X = 300, no.Y = 300, extend = T)
dimnames(p132_do$xyz.est$z) = list(p132_do$xyz.est$x, p132_do$xyz.est$y)
p132_do = melt(p132_do$xyz.est$z, varnames = c('doy', 'depth'), value.name = "DO_mg_L")
min(p132_do$DO_mg_L)
p132_do$DO_mg_L[p132_do$DO_mg_L < 0] = 0
min(p132_do$DO_mg_L)


p132_domg_plot = ggplot()+
  theme_classic()+
  geom_raster(data = p132_do, aes(x = doy, y = depth, fill = DO_mg_L))+
  scale_fill_gradientn("Dissolved Oxygen<br>Concentration (mg L<sup>-1</sup>)",colours = rev(temp_colors), limits = c(0,14))+
  geom_point(data = p132_domg_plot_dat, aes(x = doy, y = depth), size = 0.5)+
  scale_y_continuous(limits = c(-1.5,0), breaks = c(-1.5, -1, -0.5, 0))+
  theme(panel.border = element_rect(fill = NA, colour = "black"))+
  labs(y = "Depth (m)", x = ("Day of Year"), title = "Pond 132 - open water")+
  theme(axis.title.y = element_markdown(size = 8,colour = "black"), 
        axis.title.x = element_markdown(size = 8,colour = "black"),
        axis.text.x =  element_text(size = 8,colour = "black"),
        axis.text.y =  element_text(size = 8,colour = "black"),
        plot.title = element_text(size = 8,colour = "black"),
        legend.title = element_markdown(size = 8,colour = "black"),
        legend.text = element_text(size = 8,colour = "black"))
p132_domg_plot

domg_profile_plots = p123_domg_plot + p124_domg_plot + p125_domg_plot + p128_domg_plot + p131_domg_plot + p132_domg_plot +
  plot_layout(nrow = 2, ncol = 3,
              guides = "collect")
domg_profile_plots

domg_profile_plots[[1]] = domg_profile_plots[[1]] + theme(axis.title.x = element_blank(), axis.text.x = element_blank())
domg_profile_plots[[2]] = domg_profile_plots[[2]] + theme(axis.title.x = element_blank(), axis.text.x = element_blank(),
                                                                            axis.title.y = element_blank(), axis.text.y = element_blank())
domg_profile_plots[[3]] = domg_profile_plots[[3]] + theme(axis.title.x = element_blank(), axis.text.x = element_blank(),
                                                                            axis.title.y = element_blank(), axis.text.y = element_blank())
domg_profile_plots[[5]] = domg_profile_plots[[5]] + theme(axis.title.y = element_blank(), axis.text.y = element_blank())
domg_profile_plots[[6]] = domg_profile_plots[[6]] + theme(axis.title.y = element_blank(), axis.text.y = element_blank())
domg_profile_plots

ggsave(filename = "figure_2_dissolved_oxygen_mgL.png", plot = domg_profile_plots, width = 6.5, height = 4.3, units = "in")


```

2. Use the temp data to test for stratification
```{r}
#quick visual inspection
plot_temp1 <- wc_temp_do  %>% 
  ggplot(aes(x=Temp_deg_C, y=Depth_m, color=Pond)) + 
  geom_point() + geom_line()+
  scale_y_reverse()+
  theme(text = element_text(size=12))+ 
  xlab("Temp C") + ylab("Depth (m)") + 
  scale_color_manual(values=c("blue", "blue", "blue", 
                              "goldenrod", "goldenrod", "goldenrod")) + 
  theme_bw()
plot_temp1 + facet_wrap(~Date)

#visually, stratification early summer, disappears later in summer...

# calculate density for each sampling point...
density_dat = wc_temp_do
density_dat$density = (999.83952 + 16.945176*density_dat$Temp_deg_C - (0.0079870401 * (density_dat$Temp_deg_C^2)) +
  (0.000046170461*(density_dat$Temp_deg_C^3)) + (0.00000010556302*(density_dat$Temp_deg_C^4)) - 
    (0.00000000028054253*(density_dat$Temp_deg_C^5)))/(1 + (0.016897850*density_dat$Temp_deg_C))
#this is following Kell (1975) and is kg m-3

#make pairwise comparisons of density gradients between adjacent sampling depths in each pond
pairwise_grad_dat <- density_dat %>%
  arrange(doy, Pond, Depth_m) %>%
  group_by(doy, Pond, treatment) %>%
  mutate(
    depth_upper = Depth_m,
    depth_lower = lead(Depth_m),
    rho_upper = density,
    rho_lower = lead(density),
    density_gradient = (rho_lower - rho_upper) /
      (depth_lower - depth_upper)
  ) %>%
  filter(!is.na(depth_lower)) %>%
  select(
    doy, Pond, treatment,
    depth_upper, depth_lower,
    rho_upper, rho_lower,
    density_gradient
  ) %>%
  ungroup()

pairwise_grad_dat

#identify the strongest gradients
max_grad_dat <- pairwise_grad_dat %>%
  group_by(doy, Pond, treatment) %>%
  slice_max(density_gradient, n = 1, with_ties = FALSE) %>%
  ungroup()
max_grad_dat

#compare strongest gradient between treatments
max_grad_dat$Pond = as.factor(max_grad_dat$Pond)
dens_grad_mod = lmer(density_gradient ~  treatment + doy + (1|Pond), data = max_grad_dat)
summary(dens_grad_mod)
lsmeans(dens_grad_mod, pairwise ~ treatment)

fpv_strat = subset(max_grad_dat, treatment == "fpv")
mean(fpv_strat$density_gradient)
sd(fpv_strat$density_gradient)

open_strat = subset(max_grad_dat, treatment == "open")
mean(open_strat$density_gradient)
sd(open_strat$density_gradient)

day_one_strat = subset(max_grad_dat, doy == "172")
mean(day_one_strat$density_gradient)
sd(day_one_strat$density_gradient)

finalday_strat = subset(max_grad_dat, doy == "255")
mean(finalday_strat$density_gradient)
sd(finalday_strat$density_gradient)

#get the summaries by treatment for graphing
av_max_grad = max_grad_dat %>%
  group_by(doy, treatment)%>%
  dplyr::summarize(mean_grad = mean(density_gradient),
                   sd_grad = sd(density_gradient),
                   mean_depth = mean((depth_lower + depth_upper)/2),
                   sd_depth = sd((depth_lower + depth_upper)/2))

#max density plot
max_dens_plot = ggplot(data = av_max_grad, aes(x = doy, y = mean_grad, fill = treatment, color = treatment))+
  geom_hline(yintercept = 0.287, linetype = "dashed")+
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = doy, ymin = mean_grad - sd_grad, ymax = mean_grad + sd_grad), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  labs(y= "Maximum Density Gradient (kg m<sup>-3</sup> m<sup>-1</sup>)", x = "Day of Year")+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_text(size = 8,colour = "black"),
        axis.title.y = element_markdown(size = 8))+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("#ef8a62", "#67a9cf"))+
  theme(legend.position = "NULL")+
  scale_y_continuous(limits = c(0,2.5), breaks = c(0, 0.5, 1, 1.5, 2, 2.5))+
  scale_x_continuous(limits = c(170,255))
max_dens_plot

#density gradient depth plot
grad_depth_plot = ggplot(data = av_max_grad, aes(x = doy, y = mean_depth, fill = treatment, color = treatment))+
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = doy, ymin = mean_depth - sd_depth, ymax = mean_depth + sd_depth), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  labs(y= "Depth of Maximum Density Gradient (m)", x = "Day of Year")+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_text(size = 8,colour = "black"),
        axis.title.y = element_markdown(size = 8))+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("#ef8a62", "#67a9cf"))+
  theme(legend.position = "NULL")+
  scale_y_reverse(limits = c(0,1.5), breaks = c(0, 0.5, 1, 1.5))+
  scale_x_continuous(limits = c(170,255))

grad_depth_plot

#put the two graphs together:
strat_plots = max_dens_plot + grad_depth_plot +
  plot_layout(nrow = 2, ncol = 1)+
  plot_annotation(tag_levels = "A", tag_suffix = '.')& 
  theme(plot.tag = element_text(size = 8))
strat_plots

ggsave(filename = "strat_plots.png", plot = strat_plots, width = 4.5, height = 6.5, units = "in")
```



3. Compare dissolved gas concentrations throughout the summer
```{r}
diss_ch4_dat<- read_excel("fs_methane_intensive_2024_v3.xlsx", sheet = "diss_ch4_dat")
View(diss_ch4_dat)

#get means of technical replicates
treatment_mean_diss = diss_ch4_dat %>%
  group_by(treatment, doy, pond, top_bottom) %>%
  dplyr::summarize(mean_ch4 = mean(ch4_umol_L))

#and split out surface and bottom water
top_ch4_conc_dat = subset(treatment_mean_diss, top_bottom == "Top")
bottom_ch4_conc_dat = subset(treatment_mean_diss, top_bottom == "Bottom")

#then make the top averages
top_mean_diss = top_ch4_conc_dat %>%
  group_by(treatment, doy) %>%
  dplyr::summarize(sd_ch4 = sd(mean_ch4),
                   bigmean_ch4 = mean(mean_ch4))

#now make the graphs for the surface water:
max(top_ch4_conc_dat$mean_ch4) #10
top_ch4_fig = ggplot(data = top_mean_diss, aes(x = doy, y = bigmean_ch4, fill = treatment, color = treatment))+
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = doy, ymin = bigmean_ch4 - sd_ch4, ymax = bigmean_ch4 + sd_ch4), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  labs(y= "Surface [CH<sub>4</sub>]<br> (\u03bcmol L<sup>-1</sup>)", x = "Day of Year")+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("#ef8a62", "#67a9cf"))+
  theme(legend.position = "NULL")+
  scale_y_continuous(limits = c(0,11), breaks = c(0, 5, 10))+
  scale_x_continuous(limits = c(170,255))
top_ch4_fig

#compare the treatments
top_ch4_conc_dat$pond = as.factor(top_ch4_conc_dat$pond)
top_ch4_model = lmer(mean_ch4 ~ treatment + doy + (1|pond), data = top_ch4_conc_dat)
summary(top_ch4_model)
lsmeans(top_ch4_model, pairwise ~ treatment)

#and make the comparison graph
max(top_ch4_conc_dat$mean_ch4)
top_ch4_box = ggplot()+
  theme_classic()+
  geom_boxplot(data = top_ch4_conc_dat, aes(x = treatment, y = mean_ch4), outlier.shape = NA, color = "black")+
  geom_jitter(data = top_ch4_conc_dat, aes(x = treatment, y = mean_ch4, fill = treatment, color = treatment),width = 0.2, size = 2, shape = 21)+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("black", "black"))+
  scale_y_continuous(limits = c(0,11), breaks = c(0, 5, 10))+
  scale_x_discrete(labels = c("open" = "Open",
                              "fpv" = "FPV"))+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  theme(legend.position = "NULL")+
  annotate("text", x = 2, y = 11, label = "p = 0.208", size = 2.822,  fontface = "italic")
top_ch4_box


########
#repeat for the bottom water
#then make the bottom averages
bottom_mean_diss = bottom_ch4_conc_dat %>%
  group_by(treatment, doy) %>%
  dplyr::summarize(sd_ch4 = sd(mean_ch4),
                   bigmean_ch4 = mean(mean_ch4))

#now make the graphs for the bottom water:
max(bottom_ch4_conc_dat$mean_ch4) #1203
bottom_ch4_fig = ggplot(data = bottom_mean_diss, aes(x = doy, y = bigmean_ch4, fill = treatment, color = treatment))+
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = doy, ymin = bigmean_ch4 - sd_ch4, ymax = bigmean_ch4 + sd_ch4), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  labs(y= "Bottom [CH<sub>4</sub>]<br> (\u03bcmol L<sup>-1</sup>)", x = "Day of Year")+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_text(size = 8,colour = "black"),
        axis.title.y = element_markdown(size = 8))+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("#ef8a62", "#67a9cf"))+
  theme(legend.position = "NULL")+
  scale_y_continuous(limits = c(0,1250), breaks = c(0, 500, 1000))+
  scale_x_continuous(limits = c(170,255))
bottom_ch4_fig

#compare the treatments
bottom_ch4_conc_dat$pond = as.factor(bottom_ch4_conc_dat$pond)
bottom_ch4_model = lmer(mean_ch4 ~ treatment + doy + (1|pond), data = bottom_ch4_conc_dat)
summary(bottom_ch4_model)
lsmeans(bottom_ch4_model, pairwise ~ treatment)

#and make the comparison graph
max(bottom_ch4_conc_dat$mean_ch4)
bottom_ch4_box = ggplot()+
  theme_classic()+
  geom_boxplot(data = bottom_ch4_conc_dat, aes(x = treatment, y = mean_ch4), outlier.shape = NA, color = "black")+
  geom_jitter(data = bottom_ch4_conc_dat, aes(x = treatment, y = mean_ch4, fill = treatment, color = treatment),width = 0.2, size = 2, shape = 21)+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("black", "black"))+
  scale_y_continuous(limits = c(0,1250), breaks = c(0, 500, 1000))+
  scale_x_discrete(labels = c("open" = "Open",
                              "fpv" = "FPV"))+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  theme(legend.position = "NULL")+
  annotate("text", x = 2, y = 1250, label = "p = 0.003", size = 2.822,  fontface = "italic")
bottom_ch4_box

##add in plots showing surface and bottom water DO concentrations
do_dat = wc_temp_do %>%
  group_by(doy, Pond, treatment) %>%
  summarize(
    do_surface = DO_mg_L[Depth_m == min(Depth_m, na.rm = TRUE)],
    do_bottom  = DO_mg_L[Depth_m == max(Depth_m, na.rm = TRUE)],
    .groups = "drop"
  )

#make the box plots:
max(do_dat$do_surface) #10
surface_o2_box = ggplot()+
  theme_classic()+
  geom_boxplot(data = do_dat, aes(x = treatment, y = do_surface), outlier.shape = NA, color = "black")+
  geom_jitter(data = do_dat, aes(x = treatment, y = do_surface, fill = treatment, color = treatment),width = 0.2, size = 2, shape = 21)+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("black", "black"))+
  scale_y_continuous(limits = c(0,10), breaks = c(0, 5, 10))+
  scale_x_discrete(labels = c("open" = "Open",
                              "fpv" = "FPV"))+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  theme(legend.position = "NULL")
  annotate("text", x = 2, y = 1250, label = "p = 0.003", size = 2.822,  fontface = "italic")
surface_o2_box

bottom_o2_box = ggplot()+
  theme_classic()+
  geom_boxplot(data = do_dat, aes(x = treatment, y = do_bottom), outlier.shape = NA, color = "black")+
  geom_jitter(data = do_dat, aes(x = treatment, y = do_bottom, fill = treatment, color = treatment),width = 0.2, size = 2, shape = 21)+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("black", "black"))+
  scale_y_continuous(limits = c(0,10), breaks = c(0, 5, 10))+
  scale_x_discrete(labels = c("open" = "Open",
                              "fpv" = "FPV"))+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  theme(legend.position = "NULL")
  annotate("text", x = 2, y = 1250, label = "p = 0.003", size = 2.822,  fontface = "italic")
bottom_o2_box

#and the time series
do_mean = do_dat %>%
  group_by(doy, treatment) %>%
  summarize(surface_mean = mean(do_surface),
            surface_sd = sd(do_surface),
            bottom_mean = mean(do_bottom),
            bottom_sd = sd(do_bottom))

surface_O2_fig = ggplot(data = do_mean, aes(x = doy, y = surface_mean, fill = treatment, color = treatment))+
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = doy, ymin = surface_mean - surface_sd, ymax = surface_mean + surface_sd), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  labs(y= "Surface DO<br>(mg L<sup>-1</sup>)", x = "Day of Year")+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_text(size = 8,colour = "black"),
        axis.title.y = element_markdown(size = 8))+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("#ef8a62", "#67a9cf"))+
  theme(legend.position = "NULL")+
  scale_y_continuous(limits = c(0,10), breaks = c(0, 5, 10))+
  scale_x_continuous(limits = c(170,255))
surface_O2_fig

bottom_O2_fig = ggplot(data = do_mean, aes(x = doy, y = bottom_mean, fill = treatment, color = treatment))+
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = doy, ymin = bottom_mean - bottom_sd, ymax = bottom_mean + bottom_sd), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  labs(y= "Bottom DO<br>(mg L<sup>-1</sup>)", x = "Day of Year")+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_text(size = 8,colour = "black"),
        axis.title.y = element_markdown(size = 8))+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("#ef8a62", "#67a9cf"))+
  theme(legend.position = "NULL")+
  scale_y_continuous(limits = c(0,10), breaks = c(0, 5, 10))+
  scale_x_continuous(limits = c(170,255))
bottom_O2_fig

#put all the graphs together (Figure 3)
figure3_plot = top_ch4_fig + top_ch4_box + bottom_ch4_fig + bottom_ch4_box + surface_O2_fig + surface_o2_box + bottom_O2_fig + bottom_o2_box +
  plot_layout(nrow = 4, ncol = 2, widths = c(10, 5))+
  plot_annotation(tag_levels = "A", tag_suffix = '.')& 
  theme(plot.tag = element_text(size = 8))
figure3_plot

figure3_plot[[1]] = figure3_plot[[1]] + theme(axis.title.x = element_blank())
figure3_plot[[3]] = figure3_plot[[3]] + theme(axis.title.x = element_blank())
figure3_plot[[5]] = figure3_plot[[5]] + theme(axis.title.x = element_blank())
figure3_plot[[2]] = figure3_plot[[2]] + theme(axis.text.y = element_blank(),
                                        axis.title.y = element_blank(),
                                        axis.line.y = element_blank(),
                                        axis.ticks.y = element_blank(),
                                        axis.title.x = element_blank())
figure3_plot[[4]] = figure3_plot[[4]] + theme(axis.text.y = element_blank(),
                                        axis.title.y = element_blank(),
                                        axis.line.y = element_blank(),
                                        axis.ticks.y = element_blank(),
                                        axis.title.x = element_blank())
figure3_plot[[6]] = figure3_plot[[6]] + theme(axis.text.y = element_blank(),
                                        axis.title.y = element_blank(),
                                        axis.line.y = element_blank(),
                                        axis.ticks.y = element_blank(),
                                        axis.title.x = element_blank())
figure3_plot[[8]] = figure3_plot[[8]] + theme(axis.text.y = element_blank(),
                                        axis.title.y = element_blank(),
                                        axis.line.y = element_blank(),
                                        axis.ticks.y = element_blank(),
                                        axis.title.x = element_blank())
figure3_plot

##export the figure
ggsave(filename = "figure_3_dissolved_ch4.png", plot = figure3_plot, width = 6.5, height = 7, units = "in")



##get the surface and bottom water average and SD by treatment
diss_ch4_avs = treatment_mean_diss %>%
  group_by(treatment, top_bottom) %>%
  dplyr::summarize(ch4_sd = sd(mean_ch4),
                   big_mean_ch4 = mean(mean_ch4))
diss_ch4_avs

```


4. Compare diffusive fluxes 
```{r}
diff_flux_dat = read_excel("fs_methane_intensive_2024_v3.xlsx", sheet = "diff_flux_dat")
view(diff_flux_dat)

diff_flux_dat$center_edge = as.factor(diff_flux_dat$center_edge)

#separate out the center and edge
center_diff_dat = subset(diff_flux_dat, center_edge == "Center")
edge_diff_dat = subset(diff_flux_dat, center_edge == "Edge")

t.test(ch4_flux_umol_m2_h ~ treatment, data = center_diff_dat)
t.test(ch4_flux_umol_m2_h ~ treatment, data = edge_diff_dat)

#and compare center and edge within treatments
fpv_diff_dat = subset(diff_flux_dat, treatment == "fpv")
open_diff_dat = subset(diff_flux_dat, treatment == "open")

t.test(ch4_flux_umol_m2_h ~ center_edge, data = fpv_diff_dat)
t.test(ch4_flux_umol_m2_h ~ center_edge, data = open_diff_dat)



pond_diff_av = diff_flux_dat %>%
  group_by(pond, treatment) %>%
  dplyr::summarize(mean_ch4_flux = mean(ch4_flux_umol_m2_h),
                   sd_ch4_flux = sd(ch4_flux_umol_m2_h))
pond_diff_av  
mean(pond_diff_av$mean_ch4_flux)
sd(pond_diff_av$mean_ch4_flux)


pond_diff_av2 = diff_flux_dat %>%
  group_by(center_edge, treatment) %>%
  dplyr::summarize(mean_ch4_flux = mean(ch4_flux_umol_m2_h),
                   sd_ch4_flux = sd(ch4_flux_umol_m2_h))
pond_diff_av2
```


5. Compare the sediment CH4 bottle incubation data
```{r}
sed_ch4_dat = read_excel("fs_methane_intensive_2024_v3.xlsx", sheet = "sed_flux_dat")

t.test(ch4_ppm_d_gDW ~ treatment, data = sed_ch4_dat)

sed_ch4_summary = sed_ch4_dat %>%
  group_by(treatment) %>%
  dplyr::summarize(mean_flux = mean(ch4_ppm_d_gDW),
                   sd_flux = sd(ch4_ppm_d_gDW))
sed_ch4_summary
```




6. Plot methanogen and methanotroph abundances over time
```{r}
#first load the watercolumn data
fpv_wc_microb_dat = read_excel("fs_methane_intensive_2024_v3.xlsx", sheet = "wc_microbe")

wc_microb_avs = fpv_wc_microb_dat %>%
  group_by(treatment, top_bottom, methanogen_methanotroph) %>%
  dplyr::summarize(meth_sd = sd(total_cells_ml),
                   meth_mean = mean(total_cells_ml))
wc_microb_avs

#split out surface and bottom water and methogen/methanotroph
top_methanogen_dat = fpv_wc_microb_dat[fpv_wc_microb_dat$top_bottom == "top" & fpv_wc_microb_dat$methanogen_methanotroph == "Methanogen",]
top_methanotroph_dat = fpv_wc_microb_dat[fpv_wc_microb_dat$top_bottom == "top" & fpv_wc_microb_dat$methanogen_methanotroph == "Methanotroph",]
bottom_methanogen_dat = fpv_wc_microb_dat[fpv_wc_microb_dat$top_bottom == "bottom" & fpv_wc_microb_dat$methanogen_methanotroph == "Methanogen",]
bottom_methanotroph_dat = fpv_wc_microb_dat[fpv_wc_microb_dat$top_bottom == "bottom" & fpv_wc_microb_dat$methanogen_methanotroph == "Methanotroph",]

#start with surface methanogens
top_gen_sum = top_methanogen_dat %>%
  group_by(treatment, doy) %>%
  dplyr::summarize(sd_meth = sd(total_cells_ml),
                   mean_meth = mean(total_cells_ml))

max(top_gen_sum$mean_meth)
top_methanogen_fig = ggplot(data = top_gen_sum, aes(x = doy, y = mean_meth, fill = treatment, color = treatment))+
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = doy, ymin = mean_meth - sd_meth, ymax = mean_meth + sd_meth), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  labs(y= "Surface Methanogen<br>Abundance (Cells mL<sup>-1</sup>)", x = "Day of Year")+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("#ef8a62", "#67a9cf"))+
  theme(legend.position = "NULL")#+
  #scale_y_continuous(limits = c(0,11), breaks = c(0, 5, 10))+
  #scale_x_continuous(limits = c(170,250))
top_methanogen_fig

top_methanogen_dat$Pond = as.factor(top_methanogen_dat$Pond)
top_methanogen_model = lmer(total_cells_ml ~ treatment + doy + (1|Pond), data = top_methanogen_dat)
summary(top_methanogen_model)
lsmeans(top_methanogen_model, pairwise ~ treatment)

top_methanogen_box = ggplot()+
  theme_classic()+
  geom_boxplot(data = top_methanogen_dat, aes(x = treatment, y = total_cells_ml), outlier.shape = NA, color = "black")+
  geom_jitter(data = top_methanogen_dat, aes(x = treatment, y = total_cells_ml, fill = treatment, color = treatment),width = 0.2, size = 2, shape = 21)+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("black", "black"))+
  #scale_y_continuous(limits = c(0,11), breaks = c(0, 5, 10))+
  scale_x_discrete(labels = c("open" = "Open",
                              "fpv" = "FPV"))+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  theme(legend.position = "NULL")+
  annotate("text", x = 2, y = 9000, label = "p = 0.324", size = 2.822,  fontface = "italic")
top_methanogen_box


##surface methanotrophs
top_troph_sum = top_methanotroph_dat %>%
  group_by(treatment, doy) %>%
  dplyr::summarize(sd_meth = sd(total_cells_ml),
                   mean_meth = mean(total_cells_ml))

max(top_troph_sum$mean_meth)
top_methanotroph_fig = ggplot(data = top_troph_sum, aes(x = doy, y = mean_meth, fill = treatment, color = treatment))+
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = doy, ymin = mean_meth - sd_meth, ymax = mean_meth + sd_meth), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  labs(y= "Surface Methanotroph<br>Abundance (Cells mL<sup>-1</sup>)", x = "Day of Year")+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("#ef8a62", "#67a9cf"))+
  theme(legend.position = "NULL")#+
  #scale_y_continuous(limits = c(0,11), breaks = c(0, 5, 10))+
  #scale_x_continuous(limits = c(170,250))
top_methanotroph_fig

top_methanotroph_dat$Pond = as.factor(top_methanotroph_dat$Pond)
top_methanotroph_model = lmer(total_cells_ml ~ treatment + doy + (1|Pond), data = top_methanotroph_dat)
summary(top_methanotroph_model)
lsmeans(top_methanotroph_model, pairwise ~ treatment)

top_methanotroph_box = ggplot()+
  theme_classic()+
  geom_boxplot(data = top_methanotroph_dat, aes(x = treatment, y = total_cells_ml), outlier.shape = NA, color = "black")+
  geom_jitter(data = top_methanotroph_dat, aes(x = treatment, y = total_cells_ml, fill = treatment, color = treatment),width = 0.2, size = 2, shape = 21)+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("black", "black"))+
  #scale_y_continuous(limits = c(0,11), breaks = c(0, 5, 10))+
  scale_x_discrete(labels = c("open" = "Open",
                              "fpv" = "FPV"))+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  theme(legend.position = "NULL")+
  annotate("text", x = 2, y = 1500000, label = "p < 0.001", size = 2.822,  fontface = "italic")
top_methanotroph_box

####repeat for the bottom waters
bottom_gen_sum = bottom_methanogen_dat %>%
  group_by(treatment, doy) %>%
  dplyr::summarize(sd_meth = sd(total_cells_ml),
                   mean_meth = mean(total_cells_ml))

max(bottom_gen_sum$mean_meth)
bottom_methanogen_fig = ggplot(data = bottom_gen_sum, aes(x = doy, y = mean_meth, fill = treatment, color = treatment))+
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = doy, ymin = mean_meth - sd_meth, ymax = mean_meth + sd_meth), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  labs(y= "Bottom Methanogen<br>Abundance (Cells mL<sup>-1</sup>)", x = "Day of Year")+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("#ef8a62", "#67a9cf"))+
  theme(legend.position = "NULL")#+
  #scale_y_continuous(limits = c(0,11), breaks = c(0, 5, 10))+
  #scale_x_continuous(limits = c(170,250))
bottom_methanogen_fig

bottom_methanogen_dat$Pond = as.factor(bottom_methanogen_dat$Pond)
bottom_methanogen_model = lmer(total_cells_ml ~ treatment + doy + (1|Pond), data = bottom_methanogen_dat)
summary(bottom_methanogen_model)
lsmeans(bottom_methanogen_model, pairwise ~ treatment)

bottom_methanogen_box = ggplot()+
  theme_classic()+
  geom_boxplot(data = bottom_methanogen_dat, aes(x = treatment, y = total_cells_ml), outlier.shape = NA, color = "black")+
  geom_jitter(data = bottom_methanogen_dat, aes(x = treatment, y = total_cells_ml, fill = treatment, color = treatment),width = 0.2, size = 2, shape = 21)+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("black", "black"))+
  #scale_y_continuous(limits = c(0,11), breaks = c(0, 5, 10))+
  scale_x_discrete(labels = c("open" = "Open",
                              "fpv" = "FPV"))+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  theme(legend.position = "NULL")+
  annotate("text", x = 2, y = 75000, label = "p = 0.572", size = 2.822,  fontface = "italic")
bottom_methanogen_box

bottom_troph_sum = bottom_methanotroph_dat %>%
  group_by(treatment, doy) %>%
  dplyr::summarize(sd_meth = sd(total_cells_ml),
                   mean_meth = mean(total_cells_ml))

max(bottom_troph_sum$mean_meth)
bottom_methanotroph_fig = ggplot(data = bottom_troph_sum, aes(x = doy, y = mean_meth, fill = treatment, color = treatment))+
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = doy, ymin = mean_meth - sd_meth, ymax = mean_meth + sd_meth), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  labs(y= "Bottom Methanotroph<br>Abundance (Cells mL<sup>-1</sup>)", x = "Day of Year")+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("#ef8a62", "#67a9cf"))+
  theme(legend.position = "NULL")#+
  #scale_y_continuous(limits = c(0,11), breaks = c(0, 5, 10))+
  #scale_x_continuous(limits = c(170,250))
bottom_methanotroph_fig

bottom_methanotroph_dat$Pond = as.factor(bottom_methanotroph_dat$Pond)
bottom_methanotroph_model = lmer(total_cells_ml ~ treatment + doy + (1|Pond), data = bottom_methanotroph_dat)
summary(bottom_methanotroph_model)
lsmeans(bottom_methanotroph_model, pairwise ~ treatment)

bottom_methanotroph_box = ggplot()+
  theme_classic()+
  geom_boxplot(data = bottom_methanotroph_dat, aes(x = treatment, y = total_cells_ml), outlier.shape = NA, color = "black")+
  geom_jitter(data = bottom_methanotroph_dat, aes(x = treatment, y = total_cells_ml, fill = treatment, color = treatment),width = 0.2, size = 2, shape = 21)+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("black", "black"))+
  #scale_y_continuous(limits = c(0,11), breaks = c(0, 5, 10))+
  scale_x_discrete(labels = c("open" = "Open",
                              "fpv" = "FPV"))+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  theme(legend.position = "NULL")+
  annotate("text", x = 2, y = 750000, label = "p < 0.001", size = 2.822,  fontface = "italic")
bottom_methanotroph_box

#####and now the sediments
fpv_sed_microb_dat = read_excel("fs_methane_intensive_2024_v3.xlsx", sheet = "sed_microbe")

sed_troph_dat = subset(fpv_sed_microb_dat, methanogen_methanotroph == "Methanotroph")
sed_gen_dat = subset(fpv_sed_microb_dat, methanogen_methanotroph == "Methanogen")

sed_troph_sum = sed_troph_dat %>%
  group_by(treatment, doy) %>%
  dplyr::summarize(sd_meth = sd(order_abund_relative),
                   mean_meth = mean(order_abund_relative))

max(sed_troph_sum$mean_meth)
sed_methanotroph_fig = ggplot(data = sed_troph_sum, aes(x = doy, y = mean_meth, fill = treatment, color = treatment))+
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = doy, ymin = mean_meth - sd_meth, ymax = mean_meth + sd_meth), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  labs(y= "Sediment Methanotroph<br>Abundance (Proportion)", x = "Day of Year")+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("#ef8a62", "#67a9cf"))+
  theme(legend.position = "NULL")#+
  #scale_y_continuous(limits = c(0,11), breaks = c(0, 5, 10))+
  #scale_x_continuous(limits = c(170,250))
sed_methanotroph_fig

sed_troph_dat$Pond = as.factor(sed_troph_dat$Pond)
sed_methanotroph_model = lmer(order_abund_relative ~ treatment + doy + (1|Pond), data = sed_troph_dat)
summary(sed_methanotroph_model)
lsmeans(sed_methanotroph_model, pairwise ~ treatment)

sed_methanotroph_box = ggplot()+
  theme_classic()+
  geom_boxplot(data = sed_troph_dat, aes(x = treatment, y = order_abund_relative), outlier.shape = NA, color = "black")+
  geom_jitter(data = sed_troph_dat, aes(x = treatment, y = order_abund_relative, fill = treatment, color = treatment),width = 0.2, size = 2, shape = 21)+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("black", "black"))+
  #scale_y_continuous(limits = c(0,11), breaks = c(0, 5, 10))+
  scale_x_discrete(labels = c("open" = "Open",
                              "fpv" = "FPV"))+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  theme(legend.position = "NULL")+
  annotate("text", x = 2, y = .15, label = "p = 0.736", size = 2.822,  fontface = "italic")
sed_methanotroph_box

###
sed_gen_sum = sed_gen_dat %>%
  group_by(treatment, doy) %>%
  dplyr::summarize(sd_meth = sd(order_abund_relative),
                   mean_meth = mean(order_abund_relative))

max(sed_gen_sum$mean_meth)
sed_methanogen_fig = ggplot(data = sed_gen_sum, aes(x = doy, y = mean_meth, fill = treatment, color = treatment))+
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = doy, ymin = mean_meth - sd_meth, ymax = mean_meth + sd_meth), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  labs(y= "Sediment Methanogen<br>Abundance (Proportion)", x = "Day of Year")+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_text(size = 8,colour = "black"),
        axis.title.y = element_markdown(size = 8))+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("#ef8a62", "#67a9cf"))+
  theme(legend.position = "NULL")#+
  #scale_y_continuous(limits = c(0,11), breaks = c(0, 5, 10))+
  #scale_x_continuous(limits = c(170,250))
sed_methanogen_fig

sed_gen_dat$Pond = as.factor(sed_gen_dat$Pond)
sed_methanogen_model = lmer(order_abund_relative ~ treatment + doy + (1|Pond), data = sed_gen_dat)
summary(sed_methanogen_model)
lsmeans(sed_methanogen_model, pairwise ~ treatment)

sed_methanogen_box = ggplot()+
  theme_classic()+
  geom_boxplot(data = sed_gen_dat, aes(x = treatment, y = order_abund_relative), outlier.shape = NA, color = "black")+
  geom_jitter(data = sed_gen_dat, aes(x = treatment, y = order_abund_relative, fill = treatment, color = treatment),width = 0.2, size = 2, shape = 21)+
  scale_fill_manual(values = c("#ef8a62", "#67a9cf"))+
  scale_color_manual(values = c("black", "black"))+
  #scale_y_continuous(limits = c(0,11), breaks = c(0, 5, 10))+
  scale_x_discrete(labels = c("open" = "Open",
                              "fpv" = "FPV"))+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  theme(legend.position = "NULL")+
  annotate("text", x = 2, y = .15, label = "p = 0.654", size = 2.822,  fontface = "italic")
sed_methanogen_box

##put them all together
microbe_time_plot = top_methanotroph_fig + top_methanotroph_box + top_methanogen_fig + top_methanogen_box +
  bottom_methanotroph_fig + bottom_methanotroph_box + bottom_methanogen_fig + bottom_methanogen_box + 
  sed_methanotroph_fig + sed_methanotroph_box + sed_methanogen_fig + sed_methanogen_box+
  plot_layout(nrow = 6, ncol = 2, widths = c(10, 5))+
  plot_annotation(tag_levels = "A", tag_suffix = '.')& 
  theme(plot.tag = element_text(size = 8))
  
microbe_time_plot[[2]] = microbe_time_plot[[2]] + theme(axis.title.y = element_blank(), axis.line.y = element_blank(), axis.ticks.y = element_blank(), axis.text.y = element_blank())
microbe_time_plot[[4]] = microbe_time_plot[[4]] + theme(axis.title.y = element_blank(), axis.line.y = element_blank(), axis.ticks.y = element_blank(), axis.text.y = element_blank())
microbe_time_plot[[6]] = microbe_time_plot[[6]] + theme(axis.title.y = element_blank(), axis.line.y = element_blank(), axis.ticks.y = element_blank(), axis.text.y = element_blank())
microbe_time_plot[[8]] = microbe_time_plot[[8]] + theme(axis.title.y = element_blank(), axis.line.y = element_blank(), axis.ticks.y = element_blank(), axis.text.y = element_blank())
microbe_time_plot[[10]] = microbe_time_plot[[10]] + theme(axis.title.y = element_blank(), axis.line.y = element_blank(), axis.ticks.y = element_blank(), axis.text.y = element_blank())
microbe_time_plot[[12]] = microbe_time_plot[[12]] + theme(axis.title.y = element_blank(), axis.line.y = element_blank(), axis.ticks.y = element_blank(), axis.text.y = element_blank())

microbe_time_plot

#export the figure
ggsave(filename = "microbe_summary_plot.png", plot = microbe_time_plot, width = 6.5, height = 8, units = "in")
```



