---
title: "Water Column and Sediment Methanogens & Methanotrophs in FPV Ponds"
author: "Sophia Aredas & Mar Schmidt"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    highlight: default
    keep_md: yes
    theme: journal
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

# Purpose

We will create a figure to show the difference in methanogens and
methanotrophs in FSP and control ponds at different depths and between
sample types. 

Additionally, I have written the methods and results in this document
which will be added to the google document.

```{r setup, include=FALSE}
# For width of code chunks and scroll bar 
options(width=250)

knitr::opts_chunk$set(eval = TRUE, 
                      echo = TRUE, 
                      include = TRUE,
                      warning = FALSE,
                      collapse = FALSE,
                      message = FALSE,
                      dpi=300, dev = "png",
                      engine = "R", # Chunks will always have R code, unless noted
                      error = TRUE) 
```

# Load packages

```{r load-packages}
# Efficiently load packages 
pacman::p_load(ggplot2, phyloseq, ggpubr, tidyverse, patchwork, ggh4x, speedyseq, rstatix, dplyr, purrr, vegan, ANCOMBC, microViz,cowplot, grid, scales, Biostrings, stringr, lmerTest, ggtext, install = FALSE)

source("code/functions.R") # contains scale_reads
source("code/colors_and_shapes.R")

# Set our seed for reproducibility
set.seed(09091999)
```

# Load in Data

Loading in our .RData objects.

1. Water samples have flow cytometry derived absolute abundances.

2. Sediments do not have absolute abundance measurements incorporated. Instead, it has been scaled down to the minimum reads. Note: we scale down to the minimum number of reads to standardize/normalize our data to allow for more accurate comparisons between reads to account for uneven sequencing depth across samples (may arise to do sequencing runs or library prep efficiency, etc.)

And we will add in our metadata as well!

```{r load-metadata}
# water physeq with absolute abundance counts, this hasnt pruned the taxa out yet
# load("data/00_load_data/full_abs_physeq.RData")

# or we can also load in water physeq that has absolute abundance as well
load("data/00_load_data/water_physeq.RData")

# sediment scaled physeq
load("data/00_load_data/scaled_sed_physeq.RData")

## Add JDate to the sample_data 
sample_data(scaled_sed_physeq)$JDate <-
  lubridate::yday(sample_data(scaled_sed_physeq)$Date_Collected)

# physeq with water (unincorporated cell counts) + sediment samples = 188 samples total
load("data/00_load_data/new_archaea_rooted_physeq.RData")

## Add JDate to the sample_data 
sample_data(new_archaea_rooted_physeq)$JDate <-
  lubridate::yday(sample_data(new_archaea_rooted_physeq)$Date_Collected)

# load in metadata
load("data/00_load_data/meta_track_23_24.RData")
metadata <- meta_track_23_24
```


# Prepare Phyloseq Objects
Here we will work with our two sample types (water and sediment) for the entire 2024 year. 

We incorporated total cell counts from flow cytometry for water samples (full_abs_physeq.RData)
For our sediment samples, we do not have absolute abundance measures so we will need to calculate the relative abundance and rarify our samples to the minimum read depth (20822)

### Water 
First we will filter out our water samples and save it as a phyloseq object and data frame

```{r water-phyloseq}
# filter for all time points in 2024
water_physeq_24 <- subset_samples(water_physeq, Year == "2024")

# prune taxa
water_physeq_24 <- water_physeq_24 %>% 
  prune_taxa(taxa_sums(.) > 0,.)

# melt physeq into data frame for all taxa
water_physeq_ch4_df <- water_physeq_24 %>%
  speedyseq::psmelt() # melt into dataframe
```
We have created:
1. `water_physeq_24` phyloseq object of just 2024 water samples
2. `water_physeq_ch4_df` data frame from water physeq

### Sediment 
Filter out sediment samples and save it as a phyloseq object and data frame.
Same as the water samples we will filter out our phyloseq object and save it.

We do not have absolute abundance counts for our sediment samples so we will need to rarefy to the minimum sequencing depth (20826 reads)
```{r sed-phyloseq-24}
# filter phyloseq for only sediment samples
sed_phy <- subset_samples(new_archaea_rooted_physeq, SampleType == "Sediment")

# subset samples 
sed_physeq <- subset_samples(sed_phy, !(sample_names(sed_phy) %in% c("SA_D046", "SA_D047")))

# prune taxa 
sed_phy <- sed_physeq %>% 
  prune_taxa(taxa_sums(.) > 0,.)

# Intuition check of number of sequences per sample
min(sample_sums(sed_phy)) ## min = 20826

# scale water_physeq to minimum number of reads
scaled_sed_physeq <- 
  sed_phy %>% 
  scale_reads(round = "matround")

# melt physeq into data frame for all time points 
scaled_sed_physeq_24 <- scaled_sed_physeq 

# pull unique methane cycler taxonomic information 
scaled_sed_24_df <- scaled_sed_physeq_24 %>% 
  speedyseq::transform_sample_counts(function(x) x/sum(x)) %>% # transform to relative abundance here
  speedyseq::psmelt()  # melt into dataframe 

# save phyloseq object
save(scaled_sed_24_df, file = "data/01_phyloseq/scaled_sed_24_df.RData")
```  
We have created:
1. `scaled_sed_physeq_24` phyloseq object of sediment samples that have been rarified to minimum sequencing depth
2. `scaled_sed_24_df` melted sediment physeq data frame that has been transformed to relative abundance

# FAPROTAXv2 - Predict Metabolic Functions (CH4 Cyclers)
Now that we have saved our phyloseq objects from our water and sediment samples, we will be using FAPROTAXv2 to identify which ASVs are methanotrophs or methanogens. 

This is different from the original FAPROTAXv1 which uses taxonomic identification based on the SILVA database. Although it does understand GreenGenes2 naming, with the updated GreenGenes2 taxonomic identification, it does predict less than if we had used SILVA which makes sense since it is now incorporating taxonomic identities to unify metagenomic and amplicon sequencing studies.

Instead we will be using FAPROTAXv2's function to take the representative 16S rRNA sequences and place it on a phylogenetic tree using a reference tree to functionally predict ASVs. We will then take those ASVs and integrate it with our taxonomy table to see who is identified as what, preferably at the genus level.

### Water - Create FASTA File
Here we will start with a classic OTU table in TSV format but without taxonomic identities. Instead we will have 16S rRNA sequences in a fasta format to place OTUs on FAPROTAX reference tree and functionally annotate from there
```{r faprotax2-ref16S-prep-water}
# set seed
set.seed(09091999)

# 1. Create otu table 
classic_otu_table_water <- water_physeq_24 %>%
  otu_table %>%
  psmelt() %>%
  pivot_wider(names_from = Sample, values_from = Abundance) 

write_tsv(classic_otu_table_water, file = paste0("data/02A_Water_FAPROTAXv2/water_otu_table.tsv"))

# create fasta file 
otu_water_24 <- psmelt(water_physeq_24) %>%
  dplyr::select(OTU, Kingdom:ASVseqs) %>% 
    distinct(OTU, .keep_all = TRUE) %>% # i think one unique otu is fine?
  mutate(taxonomy = paste(Kingdom, Phylum, Class, Order, Family, sep=";")) %>% 
  dplyr::select(OTU,ASVseqs)


# use biostrings to create fasta file 
dna_seqs <- DNAStringSet(otu_water_24$ASVseqs)
names(dna_seqs) <- otu_water_24$OTU
writeXStringSet(dna_seqs, "data/02A_Water_FAPROTAXv2/water_16S_seqs.fasta")

# intuition check, no news is good news
# Check if OTU IDs match between the two files
otu_ids_table <- otu_water_24$OTU
otu_ids_fasta <- names(dna_seqs)

identical(sort(otu_ids_table), sort(otu_ids_fasta)) # check id table

stopifnot(identical(names(dna_seqs), otu_water_24$OTU)) # check that fasta df is good, no news is good news
```
Now that we have created our .fasta file from our water samples, we can now run FAPROTAXv2 on them to functionally predict.

### Water - Run FAPROTAXv2
First we will align our sequences to the reference sequences using mafft.
```{r water-faprotax2-fastas, engine = "bash", engine.opts = "l", eval = FALSE}
# move directory
cd data/02A_Water_FAPROTAXv2

# get mafft onto path -- make sure to copy this in terminal 
export PATH=/programs/mafft/bin:$PATH


# align our 16S rRNA fasta list to FAPROTAX2 database reference OR align through mafft GUI website if crashing
/programs/mafft/bin/mafft --auto --addfragments water_16S_seqs.fasta --thread -12 --keeplength SILVA_FOCAL_ALIGNMENTS_SANITIZED_SUBSET.fasta > water_output_alignment.fasta

# gui website code that was run
mafft --thread 12 --inputorder --keeplength --anysymbol --maxambiguous 0.0 --addfragments fragments --auto input > output

# run script to functionally annotate
# since we have already created alignment we place our aligned sequences in already and no need to add reference file

# run script to create phylogenetic tree
# first set environment to have epa-ng to create tree
export LD_LIBRARY_PATH=/usr/local/gcc-7.3.0/lib64:/usr/local/gcc-7.3.0/lib
export PATH=/programs/epa-ng-0.3.8/bin:$PATH
export LD_LIBRARY_PATH=/usr/lib64:$LD_LIBRARY_PATH # was having issues for getting Rcpp to work but this got it to go!

# run script with aligned mafft aligned fasta file and tree
# running with default for now which uses the binomial hsp algorithm
Rscript faprotax.R -i water_otu_table.tsv -a alignment_ch4water.fasta -o ch4_water_function_table_16S.tsv --out_intermediates ch4_water_intermediates2 -d /local//workdir/sna49/FS_CH4_Mech_Ray_LO_Letters/data/02A_Water_FAPROTAXv2 -r ch4_water_report.txt  --otu_names_are_in_column "OTU" --hsp_algorithm "binomial"
```

### Water - Extract Predicted FAPROTAXv2 Results
Now we will get our function table and find the predicted metabolic functions that correspond to methane cycling, methanogens or methanotrophs. It will provide the ASV as well as the predicted function.

First we will extract all functions then narrow in on CH4 cyclers. This will also be paired with taxonomic identification of known methane cyclers at the genus level.

ASVs not classified to the family or genus level were filtered out from analysis despite FAPROTAXv2 predictions. 

Each ASV will be verified through literature review to ensure it is a known methanogen or methanotroph
```{r pull-faprotaxv2-asvs}
# Read the text file of faprotaxv2 report from the previous chunk!
lines <- readLines("data/02A_Water_FAPROTAXv2/ch4_water_report.txt")

# create empty vectors
function_group <- c()
asv_ids <- c()

# temporary variable to store current function name
current_function <- NA

for (line in lines) {
  line <- str_trim(line)
  
  # function group name
  if (str_detect(line, ":$")) {
    current_function <- str_extract(line, "^[^\\s(]+")  # gets function name before the space
  }
  
  # reads asv lines (i.e. ASV_8 (P = 1))
  if (str_detect(line, "^ASV_\\d+")) {
    asv_id <- str_extract(line, "ASV_\\d+")
    function_group <- c(function_group, current_function)
    asv_ids <- c(asv_ids, asv_id)
  }
}

# Combine into dataframe
asv_df <- data.frame(function_group, asv_id = asv_ids)

# View if curious!
#View(asv_df)


# now lets specifically pull out the ch4 cyclers
water_ch4_cyclers_asv_df <- asv_df %>% 
  dplyr::filter(function_group %in% c(
                "acetoclastic_methanogenesis.value",
                "hydrogenotrophic_methanogenesis.value",
                "methanogenesis_by_CO2_reduction_with_H2.value",
                "methanogenesis_by_disproportionation_of_methyl_groups.value",
                "methanogenesis_by_reduction_of_methyl_compounds_with_H2.value",
                "methanogenesis_using_formate.value",
                "methanogenesis.value",
                "methanotrophy.value")) #%>% 
  # dplyr::mutate(
  #   CH4_Cycler = dplyr::case_when(
  #     grepl("methanogenesis", function_group) ~ "Methanogen",
  #     grepl("methanotrophy", function_group) ~ "Methanotroph",
  #     TRUE ~ NA_character_))

# quick check to see if we have any duplicated asvs
duplicates <- water_ch4_cyclers_asv_df %>%
  dplyr::group_by(asv_id) %>%
  dplyr::summarise(
    n_function_groups = dplyr::n_distinct(function_group), # number of function groups
    function_groups = paste(unique(function_group), collapse = ", ") # if theres another one where asv is duplicated
  ) %>%
  dplyr::filter(n_function_groups > 1) # count
duplicates # awesome methanogenesis functions overlap


# methanogens have multiple annotations but they are all methanogenesis, we will keep the first occurrence of each asv since the specific function subtype doesnt matter to much to me right now
water_ch4_cyclers_asv_unique <- water_ch4_cyclers_asv_df %>%
  dplyr::distinct(asv_id, .keep_all = TRUE)

# join ch4 cyclers with water_physeq_ch4_df
water_ch4_joined <- water_physeq_ch4_df %>%
  dplyr::inner_join(water_ch4_cyclers_asv_unique, 
                    by = c("ASV" = "asv_id")) # 20112 observations!

# tidy up dataframe
water_ch4_joined <- water_ch4_joined %>% 
  dplyr::select(function_group, Kingdom, Phylum, Class, Order,
                Family, Genus, Species, ASV, ASVseqs) 


# check out unqiue taxanomic levels
water_ch4_joined$Phylum %>% unique()
water_ch4_joined$Order %>% unique()
water_ch4_joined$Genus %>% unique()
water_ch4_joined$ASV %>% unique()

# filter out NA at order level
water_ch4_joined <- water_ch4_joined %>% 
  dplyr::filter(!is.na(Family)) # 16368

# create dataframe for sipmlified distinct ASV to verify
simplified_af <- water_ch4_joined %>% 
  distinct(Genus, .keep_all = TRUE) 
simplified_af

# now comparing with literature who is identified as a ch4 cycler
# filtering out incorrectly identified asvs, or asvs with no literature supporting their predicted metabolic function
water_ch4_joined_clean_faprotax <- water_ch4_joined %>% 
  dplyr::filter(
    !Genus %in% c(
    "Derxia", "UBA10450", "UBA11358", "Alsobacter", "UBA3015", "GAS474", "JAAUTS01",
    "UBA6215", "LW23", "WYBW01", "Pedosphaera", "Azospirillum", "YA12-FULL-60-10",
    "Beijerinckia", "SXTU01", "AV2", "Lenti-01", "PWTM01", "UBA95"),
    !Family %in% c("J093", "UBA10450"),
    !ASV %in% c("ASV_23314", "ASV_301", "ASV_3420", "ASV_2686", "ASV_2686", "ASV_3289")) # total 10560

# notes for my exclusions:
# 1. cannot find any information for derxia being methanotroph
# 2. UBA10450 (family) is within Chthoniobacterales (order) within Verrucomicrobiota. since this is at family level its hard to say what is function is. im finding its involved in complex carbon degradation and could maybe be a methanotroph and have other pathways to break this down but with this uncertainty and especially since no genus level identification will omit for now. most are apparently aerobic/anaerobic chemoorganotrophs
  # verruco methanotrophs are within methylacidiphilaceae family https://www.mdpi.com/2076-2607/12/11/2271
# 3. UBA11358 (family and genus) within Verrucomicrobiota i cant find solid evidence for methanotrophy either but based on these MAGs seems they are involved in other h2 metabolisms and stuff; biorxiv https://www.biorxiv.org/content/10.1101/2024.07.02.601631v1
# 4. aslobacter soli is a chemo-organotroph incapable of growth on C1 substrates [https://www.microbiologyresearch.org/content/journal/ijsem/10.1099/ijsem.0.003088#R1]
# 5. Methylacidiphilales (order), UBA3015 (family), UBA3015 (genus). not methanotroph likely heterotroph invovled in plant degradation. [https://registry.seqco.de/names/48695; https://www.nature.com/articles/s41467-025-63266-9]
# 6. GAS474 lacks genes for methanotrophy likely methylotroph or complex carbon degrader [https://www.nature.com/articles/s41522-024-00583-9]
# 7. JAAUTS01 no evidence for known methanotrophy
# 8. UBA6215 within Omnitrophota are not known to be methanotrophs but have diverse metabolic capabilities like acetogenesis and potentially parasitic? [https://www.nature.com/articles/s41564-022-01319-1]
# 9. LW23 likely methylotroph [https://www.mdpi.com/2076-2607/9/12/2566?utm_source=researchgate.net&medium=article]
# 10. WYBW01 methylotroph [https://www.biorxiv.org/content/10.1101/2025.02.12.637893v3.full]
# 11. Pedosphaera no evidence for methanotrophy
# 12. Azospirillum is a methylotroph, no evidence for methanotrophy [https://www.nature.com/articles/s43247-024-01656-5]
# 13. YA12-FULL-60-10 no evidence for methanotrphy [https://www.mdpi.com/2076-2607/8/6/920?utm_source=researchgate.net&medium=article]
# 14. Beijerinckia chemoheterotroph, not capable of methylotrophic or methanotrophy. [https://journals.asm.org/doi/10.1128/jb.00656-10]
# 15. SXTU01 not known methanotroph, no literature supporting this or found
# 16. AV2, no evidence for methanotrophy
# 17. Lenti-01, no evidence for methanotrophy. specialists for degrading sulfated polysaccharides [https://pmc.ncbi.nlm.nih.gov/articles/PMC8857213/#:~:text=A%20phylogenetic%20reconstruction%20using%20conserved,also%20determined%20(Supplementary%20results).]
# 18. PWTM01 no reports of methanotrophy
# 19. UBA95 could not find support for methanogenesis, seems to have other metabolic capabilities like secondary metabolites and nitrogen cycling.
# 20. JO93 not methanotroph likely, in different order with that doesnt have known methanotrophs
# 21. removing ASVs within Beijerinckiaceae family that are NA at genus level as not all genera are capable of methanotrophy


# distinct at family level for ch4 cycler
# one NA at genus but family is all known methanotrophs
# these are just intuition checks!
water_ch4_joined_clean_faprotax$Order %>% unique()
faprotaxv2_family <- unique(water_ch4_joined_clean_faprotax$Family)


# now lets create data frame to search in case faprotax missed a literature-based identified methanotroph 
water_24_taxonomysearch <- water_physeq_ch4_df %>%
  dplyr::select(OTU, Kingdom:ASVseqs) %>% 
  distinct(OTU, .keep_all = TRUE) %>% # one ASV per row
  dplyr::filter(!is.na(Family)) # cannot be na and family level and below


# rationale to include this above:
  #1. 2-02-FULL-66-22 contains nitrite-dependent anaerobic methane oxidizing (N-DAMO) bacteria that couple anaerobic methane oxidation to nitrite reduction.
  #2. Methanomethylophilaceae within Methanomassiliicoccales is a methanogen

# additional information from searching:
  #1. UBA12499 within Rokubacteriales in Methylomirobiota is a methylotroph, no known genes to oxidize methane.


# create dataframes for manual search 
water_24_manual <- water_24_taxonomysearch %>%
  dplyr::filter(Family %in% c("2-02-FULL-66-22", "Methanomethylophilaceae", "UBA472", 
                              "Methanomassiliicoccaceae", "Methylomonadaceae", "Methanobacteriaceae") |
                Order %in% c("Methylococcales"))# %>% 
  #dplyr::filter
  # family level to also include manually (e.g., specific N-DAMO methanotroph and methanogen
  #dplyr::distinct(ASV, .keep_all = TRUE) # one ASV per row
    

# finalize dataframe column orders before merging
water_ch4_joined_clean_faprotax_trim <- water_ch4_joined_clean_faprotax %>%
  dplyr::select(-function_group)

water_ch4_manual_trim <- water_24_manual %>% 
  dplyr::select(-OTU)


# bind rows
water_ch4_cyclers_complete_df <- dplyr::bind_rows(
  water_ch4_joined_clean_faprotax_trim,
  water_ch4_manual_trim
)

# list of all our CH4 cycler ASVs to filter phyloseq 
ch4_asv_list <- water_ch4_cyclers_complete_df$ASV

# filter phyloseq for CH4 cyclers at the ASV level
water_ch4_cyclers_physeq <- water_physeq_24 %>% 
  subset_taxa(ASV %in% ch4_asv_list)


# create list of CH4 cylcers at the family level!
water_methanogens <- c("Methanotrichaceae", "Methanospirillaceae_2121", "Methanobacteriaceae",
                 "Methanospirillaceae_2125", "Methanosarcinaceae","Methanocellaceae", 
                 "Methanomicrobiaceae", "Methanofastidiosaceae", "Methanomethylophilaceae", "UBA472", "Methanomassiliicoccaceae")
water_methanotrophs <- c("Methylomonadaceae", "Methylococcales","Beijerinckiaceae", "2-02-FULL-66-22")

# # create list of CH4 cylcers at the order level!
# water_ch4_cyclers_complete_df$Order %>% unique()
# water_methanogens_o <- c("Methanotrichaceae", "Methanospirillaceae_2121", "Methanobacteriaceae",
#                  "Methanospirillaceae_2125", "Methanosarcinaceae","Methanocellaceae", 
#                  "Methanomicrobiaceae", "Methanofastidiosaceae", "Methanomethylophilaceae", "UBA472", "Methanomassiliicoccaceae")
# water_methanotrophs_o <- c("Methylomonadaceae", "Methylococcaceae","Beijerinckiaceae", "2-02-FULL-66-22")

# add if methanogen or methanotroph to column
tax_tab <- as.data.frame(tax_table(water_ch4_cyclers_physeq))

tax_tab$CH4_Cycler <- dplyr::case_when(
  tax_tab$Family %in% water_methanogens ~ "Methanogen",
  tax_tab$Family %in% water_methanotrophs   ~ "Methanotroph",
  tax_tab$Order %in% water_methanotrophs   ~ "Methanotroph",
  TRUE ~ "NA"
)


tax_table(water_ch4_cyclers_physeq) <- tax_table(as.matrix(tax_tab))

water_ch4_cyclers_physeq # 255 total ASVs

# save phyloseq object
save(water_ch4_cyclers_physeq, file = "data/01_phyloseq/water_ch4_cyclers_physeq.RData")

# melt into dataframe for plotting CH4 cyclers
water_ch4_cyclers_df <- water_ch4_cyclers_physeq %>%
  speedyseq::psmelt() %>%  # melt into dataframe
  dplyr::mutate(
    solar_progress = recode(solar_progress, "FPV" = "FPV", "No FPV" = "Open"), # solar progress
    Depth_Class = recode(Depth_Class,  # depth class
      "S" = "Surface Water",
      "B" = "Bottom Water")) %>% 
  dplyr::select(DNA_ID, Abundance, Pond, solar_progress, CH4_Cycler, JDate, Date_Collected, Depth_Class,
                Phylum, Class, Order, Family, Genus, ASV, Species)

# save water CH4 cyclers df as .RData
save(water_ch4_cyclers_df, file = "data/01_phyloseq/water_ch4_cyclers_df.RData")


# write out to excel file 
write.csv(water_ch4_cyclers_df, "data/01_phyloseq/water_ch4_cyclers.csv")

```
Here we used FAPROTAXv2 along with manual identification to functionally predict methanogens and methanotrophs in the water column. 

We created:
1. **water_ch4_cylers_physeq** phyloseq object that has all 2024 time points and filtered for *only* methane cyclers

2. **water_ch4_cyclers_df** data frame that will be used for plotting that includes total cell counts and reformatted names such as depth and solar treatment. This was also saved as a .csv file called "water_ch4_cyclers.csv"

### Sediment - Create FASTA File
Here we will start with a classic OTU table in TSV format but without taxonomic identities. Instead we will have 16S rRNA sequences in a fasta format to place OTUs on FAPROTAX reference tree and functionally annotate from there
```{r faprotax2-ref16S-prep-sed}
# set seed
set.seed(09091999)

# 1. Create otu table 
classic_otu_table_sed <- scaled_sed_physeq %>%
  otu_table %>%
  psmelt() %>%
  pivot_wider(names_from = Sample, values_from = Abundance) 

write_tsv(classic_otu_table_sed, file = paste0("data/02B_Sediment_FAPROTAXv2/sed_otu_table.tsv"))

# create fasta file 
otu_sed_24 <- psmelt(scaled_sed_physeq) %>%
  dplyr::select(OTU, Kingdom:ASVseqs) %>% 
    distinct(OTU, .keep_all = TRUE) %>% 
  mutate(taxonomy = paste(Kingdom, Phylum, Class, Order, Family, sep=";")) %>% 
  dplyr::select(OTU,ASVseqs)


# use biostrings to create fasta file 
dna_seqs <- DNAStringSet(otu_sed_24$ASVseqs)
names(dna_seqs) <- otu_sed_24$OTU
writeXStringSet(dna_seqs, "data/02B_Sediment_FAPROTAXv2/sed_16S_seqs.fasta")

# intuition check, no news is good news
# Check if OTU IDs match between the two files
otu_ids_table <- otu_sed_24$OTU
otu_ids_fasta <- names(dna_seqs)

identical(sort(otu_ids_table), sort(otu_ids_fasta)) # check id table

stopifnot(identical(names(dna_seqs), otu_sed_24$OTU)) # check that fasta df is good, no news is good news
```
Now that we have created our .fasta file from our water samples, we can now run FAPROTAXv2 on them to functionally predict.

### Sediment - Run FAPROTAXv2
First we will align our sequences to the reference sequences using mafft.
```{r sed-faprotax2-fastas, engine = "bash", engine.opts = "l", eval = FALSE}
# move directory
cd data/02B_Sediment_FAPROTAXv2

# get mafft onto path -- make sure to copy this in terminal or nothing happens <3
export PATH=/programs/mafft/bin:$PATH


# align our 16S rRNA fasta list to FAPROTAX2 database reference OR align through mafft GUI website if crashing
/programs/mafft/bin/mafft --auto --addfragments sed_16S_seqs.fasta --thread -12 --keeplength SILVA_FOCAL_ALIGNMENTS_SANITIZED_SUBSET.fasta > sed_output_alignment.fasta

# gui website code that was run
mafft --thread 12 --inputorder --keeplength --anysymbol --maxambiguous 0.0 --addfragments fragments --auto input > output

# run script to functionally annotate
# since we have already created alignment we place our aligned sequences in already and no need to add reference file

# run script to create phylogenetic tree
# first set environment to have epa-ng to create tree
export LD_LIBRARY_PATH=/usr/local/gcc-7.3.0/lib64:/usr/local/gcc-7.3.0/lib
export PATH=/programs/epa-ng-0.3.8/bin:$PATH
export LD_LIBRARY_PATH=/usr/lib64:$LD_LIBRARY_PATH # was having issues for getting Rcpp to work but this got it to go!

# run script with aligned mafft aligned fasta file and tree
# running with default for now which uses the binomial hsp algorithm
Rscript faprotax.R -i sed_otu_table.tsv -a alignment_ch4sed.fasta -o ch4_sed_function_table_16S.tsv --out_intermediates ch4_sed_intermediates -d /local/workdir/sna49/FS_CH4_Mech_Ray_LO_Letters/data/02B_Sediment_FAPROTAXv2 -r ch4_sed_report.txt  --otu_names_are_in_column "OTU" --hsp_algorithm "binomial"
```

### Sediment - Extract Predicted FAPROTAXv2 Results
Now we will get our function table and find the predicted metabolic functions that correspond to methane cycling, methanogens or methanotrophs. It will provide the ASV as well as the predicted function.

First we will extract all functions then narrow in on CH4 cyclers. This will also be paired with taxonomic identification of known methane cyclers at the genus level.

ASVs not classified to the family or genus level were filtered out from analysis despite FAPROTAXv2 predictions. 

Each ASV will be verified through literature review to ensure it is a known methanogen or methanotroph
```{r pull-faprotaxv2-sed-asvs}
# Read the text file of faprotaxv2 report from the previous chunk!
lines <- readLines("data/02B_Sediment_FAPROTAXv2/ch4_sed_report.txt")

# create empty vectors
function_group <- c()
asv_ids <- c()

# temporary variable to store current function name
current_function <- NA

for (line in lines) {
  line <- str_trim(line)
  
  # function group name
  if (str_detect(line, ":$")) {
    current_function <- str_extract(line, "^[^\\s(]+")  # gets function name before the space
  }
  
  # reads asv lines (i.e. ASV_8 (P = 1))
  if (str_detect(line, "^ASV_\\d+")) {
    asv_id <- str_extract(line, "ASV_\\d+")
    function_group <- c(function_group, current_function)
    asv_ids <- c(asv_ids, asv_id)
  }
}

# Combine into dataframe
asv_df <- data.frame(function_group, asv_id = asv_ids)

# View if curious!
#View(asv_df)


# now lets specifically pull out the ch4 cyclers
sed_ch4_cyclers_asv_df <- asv_df %>% 
  dplyr::filter(function_group %in% c(
                "acetoclastic_methanogenesis.value",
                "hydrogenotrophic_methanogenesis.value",
                "methanogenesis_by_CO2_reduction_with_H2.value",
                "methanogenesis_by_disproportionation_of_methyl_groups.value",
                "methanogenesis_by_reduction_of_methyl_compounds_with_H2.value",
                "methanogenesis_using_formate.value",
                "methanogenesis.value",
                "methanotrophy.value")) 

# quick check to see if we have any duplicated asvs
duplicates <- sed_ch4_cyclers_asv_df %>%
  dplyr::group_by(asv_id) %>%
  dplyr::summarise(
    n_function_groups = dplyr::n_distinct(function_group), # number of function groups
    function_groups = paste(unique(function_group), collapse = ", ") # if theres another one where asv is duplicated
  ) %>%
  dplyr::filter(n_function_groups > 1) # count
duplicates # awesome methanogenesis functions overlap


# methanogens have multiple annotations but they are all methanogenesis, we will keep the first occurrence of each asv since the specific function subtype doesnt matter to much to me right now
sed_ch4_cyclers_asv_unique <- sed_ch4_cyclers_asv_df %>%
  dplyr::distinct(asv_id, .keep_all = TRUE)

# join ch4 cyclers with sed_physeq_ch4_df
sed_ch4_joined <- scaled_sed_24_df %>%
  dplyr::inner_join(sed_ch4_cyclers_asv_unique, 
                    by = c("ASV" = "asv_id")) # 36202 observations!

# tidy up dataframe
sed_ch4_joined <- sed_ch4_joined %>% 
  dplyr::select(function_group, Kingdom, Phylum, Class, Order,
                Family, Genus, Species, ASV, ASVseqs) 


# check out unqiue taxanomic levels
sed_ch4_joined$Phylum %>% unique()
sed_ch4_joined$Order %>% unique()
sed_ch4_joined$Genus %>% unique()
sed_ch4_joined$ASV %>% unique()

# filter out NA at order level
sed_ch4_joined <- sed_ch4_joined %>% 
  dplyr::filter(!is.na(Family)) # 23628

# create dataframe for simplified distinct ASV to verify
simplified_af <- sed_ch4_joined %>% 
  distinct(Family, .keep_all = TRUE) 
simplified_af

# now comparing with literature who is identified as a ch4 cycler
# filtering out incorrectly identified asvs, or asvs with no literature supporting their predicted metabolic function
sed_ch4_joined_clean_faprotax <- sed_ch4_joined %>% 
  dplyr::filter(
    !Genus %in% c("AV2", "Alsobacter", "Beijerinckia", "BOG-1460", "DP16D-bin41", "MWDV01", "SLKG01", "UBA7542", 	
                  "WLWN01", "UBA11358",  "UBA3015", "GAS474", "JAAUTS01","UBA6215", "LW23", "WYBW01", 
                  "Pedosphaera", "YA12-FULL-60-10", "UBA10450", "UBA3939", "PWTM01", "UBA95", "Lenti-01"),
    !Family %in% c("Pontiellaceae","J093", "UBA10450", "JAAZAB01", "UBA11358"),
    !ASV %in% c("ASV_3646", "ASV_468", "ASV_24935", "ASV_3646", "ASV_468")) # total 13200

# notes for my exclusions:
# 1. cannot find any information for derxia being methanotroph
# 2. UBA10450 (family) is within Chthoniobacterales (order) within Verrucomicrobiota. since this is at family level its hard to say what is function is. im finding its involved in complex carbon degradation and could maybe be a methanotroph and have other pathways to break this down but with this uncertainty and especially since no genus level identification will omit for now. most are apparently aerobic/anaerobic chemoorganotrophs
  # verruco methanotrophs are within methylacidiphilaceae family https://www.mdpi.com/2076-2607/12/11/2271
# 3. UBA11358 (family and genus) within Verrucomicrobiota i cant find solid evidence for methanotrophy either but based on these MAGs seems they are involved in other h2 metabolisms and stuff; biorxiv https://www.biorxiv.org/content/10.1101/2024.07.02.601631v1
# 4. aslobacter soli is a chemo-organotroph incapable of growth on C1 substrates [https://www.microbiologyresearch.org/content/journal/ijsem/10.1099/ijsem.0.003088#R1]
# 5. Methylacidiphilales (order), UBA3015 (family), UBA3015 (genus). not methanotroph likely heterotroph invovled in plant degradation. [https://registry.seqco.de/names/48695; https://www.nature.com/articles/s41467-025-63266-9]
# 6. GAS474 lacks genes for methanotrophy likely methylotroph or complex carbon degrader [https://www.nature.com/articles/s41522-024-00583-9]
# 7. JAAUTS01 no evidence for known methanotrophy
# 8. UBA6215 within Omnitrophota are not known to be methanotrophs but have diverse metabolic capabilities like acetogenesis and potentially parasitic? [https://www.nature.com/articles/s41564-022-01319-1]
# 9. LW23 likely methylotroph [https://www.mdpi.com/2076-2607/9/12/2566?utm_source=researchgate.net&medium=article]
# 10. WYBW01 methylotroph [https://www.biorxiv.org/content/10.1101/2025.02.12.637893v3.full]
# 11. Pedosphaera no evidence for methanotrophy
# 12. Azospirillum is a methylotroph, no evidence for methanotrophy [https://www.nature.com/articles/s43247-024-01656-5]
# 13. YA12-FULL-60-10 no evidence for methanotrphy [https://www.mdpi.com/2076-2607/8/6/920?utm_source=researchgate.net&medium=article]
# 14. Beijerinckia chemoheterotroph, not capable of methylotrophic or methanotrophy. [https://journals.asm.org/doi/10.1128/jb.00656-10]
# 15. SXTU01 not known methanotroph, no literature supporting this or found
# 16. AV2, no evidence for methanotrophy
# 17. Lenti-01, no evidence for methanotrophy. specialists for degrading sulfated polysaccharides [https://pmc.ncbi.nlm.nih.gov/articles/PMC8857213/#:~:text=A%20phylogenetic%20reconstruction%20using%20conserved,also%20determined%20(Supplementary%20results).]
# 18. PWTM01 no reports of methanotrophy
# 19. UBA95 could not find support for methanogenesis, seems to have other metabolic capabilities like secondary metabolites and nitrogen cycling.
# 20. JO93 not methanotroph likely, in different order with that doesnt have known methanotrophs
# 21. removing ASVs within Beijerinckiaceae family that are NA at genus level as not all genera are capable of methanotrophy
# 22. BOG-1460 is not known methanotroph
# 23. DP16D-bin41 not known methanotroph
# 24. VSJD01 not known methanotroph 


# distinct at family level for ch4 cycler
# one NA at genus but family is all known methanotrophs
# these are just intuition checks!
sed_ch4_joined_clean_faprotax$Family %>% unique()
faprotaxv2_family <- unique(sed_ch4_joined_clean_faprotax$Family)


# now lets create data frame to search in case faprotax missed a literature-based identified methanotroph 
sed_24_taxonomysearch <- scaled_sed_24_df %>%
  dplyr::select(OTU, Kingdom:ASVseqs) %>% 
  distinct(OTU, .keep_all = TRUE) %>% # one ASV per row
  dplyr::filter(!is.na(Family)) # cannot be na and family level and below

# rationale to include this above:
  #1. Methanomassiliicoccales (order level) methanogens to capture family and genus level; methylotrophic methanogen
  #2. Methanomethylicaceae (family) methyltrophic methanogen
  #3. 2-02-FULL-66-22 contains nitrite-dependent anaerobic methane oxidizing (N-DAMO) bacteria that couple anaerobic methane oxidation to nitrite reduction.


# create dataframes for manual search 
sed_24_manual <- sed_24_taxonomysearch %>%
  dplyr::filter(Family %in% c("2-02-FULL-66-22", "Methanomethylophilaceae", "Methylomonadaceae", "Methanotrichaceae", "Methanomethylicaceae") |
                Order %in% c("Methanomassiliicoccales")) %>%  # family level to also include manually (e.g., specific N-DAMO methanotroph and methanogen
  dplyr::distinct(ASV, .keep_all = TRUE) # one ASV per row
    

# finalize dataframe column orders before merging
sed_ch4_joined_clean_faprotax_trim <- sed_ch4_joined_clean_faprotax %>%
  dplyr::select(-function_group)

sed_ch4_manual_trim <- sed_24_manual %>% 
  dplyr::select(-OTU)


# bind rows
sed_ch4_cyclers_complete_df <- dplyr::bind_rows(
  sed_ch4_joined_clean_faprotax_trim,
  sed_ch4_manual_trim
)

# unique asv counts
test<- sed_ch4_cyclers_complete_df$ASV %>% unique() # correclty 347
# list of all our CH4 cycler ASVs to filter phyloseq 
ch4_asv_list <- sed_ch4_cyclers_complete_df$ASV

# filter phyloseq for CH4 cyclers at the ASV level
sed_ch4_cyclers_physeq <- scaled_sed_physeq %>% 
  speedyseq::transform_sample_counts(function(x) x/sum(x)) %>%  # calc rel abundance
  subset_taxa(ASV %in% ch4_asv_list)


# create list of CH4 cylcers at the family level!
sed_methanogens <- c("Methanotrichaceae", "Methanospirillaceae_2121", "Methanobacteriaceae",
                 "Methanospirillaceae_2125", "Methanosarcinaceae","Methanocellaceae", 
                 "Methanomicrobiaceae", "UBA472", "Methanomassiliicoccaceae", "Methanomethylicaceae") # note: methanomethylicaceae is involved in methylotrophic methanogenesis; produce methane from methyl compounds
sed_methanotrophs <- c("Methylomonadaceae", "Methylococcaceae","Beijerinckiaceae", "2-02-FULL-66-22",  "Methanoperedenaceae") # note:  Methanoperedenaceae is methanotrophic archaea (AOM)

# add if methanogen or methanotroph to column
tax_tab <- as.data.frame(tax_table(sed_ch4_cyclers_physeq))

tax_tab$CH4_Cycler <- dplyr::case_when(
  tax_tab$Family %in% sed_methanogens ~ "Methanogen",
  tax_tab$Family %in% sed_methanotrophs   ~ "Methanotroph",
  TRUE ~ "NA"
)


tax_table(sed_ch4_cyclers_physeq) <- tax_table(as.matrix(tax_tab))

sed_ch4_cyclers_physeq # 347 total ASVs, 44 samples

# save phyloseq object
save(sed_ch4_cyclers_physeq, file = "data/01_phyloseq/sed_ch4_cyclers_physeq.RData")

# melt into dataframe for plotting CH4 cyclers
sed_ch4_cyclers_df <- sed_ch4_cyclers_physeq %>%
  #speedyseq::tax_glom(taxrank = "ASV") %>% 
  # Calculate the relative abundance
  #speedyseq::transform_sample_counts(function(x) {x/sum(x)}) %>%
  psmelt() %>% 
  dplyr::mutate(
    solar_progress = recode(solar_progress, "FPV" = "FPV", "No FPV" = "Open"), # solar progress
    Depth_Class = recode(Depth_Class,  # depth class
      "SD" = "Sediment")) %>% 
  dplyr::select(DNA_ID, Abundance, Pond, solar_progress, CH4_Cycler, JDate, Date_Collected, Depth_Class,
                Phylum, Class, Order, Family, Genus, ASV, Species) 

# save water CH4 cyclers df as .RData
save(sed_ch4_cyclers_df, file = "data/01_phyloseq/sed_ch4_cyclers_df.RData")


# write out to excel file 
write.csv(sed_ch4_cyclers_df, "data/01_phyloseq/sed_ch4_cyclers.csv")
```
Here we used FAPROTAXv2 along with manual identification to functionally predict methanogens and methanotrophs in the sediments. 

We created:
1. **sed_ch4_cylers_physeq** phyloseq object that has all 2024 time points and filtered for *only* methane cyclers. This was from our scaled phyloseq that was scaled down to the minimum read number 20826

2. **sed_ch4_cyclers_df** data frame that will be used for plotting that includes relative abundance and reformatted names such as depth and solar treatment. This was also saved as a .csv file called "sed_ch4_cyclers.csv"


# Normality of Methane Cyclers in Ponds
**Is our data normally distributed?**

Lets configure our dataframe and then check to see how our data is distributed with Q-Q plots, density histogram, and Shapiro-Wilk test.

We will split this up by sample type where we will do water first then sediments in next chunk

### Water - Normality
```{r normality-water}
# factor solar progress
water_ch4_cyclers_df$solar_progress <- factor(
  water_ch4_cyclers_df$solar_progress,
  levels = c("FPV", "Open"))

# factor depth
water_ch4_cyclers_df$Depth_Class <- factor(
  water_ch4_cyclers_df$Depth_Class,
  levels = c("Surface Water", "Bottom Water"))

# now add interaction column to our df
water_ch4_cyclers_df <- water_ch4_cyclers_df %>%
  mutate(group = interaction(CH4_Cycler, Depth_Class, sep = " ")) %>% 
  group_by(Pond, solar_progress, Depth_Class, group, JDate, CH4_Cycler, DNA_ID) %>%
  summarise(
    total_abundance = sum(Abundance, na.rm = TRUE), # total across all samples
    .groups = "drop"
  )


# qq plot to visualize normality
ggplot(water_ch4_cyclers_df, aes(sample = total_abundance)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~ group, scales = "free") +
  theme_minimal() +
  labs(title = "Q-Q Plots: Water Abundance by Group")

# lets plot density histogram by group too
ggplot(water_ch4_cyclers_df, aes(x = total_abundance, fill = group)) +
  geom_histogram(aes(y = ..density..), bins = 30, alpha = 0.5, color = "black") +
  geom_density(alpha = 0.6) +
  facet_wrap(~ group, scales = "free") +
  theme_minimal() +
  labs(title = "Histogram and Density of Abundance by Group",
       x = "Absolute Cell Abundance",
       y = "Density") +
  theme(legend.position = "none")

# now lets plot density histogram by further facetting by treatment
ggplot(water_ch4_cyclers_df, aes(x = total_abundance, fill = group)) +
  geom_histogram(aes(y = ..density..), bins = 30, alpha = 0.5, color = "black") +
  geom_density(alpha = 0.6) +
  facet_wrap(~ group + solar_progress, scales = "free") +
  theme_minimal() +
  labs(title = "Histogram and Density of Order Abundance by Group and Solar Progress",
       x = "Absolute Cell Abundance",
       y = "Density") +
  theme(legend.position = "none")

# shapiro test
water_ch4_cyclers_df %>%
  group_by(group) %>%
  summarise(
    shapiro_p = shapiro.test(total_abundance)$p.value,
    n = n()
  )

```

Based on our data with the qq plots we see that the points more or less fit the line. but when we investigate further with the density histograms, the data lacks a clear unimodal distribution. Now, let's check this out further with a Shapiro-Wilk test. When we run our Shapiro-Wilk test to also test for a normal distribution if p > 0.05 the data is likely normal but if p < 0.05 then the data is not normal.

# A tibble: 4 × 3
  group                        shapiro_p     n
  <fct>                            <dbl> <int>
1 Methanogen Surface Water   0.00000185     24
2 Methanotroph Surface Water 0.000267       24
3 Methanogen Bottom Water    0.000000393    24
4 Methanotroph Bottom Water  0.000445       24

The data is not normal! 

Therefore, we will need to use non-parametric stats to statitistically test the data. 


### Sediment - Normality 

We will be running this for all 2024 time points 
```{r normality-sed}
# level
sed_ch4_cyclers_df$solar_progress <- factor(
  sed_ch4_cyclers_df$solar_progress,
  levels = c("FPV", "Open"))

# interaction grouping and relative abundance calc
sed_ch4_cyclers_df <- sed_ch4_cyclers_df %>%
  mutate(group = interaction(CH4_Cycler, Depth_Class, sep = " ")) %>% 
  group_by(Pond, solar_progress, group, JDate, CH4_Cycler) %>%
  summarise(
    rel_abundance = sum(Abundance, na.rm = TRUE), # total across all samples
    .groups = "drop"
  )


# qq plot to visualize normality
ggplot(sed_ch4_cyclers_df, aes(sample = rel_abundance)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~ group, scales = "free") +
  theme_minimal() +
  labs(title = "Q-Q Plots: Order Abundance by Group")

# lets plot density histogram by group too
ggplot(sed_ch4_cyclers_df, aes(x = rel_abundance, fill = group)) +
  geom_histogram(aes(y = ..density..), bins = 30, alpha = 0.5, color = "black") +
  geom_density(alpha = 0.6) +
  facet_wrap(~ group, scales = "free") +
  theme_minimal() +
  labs(title = "Histogram and Density of Relative Abundance by Group",
       x = "Relative Abundance",
       y = "Density") +
  theme(legend.position = "none")

# now lets plot density histogram by further facetting by treatment
ggplot(sed_ch4_cyclers_df, aes(x = rel_abundance, fill = group)) +
  geom_histogram(aes(y = ..density..), bins = 30, alpha = 0.5, color = "black") +
  geom_density(alpha = 0.6) +
  facet_wrap(~ group + solar_progress, scales = "free") +
  theme_minimal() +
  labs(title = "Histogram and Density of Relative Abundance by Group and Solar Progress",
       x = "Relative Abundance",
       y = "Density") +
  theme(legend.position = "none")

# shapiro test
sed_ch4_cyclers_df %>%
  group_by(group) %>%
  summarise(
    shapiro_p = shapiro.test(rel_abundance)$p.value,
    n = n()
  )
```

Based on our data with the qq plots and the density histograms, data does not seem to be clearly normally distributed. Shapiro-Wilk test to also test for a normal distribution if p > 0.05 the data is likely normal but if p < 0.05 then the data is not normal.

  group                 shapiro_p     n
  <fct>                     <dbl> <int>
1 Methanogen Sediment      0.137     23
2 Methanotroph Sediment    0.0184    23

Methanotrophs are not normally distributed but but methanogens are which is interesting. but will still go with two-sample Wilcoxon tests

# Fig 2 - Abundance of Methane Cyclers

In this plot we will calculate the absolute abundance of methane cyclers in water column and the relative abundance in sediments to see methane cyclers overtime and with a box plot to show abundance as well.
```{r fig-2-ch4-cycler-abundance}
# calculate total cell abundance mean and standard deviation 
wc_total_abund <- water_ch4_cyclers_df %>% 
  group_by(solar_progress, Depth_Class, CH4_Cycler) %>% 
  dplyr::summarize(ch4_water_sd = sd(total_abundance),
                   mean_ch4_water = mean(total_abundance),
                   .groups = "drop")
wc_total_abund
  

##### surface methanogens #####
# 1. calculate stats
# 1a. calculate abundances
methanogen_surface_sum <- water_ch4_cyclers_df %>%
  dplyr::filter(Depth_Class == "Surface Water",
                CH4_Cycler == "Methanogen") %>% 
  group_by(solar_progress, JDate) %>% 
  dplyr::summarize(sd_meth = sd(total_abundance),
                   mean_meth = mean(total_abundance),
                   .groups = "drop")
methanogen_surface_sum
max(methanogen_surface_sum$mean_meth)


# 1b. linear mixed model
methanogen_surface_data <- water_ch4_cyclers_df %>%
  dplyr::filter(Depth_Class == "Surface Water",
                CH4_Cycler == "Methanogen") 
methanogen_surface_data$Pond <- as.factor(methanogen_surface_data$Pond)

surface_methanogen_model <- lmerTest::lmer(total_abundance ~ solar_progress + JDate + (1|Pond), data = methanogen_surface_data)
summary(surface_methanogen_model) #0.876
emmeans::emmeans(surface_methanogen_model, pairwise ~ solar_progress) # 0.8820

# 2. plot surface water methanogens overtime
methanogen_surface <- methanogen_surface_sum %>% 
  ggplot(aes(x = JDate, y = mean_meth, fill = solar_progress, color = solar_progress)) +
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = JDate, ymin = mean_meth - sd_meth, ymax = mean_meth + sd_meth), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  labs(y= "Surface Methanogen<br>Abundance (Cells mL<sup>-1</sup>)", x = "Day of Year")+
  scale_y_continuous(
    limits = c(NA, 8250),
    # breaks = c(0, 2500, 5000, 7500),
    # limits = c(NA, 8e3),
    # breaks = breaks_extended(n = 4),
    breaks = c(0, 4000, 8000),
    labels = label_number(scale_cut = cut_short_scale())) +
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  scale_fill_manual(values = solar_colors) +
  scale_color_manual(values = solar_colors) +
  theme(legend.position = "none")#+
  #scale_x_continuous(limits = c(170,250))
methanogen_surface

# 3. box plot
methanogen_surface_box <- methanogen_surface_data %>% 
  ggplot(aes(x = solar_progress, y = total_abundance, color = solar_progress)) +
  geom_boxplot(outlier.shape = NA, color = "black") +
  geom_jitter(aes(color = solar_progress, fill = solar_progress), width = 0.2, size = 2, shape = 21) +
  scale_fill_manual(values = solar_colors)+
  scale_color_manual(values = c("black", "black"))+
  scale_y_continuous(
    limits = c(NA, 8250),
    #breaks = c(0, 2500, 5000, 7500),
     breaks = c(0, 4000, 8000),
    labels = label_number(scale_cut = cut_short_scale())) +
  theme_classic()+
  theme(axis.text.x = element_text(size = 8,colour = "black"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none") +
 annotate("text", x = 2, y = 8000, label = "p = 0.882", size = 2.822,  fontface = "italic") #.882
methanogen_surface_box

##### surface methanotrophs #####
# 1. calculate stats
# 1a. calculate abundances
methanotroph_surface_sum <- water_ch4_cyclers_df %>%
  dplyr::filter(Depth_Class == "Surface Water",
                CH4_Cycler == "Methanotroph") %>% 
  group_by(solar_progress, JDate) %>%
  dplyr::summarize(sd_meth = sd(total_abundance),
                   mean_meth = mean(total_abundance),
                   .groups = "drop")
methanotroph_surface_sum
max(methanotroph_surface_sum$mean_meth)

# calculate abundances between treatments 
methanotroph_surface_sum_text <- water_ch4_cyclers_df %>%
  dplyr::filter(Depth_Class == "Surface Water",
                CH4_Cycler == "Methanotroph") %>% 
  group_by(solar_progress) %>% 
  dplyr::summarize(sd_meth = sd(total_abundance/1e5),
                   mean_meth = mean(total_abundance/1e5),
                   .groups = "drop")
methanotroph_surface_sum_text
max(methanotroph_surface_sum_text$mean_meth)


# 1b. linear mixed model
methanotroph_surface_data <- water_ch4_cyclers_df %>%
  dplyr::filter(Depth_Class == "Surface Water",
                CH4_Cycler == "Methanotroph") 
methanotroph_surface_data$Pond <- as.factor(methanotroph_surface_data$Pond)

max(methanotroph_surface_data$total_abundance)

surface_methanotroph_model <- lmer(total_abundance ~ solar_progress + JDate + (1|Pond), data = methanotroph_surface_data)
summary(surface_methanotroph_model) # 0.000217 ***
emmeans::emmeans(surface_methanotroph_model, pairwise ~ solar_progress) # 0.0209


# 2. plot surface water methanotrophs
methanotroph_surface <- methanotroph_surface_sum %>% 
  ggplot(aes(x = JDate, y = mean_meth, fill = solar_progress, color = solar_progress)) +
  #geom_smooth(aes(group = solar_progress), se = FALSE) +
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = JDate, ymin = mean_meth - sd_meth, ymax = mean_meth + sd_meth), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  labs(y= "Surface Methanotroph<br>Abundance (Cells mL<sup>-1</sup>)", x = "Day of Year")+
  scale_y_continuous(
    limits = c(0, 1.27e6),
    breaks = c(0, 6.25e5, 1.25e6),
    labels = c("0", "625K", "1.25M"))+
  # scale_y_continuous(
  #    limits = c(0, NA),
  #   # breaks = c(0, 7.5e5, 1.5e6),
  #   # labels = c("0", "750K", "1.5M"))+
  #   labels = label_number(scale_cut = cut_short_scale())) +
  #   #labels = label_number(scale_cut = cut_short_scale(), accuracy = 0.01)) +
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  scale_fill_manual(values = solar_colors)+
  scale_color_manual(values = solar_colors)+
  theme(legend.position = "none")
methanotroph_surface


# 3. box plot
methanotroph_surface_box <- methanotroph_surface_data %>% 
  ggplot(aes(x = solar_progress, y = total_abundance, color = solar_progress)) +
  geom_boxplot(outlier.shape = NA, color = "black") +
  geom_jitter(aes(color = solar_progress, fill = solar_progress), width = 0.2, size = 2, shape = 21) +
  scale_fill_manual(values = solar_colors)+
  scale_color_manual(values = c("black", "black"))+
  scale_y_continuous(
    limits = c(0, 1.27e6),
    breaks = c(0, 6.25e5, 1.25e6),
    labels = c("0", "625K", "1.25M"))+
  # scale_y_continuous(
  #   limits = c(0, 1.27e6),
  #   breaks = c(0, 2.5e5, 7.5e5, 1.25e6),
  #   labels = c("0","250K", "750K", "1.25M"))+
  #   # breaks = c(0, 5e5, 1e6, 1.5e6),
  #   #labels = label_number(scale_cut = cut_short_scale())) +
  theme_classic()+
  theme(axis.text.x = element_text(size = 8,colour = "black"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none") +
  # stat_compare_means(method = "wilcox.test",
  #                    #comparisons = list(c("FPV", "Open")),
  #                    label = "p.format",
  #                    size = 3,
  #                    label.y.npc = 0.9,
  #                    fontface = "italic")
  #label.y = c(8000, 100000, 500000, 400000, 400000), # get pvalue
 annotate("text", x = 2, y = 1.25e6, label = "p = 0.021", size = 2.822,  fontface = "italic") # add p value
methanotroph_surface_box

##### bottom methanogens #####
# 1. statistics
# 1a. abundance stats
methanogen_bottom_sum <- water_ch4_cyclers_df %>%
  dplyr::filter(Depth_Class == "Bottom Water",
                CH4_Cycler == "Methanogen") %>% 
  group_by(solar_progress, JDate) %>% 
  dplyr::summarize(sd_meth = sd(total_abundance),
                   mean_meth = mean(total_abundance))
methanogen_bottom_sum
max(methanogen_bottom_sum$mean_meth)

  
# 1b. linear mixed model
methanogen_bottom_data <- water_ch4_cyclers_df %>%
  dplyr::filter(Depth_Class == "Bottom Water",
                CH4_Cycler == "Methanogen") 
methanogen_bottom_data$Pond <- as.factor(methanogen_bottom_data$Pond)

bottom_methanogen_model <- lmer(total_abundance ~ solar_progress + JDate + (1|Pond), data = methanogen_bottom_data)
summary(bottom_methanogen_model) # 0.514
emmeans::emmeans(bottom_methanogen_model, pairwise ~ solar_progress) # 0.514

max(methanogen_bottom_data$total_abundance)

# 2. plot bottom water methanogens overtime 
methanogen_bottom <- methanogen_bottom_sum %>% 
  ggplot(aes(x = JDate, y = mean_meth, fill = solar_progress, color = solar_progress)) +
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = JDate, ymin = mean_meth - sd_meth, ymax = mean_meth + sd_meth), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  labs(y= "Bottom Methanogen<br>Abundance (Cells mL<sup>-1</sup>)", x = "Day of Year")+
  scale_y_continuous(
    limits = c(NA, 1.1e5),
    breaks = c(0, 5e4, 1e5),
    labels = label_number(scale_cut = cut_short_scale())) +
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  scale_fill_manual(values = solar_colors)+
  scale_color_manual(values = solar_colors) +
  theme(legend.position = "none")#+
  #scale_y_continuous(limits = c(0,11), breaks = c(0, 5, 10))+
  #scale_x_continuous(limits = c(170,250))
methanogen_bottom


# 3. box plot
methanogen_bottom_box <- methanogen_bottom_data %>% 
  ggplot(aes(x = solar_progress, y = total_abundance, color = solar_progress)) +
  geom_boxplot(outlier.shape = NA, color = "black") +
  geom_jitter(aes(color = solar_progress, fill = solar_progress), width = 0.2, size = 2, shape = 21) +
  scale_fill_manual(values = solar_colors)+
  scale_color_manual(values = c("black", "black"))+
  scale_y_continuous(
    limits = c(NA, 1.1e5),
    breaks = c(0, 5e4, 1e5),
    labels = label_number(scale_cut = cut_short_scale())) +
  theme_classic()+
  theme(axis.text.x = element_text(size = 8,colour = "black"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none") +
  # stat_compare_means(method = "wilcox.test",
  #                    #comparisons = list(c("FPV", "Open")),
  #                    label = "p.format",
  #                    size = 3,
  #                    label.y.npc = 0.9,
  #                    fontface = "italic")
                     #label.y = c(8000, 100000, 500000, 400000, 400000), # get pvalue
 annotate("text", x = 2, y = 1e5, label = "p = 0.514", size = 2.822,  fontface = "italic")
methanogen_bottom_box

##### bottom water methanotrophs #####
# 1. statistics
# 1a. calculate abundance stats
methanotroph_bottom_sum <- water_ch4_cyclers_df %>%
  dplyr::filter(Depth_Class == "Bottom Water",
                CH4_Cycler == "Methanotroph") %>% 
  group_by(solar_progress, JDate) %>% 
  dplyr::summarize(sd_meth = sd(total_abundance),
                   mean_meth = mean(total_abundance))
methanotroph_bottom_sum
max(methanotroph_bottom_sum$mean_meth)

# calculate abundances between treatments 
methanotroph_bottom_sum_text <- water_ch4_cyclers_df %>%
  dplyr::filter(Depth_Class == "Bottom Water",
                CH4_Cycler == "Methanotroph") %>% 
  group_by(solar_progress) %>% 
  dplyr::summarize(sd_meth = sd(total_abundance/1e5),
                   mean_meth = mean(total_abundance/1e5),
                   .groups = "drop")
methanotroph_bottom_sum_text
max(methanotroph_bottom_sum_text$mean_meth)

# 1b. linear mixed model 
methanotroph_bottom_data <- water_ch4_cyclers_df %>%
  dplyr::filter(Depth_Class == "Bottom Water",
                CH4_Cycler == "Methanotroph") 
methanotroph_bottom_data$Pond <- as.factor(methanotroph_bottom_data$Pond)

bottom_methanotroph_model <- lmerTest::lmer(total_abundance ~ solar_progress + JDate + (1|Pond), data = methanotroph_bottom_data)
summary(bottom_methanotroph_model) # 0.00272 **
emmeans::emmeans(bottom_methanotroph_model, pairwise ~ solar_progress) # 0.0400

# 2. plot bottom water methanotrophs
methanotroph_bottom <- methanotroph_bottom_sum %>% 
  ggplot(aes(x = JDate, y = mean_meth, fill = solar_progress, color = solar_progress)) +
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = JDate, ymin = mean_meth - sd_meth, ymax = mean_meth + sd_meth), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  labs(y= "Bottom Methanotroph<br>Abundance (Cells mL<sup>-1</sup>)", x = "Day of Year")+
  scale_y_continuous(
    breaks = c(0, 6.25e5, 1.25e6),
    labels = c("0", "625K", "1.25M"))+
  # scale_y_continuous(
  #   breaks = c(0, 2.5e5, 7.5e5, 1.25e6),
  #   labels = c("0", "250K", "750K", "1.25M"))+
  # labels = label_number(scale_cut = cut_short_scale(), accuracy = 0.01)) +
  # scale_y_continuous(
  #   
  #   limits = c(0, 1.25e6),
  #   breaks = c(0, 5e5, 1e6, 1.25e6),
  #   labels = label_number(scale_cut = cut_short_scale())) +
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  scale_fill_manual(values = solar_colors)+
  scale_color_manual(values = solar_colors)+
  theme(legend.position = "none")
methanotroph_bottom



# 3. box plot
methanotroph_bottom_box <- methanotroph_bottom_data %>% 
  ggplot(aes(x = solar_progress, y = total_abundance, color = solar_progress)) +
  geom_boxplot(outlier.shape = NA, color = "black") +
  geom_jitter(aes(color = solar_progress, fill = solar_progress), width = 0.2, size = 2, shape = 21) +
  scale_fill_manual(values = solar_colors) +
  scale_color_manual(values = c("black", "black"))+
  scale_y_continuous(
    breaks = c(0, 6.25e5, 1.25e6),
    labels = c("0", "625K", "1.25M"))+
  # scale_y_continuous(
  #   breaks = c(0, 2.5e5, 7.5e5, 1.25e6),
  #   labels = c("0","250K", "750K", "1.25M"))+
    #labels = label_number(scale_cut = cut_short_scale())) +
  theme_classic()+
  theme(axis.text.x = element_text(size = 8,colour = "black"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none") +#+
  # stat_compare_means(method = "wilcox.test",
  #                    #comparisons = list(c("FPV", "Open")),
  #                    label = "p.format",
  #                    size = 3,
  #                    label.y.npc = 0.9,
  #                    fontface = "italic")
  #                    #label.y = c(8000, 100000, 500000, 400000, 400000), # get pvalue
 annotate("text", x = 2, y = 1.25e6, label = "p = 0.040", size = 2.822,  fontface = "italic") # add p value
methanotroph_bottom_box


##### sediment methanogen #####
# calculate stats
# 1a. calcualte abundance
methanogen_sed_sum <- sed_ch4_cyclers_df %>%
  dplyr::filter(CH4_Cycler == "Methanogen") %>% 
 # group_by(solar_progress) %>% # when this is commented out then roughly 20% of community is methanogens
 # group_by(solar_progress) %>%  # when this is run then fpv has 18.9% methanogens and controls have more with 20.5%
  group_by(solar_progress, JDate) %>% 
  dplyr::summarize(sd_meth = sd(rel_abundance),
                   mean_meth = mean(rel_abundance))
methanogen_sed_sum
max(methanogen_sed_sum$mean_meth)

# 1b. linear mixed model 
methanogen_sed_data <- sed_ch4_cyclers_df %>%
  dplyr::filter(CH4_Cycler == "Methanogen") 
methanogen_sed_data$Pond <- as.factor(methanogen_sed_data$Pond)

sed_methanogen_model <- lmer(rel_abundance ~ solar_progress + JDate + (1|Pond), data = methanogen_sed_data)
summary(sed_methanogen_model) # 0.6678 
emmeans::emmeans(sed_methanogen_model, pairwise ~ solar_progress) # 0.667

# plot sediment methanogens
methanogen_sed <- methanogen_sed_sum %>% 
  ggplot(aes(x = JDate, y = mean_meth, fill = solar_progress, color = solar_progress)) +
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = JDate, ymin = mean_meth - sd_meth, ymax = mean_meth + sd_meth), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  scale_fill_manual(values = solar_colors)+
  scale_color_manual(values = solar_colors)+
  scale_y_continuous(
    limits = c(.08, .3),
    breaks = c(0.1, .2, .30),
    labels = c("0.10", "0.20", "0.30"))+
    #labels = label_number(scale_cut = cut_short_scale())) +
  labs(y= "Sediment Methanogen<br> Abundance (%)", x = "Day of Year")+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_text(size = 8,colour = "black"),
        axis.title.y = element_markdown(size = 8))+
  theme(legend.position = "none")
methanogen_sed


# 3. box plot
methanogen_sed_box <- methanogen_sed_data %>% 
  ggplot(aes(x = solar_progress, y = rel_abundance, color = solar_progress)) +
  geom_boxplot(outlier.shape = NA, color = "black") +
  geom_jitter(aes(color = solar_progress, fill = solar_progress), width = 0.2, size = 2, shape = 21) +
  scale_fill_manual(values = solar_colors)+
  scale_color_manual(values = c("black", "black"))+
  scale_y_continuous(
    limits = c(.08, .3),
    breaks = c(0.1, .2, .30),
    labels = c("0.10", "0.20", "0.30"))+
  # scale_y_continuous(
  #   #limits = c(.13, .3),
  #   breaks = breaks_extended(n = 4),
  #   #breaks = c(.15, .2, .25, .30),
  #   labels = label_number(scale_cut = cut_short_scale(), accuracy = 0.01)) +
  theme_classic()+
  theme(axis.text.x = element_text(size = 8,colour = "black"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none") +
  # stat_compare_means(method = "wilcox.test",
  #                    #comparisons = list(c("FPV", "Open")),
  #                    label = "p.format",
  #                    size = 3,
  #                    label.y.npc = 0.9,
  #                    fontface = "italic")
  # label.y = c(8000, 100000, 500000, 400000, 400000) # get pvalue
 annotate("text", x = 2, y = .3, label = "p = 0.667", size = 2.822,  fontface = "italic")
methanogen_sed_box


##### sediment methanotrophs #####
#1. calculate stats
#1a. calclate abundance
methanotroph_sed_sum <- sed_ch4_cyclers_df %>%
  dplyr::filter(CH4_Cycler == "Methanotroph") %>% 
  # group_by(solar_progress) %>% # when this is commented out then roughly 6% of community is methanotrophs
  #group_by(solar_progress) %>%  # when this is run then fpv has 6% methanotrophs and controls have more with 7% methanotrophs
  group_by(solar_progress, JDate) %>% 
  dplyr::summarize(sd_meth = sd(rel_abundance),
                   mean_meth = mean(rel_abundance))
methanotroph_sed_sum
max(methanotroph_sed_sum$mean_meth)

#1b linear mixed model
methanotroph_sed_data <- sed_ch4_cyclers_df %>%
  dplyr::filter(CH4_Cycler == "Methanotroph") 
methanotroph_sed_data$Pond <- as.factor(methanotroph_sed_data$Pond)

sed_methanotroph_model <- lmerTest::lmer(rel_abundance ~ solar_progress + JDate + (1|Pond), data = methanotroph_sed_data)
summary(sed_methanotroph_model) # 0.39913 
emmeans::lsmeans(sed_methanotroph_model, pairwise ~ solar_progress) #0.4401
  

# 2. plot sed methanotrophs overtime 
methanotroph_sed<- methanotroph_sed_sum %>% 
  ggplot(aes(x = JDate, y = mean_meth, fill = solar_progress, color = solar_progress)) +
  geom_line()+
  geom_point()+
  theme_classic()+
  geom_errorbar(aes(x = JDate, ymin = mean_meth - sd_meth, ymax = mean_meth + sd_meth), width = 0, color = "black")+
  geom_point(size = 3, shape = 16)+
  geom_point(size = 3, shape = 1, color = "black")+
  labs(y= "Sediment Methanotroph<br>Abundance (%)", x = "Day of Year")+
  theme(axis.text.x =  element_text(size = 8,colour = "black"),
         axis.text.y =  element_text(size = 8,colour = "black"),
         axis.title.x = element_blank(),
        axis.title.y = element_markdown(size = 8))+
  scale_fill_manual(values = solar_colors)+
  scale_color_manual(values = solar_colors)+
  scale_y_continuous(
    limits = c(0.02, .15),
    breaks = c(0.05, 0.10, 0.15),
    labels = label_number(scale_cut = cut_short_scale())) +
  theme(legend.position = "none")
methanotroph_sed


# box plot
methanotroph_sed_box <- methanotroph_sed_data %>% 
  ggplot(aes(x = solar_progress, y = rel_abundance, color = solar_progress)) +
  geom_boxplot(outlier.shape = NA, color = "black") +
  geom_jitter(aes(color = solar_progress, fill = solar_progress), width = 0.2, size = 2, shape = 21) +
  scale_fill_manual(values = solar_colors)+
  scale_color_manual(values = c("black", "black"))+
  scale_y_continuous(
    limits = c(0.02, .15),
    breaks = c(0.05, 0.10, 0.15),
    labels = label_number(scale_cut = cut_short_scale())) +
  theme_classic()+
  theme(axis.text.x = element_text(size = 8, colour = "black"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none") +#+
  # stat_compare_means(method = "wilcox.test",
  #                    #comparisons = list(c("FPV", "Open")),
  #                    label = "p.format",
  #                    size = 3,
  #                    label.y.npc = 0.9,
  #                    fontface = "italic")
  #                    #label.y = c(8000, 100000, 500000, 400000, 400000), # get pvalue
 annotate("text", x = 2, y = .15, label = "p = 0.440", size = 2.822,  fontface = "italic") # add p value
methanotroph_sed_box


## plot all together
fig2 <- methanotroph_surface + methanotroph_surface_box + methanogen_surface + methanogen_surface_box +
  methanotroph_bottom + methanotroph_bottom_box + methanogen_bottom + methanogen_bottom_box +
  methanotroph_sed + methanotroph_sed_box + methanogen_sed + methanogen_sed_box +
  plot_layout(nrow = 6, ncol = 2,
              widths = c(10, 4)) +
  plot_annotation(tag_levels = "A", tag_suffix = '.') +
  theme(
    plot.tag = element_text(size = 8))
fig2



#export the figure
ggsave(fig2, width = 6.5, height = 8.5, units = "in", filename = "figures/Fig_2/fig2.png") # could only save at height 8.5, orignally h = 8

```
We have created main text `fig2` which shows the abundance of water column (absolute abundance) and sediment (relative abundance) methanotrophs and methanogens over time between FPV and Open ponds. 

Each row corresponds to depth (surface water, bottom water, or sediments) and methane cycler (methanotroph or methanogen). The left panel demonstrates the temporal relationships while the right panel is a box plot comparison with linear mixed effects statistical analysis between FPV and Open ponds.


# Figure 3
Now plot beta diversity with Bray-Curtis dissimilarity matrix for all methane cyclers (methanogens and methanotrophs) in water column and sediments

### Fig 3A: Water PCoA 
```{r fig3a-pcoa-water}
# water methane cyclers

# Calculate Bray-Curtis Dissimilarity 
water_BC_pcoa <- 
  ordinate(
    physeq = water_ch4_cyclers_physeq,
    method = "PCoA",
    distance = "bray", 
    binary = FALSE
  )



#### Grab the data for the plot 
water_all_ord_df <- 
  plot_ordination(
  physeq = water_ch4_cyclers_physeq,
  ordination = water_BC_pcoa,
  color = "solar_progress",
  shape = "Pond",
  justDF = TRUE)

# now lets mutate the columns
water_all_ord_df <- water_all_ord_df %>% 
dplyr::mutate(
    solar_progress = recode(solar_progress, "FPV" = "FPV", "No FPV" = "Open"), # solar progress
    Depth_Class = recode(Depth_Class,  # depth class
      "S" = "Surface Water",
      "B" = "Bottom Water"))


### Now, plot Figure 3A: WATER 
fig3a_water_pcoa <- 
  ggplot(data = water_all_ord_df, 
       aes(x = Axis.1, 
           y = Axis.2,
           color = solar_progress,
           shape = Pond)) + 
  geom_point(size = 3, alpha = 0.8, stroke = 0.8) +
  scale_shape_manual(values = pond_shapes) + 
  scale_color_manual(values = solar_colors) +
  labs(color = "Treatment",
       shape = "Pond",
       x = "Axis.1 [27.4%]",
       y = "Axis.2 [11.3%]",
       title = expression("Water CH"[4]*" Cyclers")) + 
  guides(
    color = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5)),
    shape = guide_legend(
      nrow = 2,
      byrow = TRUE,
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5))
  ) +
  theme_classic() +
  theme(legend.position = "bottom",
        #legend.spacing = unit(0, "cm"),
        legend.box.background = element_blank(),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        plot.title = element_text(hjust = 0.5, size = 9, colour = "black"),
        axis.text.x = element_text(size = 8, colour = "black"),
        axis.text.y = element_text(size = 8, colour = "black"),
        axis.title.x = element_text(size = 8, colour = "black"),
        axis.title.y = element_text(size = 8, colour = "black"))

# Show the plot
fig3a_water_pcoa
### Sophia's plot - for axes labels
# PCoA of water samples color by treatment shape by pond
# s1a_water_pcoa <- plot_ordination(
#   physeq = water_physeq_24,
#   ordination = water_BC_pcoa,
#   color = "solar_progress",
#   shape = "Pond",
#   title = "Water Column Methane Cyclers") +
#   geom_point(size = 5, alpha = 0.5,
#              aes(fill = solar_progress, color = solar_progress, shape = Pond)) +
#   scale_fill_manual(values = solar_colors) +
#   scale_color_manual(values = solar_colors) +
#   scale_shape_manual(values = pond_shapes) +
#   guides(color = guide_legend(nrow = 1,
#                               title = NULL,
#                               override.aes = list(size = 2.7)),
#          fill = "none",
#          shape = guide_legend(nrow = 2,
#                               byrow = TRUE,
#                               title = NULL,
#                               override.aes = list(size = 2.7))) +
#   theme_classic() +
#   theme(
#     legend.position = c(0.01, 0.01),  # inside bottom-left
#     legend.justification = c(.01, .01),
#     legend.spacing = unit(0.01, "cm"),
#     legend.spacing.x = unit(0.1, "cm"),
#     legend.background = element_rect(color = NA, fill = NA),
#     legend.key.width = unit(0.2, "cm"),
#     legend.key.height = unit(0.4, "cm"),
#     legend.text = element_text(size = 6),
#     legend.box.just = "center",
#     legend.box.background = element_rect(size = 0.2, linetype = "solid", color = "black"),
#     legend.margin = margin(1, 2, 1, 1))
#
# Plot it
# s1a_water_pcoa
```

### Fig 3B: Sediment PCoA
```{r fig3b-pcoa-sediments}
# Calculate Bray-Curtis Dissimilarity 
scaled_sed_BC_pcoa <- 
  ordinate(
    physeq = sed_ch4_cyclers_physeq,
    method = "PCoA",
    distance = "bray", 
    binary = FALSE
  )


#### Grab the data for the plot 
sed_all_ord_df <- 
  plot_ordination(
  physeq = sed_ch4_cyclers_physeq,
  ordination = scaled_sed_BC_pcoa,
  color = "solar_progress",
  shape = "Pond",
  justDF = TRUE)

# update metadata for plotting 
sed_all_ord_df <- sed_all_ord_df %>% 
dplyr::mutate(
    solar_progress = recode(solar_progress, "FPV" = "FPV", "No FPV" = "Open"), # solar progress
    Depth_Class = recode(Depth_Class,  # depth class
      "S" = "Surface Water",
      "B" = "Bottom Water"))

# Now plot it! 
fig3b_sed_pcoa <- 
  ggplot(data = sed_all_ord_df, 
       aes(x = Axis.1, 
           y = Axis.2,
           color = solar_progress,
           shape = Pond)) + 
  geom_point(size = 3, alpha = 0.8, stroke = 0.8) +
  scale_shape_manual(values = pond_shapes) + 
  scale_color_manual(values = solar_colors) +
  labs(color = "Treatment",
       shape = "Pond",
       x = "Axis.1 [32.5%]",
       y = "Axis.2 [15.8%]",
       title = expression("Sediment CH"[4]*" Cyclers")) + 
  guides(
    color = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5)),
    shape = guide_legend(
      nrow = 2,
      byrow = TRUE,
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5))
  ) +
  theme_classic() +
  theme(legend.position = "bottom",
        #legend.spacing = unit(0, "cm"),
        legend.box.background = element_blank(),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        plot.title = element_text(hjust = 0.5, size = 9, colour = "black"),
        axis.text.x = element_text(size = 8, colour = "black"),
        axis.text.y = element_text(size = 8, colour = "black"),
        axis.title.x = element_text(size = 8, colour = "black"),
        axis.title.y = element_text(size = 8, colour = "black"))

# Show the plot
fig3b_sed_pcoa



# Sophia's plot 
# PCoA of sediments color by treatment shaped by pond
# s1b_sed_pcoa <-
#   plot_ordination(
#   physeq = sed_ch4_cyclers_physeq,
#   ordination = scaled_sed_BC_pcoa,
#   color = "solar_progress",
#   shape = "Pond",
#   title = "Sediment Methane Cyclers") +
#   geom_point(size = 5, alpha = 0.5,
#              aes(color = solar_progress, fill = solar_progress, shape = Pond)) +
#   scale_color_manual(values = solar_colors) +
#   scale_fill_manual(values = solar_colors) +
#   scale_shape_manual(values = pond_shapes) +
#   guides(color = "none",
#          fill = "none",
#          shape = "none") +
#   theme_classic()
#   # theme(
#   # legend.position = c(0.82, 0.01),  # inside bottom-left
#   # legend.justification = c(0, 0),  # anchor the legend's top-left corner there
#   # legend.spacing = unit(0.1, "cm"),
#   # legend.background = element_rect(color = NA, fill = NA),
#   # legend.box.background = element_rect(size = 0.1, linetype = "solid", color = "black"),
#   # legend.text = element_text(size = 6),
#   # legend.margin = margin(2, 2, 2, 2))
# s1b_sed_pcoa

# ggsave(s1b_sed_pcoa, width = 8, height = 7, units = "in",
#         filename = "figures/s1b/s1b_sed_pcoa.png")
```
Sediment samples are still distinct from other and separate along first axis

### Save Figure 3
```{r fig-3}
### Final Plot for Submission 
plot_fig3 <- 
  fig3a_water_pcoa + theme(plot.title = element_text(margin = margin(b = 0))) + 
  fig3b_sed_pcoa + theme(plot.title = element_text(margin = margin(b = 0))) +
  plot_annotation(tag_levels = "A", tag_suffix = ".") + 
    plot_layout(guides = "collect") &
  theme(
    plot.tag = element_text(size = 8, colour = "black"),
    legend.position = "bottom",
    # legend.title = element_text(size = 9),
    # legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm"),
    legend.spacing.x = unit(0.7, "cm"),
    legend.margin = margin(t = -5, unit = "pt")
  )
plot_fig3
  
ggsave(plot_fig3, width = 6.5, height = 4.5, dpi = 300,
        filename = "figures/Fig_3/Fig_3.png")
```



## Fig 3: PERMANOVA

PERMANOVA (Permutational Multivariate Analysis of Variance) is a non-parametric, permutation-based test used to compare groups of objects based on a distance matrix. The goal is to test the null hypothesis that the centroids and dispersion of groups are equivalent in the space defined by the dissimilarity measure. 

```{r fig3-water-sed-permanova}
#### WATER COLUMN: All methane cyclers
# calculate Bray-Curtis PERMANOVA using phyloseq distance
water_bray <- phyloseq::distance(water_ch4_cyclers_physeq, method = "bray", binary = FALSE)

# pull out metadata 
water_metadata <- water_ch4_cyclers_physeq %>%
  sample_data() %>%
  data.frame()


#### SEDIMENT: All methane cyclers
# calculate Bray-Curtis PERMANOVA using phyloseq distance
sed_bray <- phyloseq::distance(sed_ch4_cyclers_physeq, method = "bray", binary = FALSE)

# pull out metadata 
sed_metadata <- sed_ch4_cyclers_physeq %>%
  sample_data() %>%
  data.frame()
```


### Water 

Here we are performing a PERMANOVA on the water column methane cyclers throughout the entire sampling season.

```{r permanova-water}
# Permutational Multivariate Analysis of Variance Using Distance Matrices
# aka PERMANOVA using the adonis2 function from vegan 

#1. Test the individual terms for significance
# Testing if the centroids of solar progress are different: significant p = 0.001 ***
adonis2(water_bray ~ solar_progress, data = water_metadata, by = "terms")

# Now testing to see if centroids of depth_class are different: not significant p = 0.19
adonis2(water_bray ~ Depth_Class, data = water_metadata, by = "terms")

# Does pond matter? significant p = 0.001 ***
adonis2(water_bray ~ Pond, data = water_metadata, by = "terms")

# Does date matter? significant p = 0.001 ***
adonis2(water_bray ~ JDate, data = water_metadata, by = "terms")

#2. Test the terms together
# Now lets see the effect of each pond by date_collected and solar progress
water_permanova <- adonis2(water_bray ~ solar_progress * Pond * JDate, data = water_metadata, by = "terms"); water_permanova
```
With our PERMANOVA we find that treatment (solar_progress), day of year sampled (JDate - Julian date), and Pond is significant but depth class alone is not.

When we create a model with treatment, pond, and date these are all significant (p < 0.001 ***). 
- Solar progress is responsible for explaining  12.8% of variance and has a strong structuring effect on the water column community composition. This has the highest F value meaning that the between group differences are larger than within group variance (F = 9.4)

- Pond is also significant explains 15.5% of variance but is weaker than treatment (solar_progress) (F = 2.8). Even though it explains more variation than treatment, there is more variation in ponds than between ponds. 

- Date (JDate) is also strong and explains 8.6% of variance and is also weighted heavier (F = 6.3). The temporal effect is seen in the first axis as time progresses throughout the season.

- The interaction between treatment and date explains 5.9% of variance and is an important but moderate factor in its effect size for structuring community (F = 4.3)


Together the PERMANOVA explains only about half the variance seen with 48.9% remaining

### Sediments 
Here we are performing a PERMANOVA on the sediment methane cyclers throughout the entire sampling season
```{r permanova-sediments}
# Permutational Multivariate Analysis of Variance Using Distance Matrices
# aka PERMANOVA using the adonis2 function from vegan 

#1. Test the individual terms for significance
# Testing if the centroids of solar progress are different: significant p = 0.001 ***
adonis2(sed_bray ~ solar_progress, data = sed_metadata, by = "terms")

# Does pond matter? significant p = 0.001 ***
adonis2(sed_bray ~ Pond, data = sed_metadata, by = "terms")

# Does date matter? significant p = 0.001 ***
adonis2(sed_bray ~ JDate, data = sed_metadata, by = "terms")

#2. Test the terms together
# Now lets see the effect of each pond by date_collected and solar progress
sediment_permanova <- adonis2(sed_bray ~ solar_progress * Pond * JDate, data = sed_metadata, by = "terms"); sediment_permanova
```
With our PERMANOVA we find that treatment (solar_progress), day of year sampled (JDate - Julian date), and Pond is significant factors.

When we create a model with treatment, pond, and date these are all significant (p < 0.001 ***). 
- treatment = solar_progress is important for explaining 13.5% of variation and has the largest effect on structuring the community (F = 12.0). This explains the separation along the second axis

- pond explains the most variation (28.5%) and also has a substantial effect on structuring the community (F = 6.32) 

- Day of Year explains 10.7% of variation and also has strong temporal effect (F = 9.48) that shapes communities along first axis


This explains 63.9% of variance with 36.1% in the residuals. 

## Fig 3: Betadisper
We are running betadispr to test variances/dispersions

When computing PERMANOVA, we must also perform betadispr analysis when analyzing beta diversity in microbial ecology. We must do it after PERMANOVA because we need to look into the assumption of PERMANOVA which is the homogeneity of group dispersions aka variances. If this assumption is violated then the PERMANOVA results might be driven by dispersion rather than true differences in community composition.

Always run betadisper() and permutest() after PERMANOVA to test whether the groups have similar within-group variation.

It works by first taking ina. distance matrix and calculates the centroid of each group in multivariate space (note it does not test for significane). After computing within-group distances, we will run permutest() to see whether those dispersiosn differ significantly between groups using PERMANOVA

The permutest works like this: - null hypothesis (H0): all groups have equal multivariate dispersion and a compute a new F-statistic for each permutation.

The p-value is the proportion of permutations where the F is as extreme or more extreme than the observed F.
The result from permutest() is a robust non-parametric p-value testing whether dispersion differs across groups.
If p > 0.05 (not significant), the PERMANOVA result is reliable.

If p = 0.05 (significant), be cautious—group differences may be due to dispersion, not composition! However, not all is lost as we may expect this to be biologically true.

adonis - compares centroids to see if significant difference. betadispr - compares variance/distance from centroid

### Water
```{r betadispr-water}
# Homogeneity of Disperson test with beta dispr

## Bray-Curtis
betadispr_water_pond <- betadisper(water_bray, water_metadata$Pond)
betadispr_water_solar <- betadisper(water_bray, water_metadata$solar_progress)
betadispr_water_depth <- betadisper(water_bray, water_metadata$Depth_Class)
betadispr_water_JDate <- betadisper(water_bray, water_metadata$JDate)

# permutest() performs a non-parametric permutation test, which is robust and valid for the kind of data used in beta diversity analysis (e.g., dissimilarity matrices).
permutest(betadispr_water_pond) # not significant p = 0.256 

permutest(betadispr_water_solar) # significant p = 0.011 **

permutest(betadispr_water_depth) # not significant p = 0.417 

permutest(betadispr_water_JDate) # significant p = 0.006 **
```
With betadispr we find the PERMANOVA results are mostly valid where pond variation is consistent. The PERMANOVA and betadispr find that depth is not significant meaning that depth doesnt structure the community. This makes sense between the ponds are so shallow so they are more likely to be similar.

But treatment and date are significant meaning the differences may be due to dispersion and not composition. This could be because of temporal reasons where communities change over time with the season.

### Sediment 
```{r betadispr-sed}
# Homogeneity of Disperson test with beta dispr
## Bray-Curtis
betadispr_sed_pond <- betadisper(sed_bray, sed_metadata$Pond)
betadispr_sed_solar <- betadisper(sed_bray, sed_metadata$solar_progress)
betadispr_sed_JDate <- betadisper(sed_bray, sed_metadata$JDate)

# permutest() performs a non-parametric permutation test, which is robust and valid for the kind of data used in beta diversity analysis (e.g., dissimilarity matrices).
permutest(betadispr_sed_pond) # not significant p = 0.661

permutest(betadispr_sed_solar) # not significant p = 0.673

permutest(betadispr_sed_JDate) # not significant p = 0.162
```
With betadispr we find the PERMANOVA results are are valid as pond, treatment, and date are not significant but significant in the PERMANOVA. Thus our PERMANOVA result is reliable and the differences between groups are due to location/centroids of groups rather than differences in variation within groups 

# Figure 4

## ANCOMBC-II - Differential Abundance 

Now we will calculate the differential abundance between our water and sediment samples. First I will try to do the original methane cyclers in water and sediments. but i may also further break it down into sediment methane cycler type.

### Water
Note as of this step the results do change in that methanobacteriales is not differentially abundant anymore and we have a reduction in asv_1479 and 976. but all asvs belong to methylococcales order
```{r diff-abund-water}
# filter out for ASVs with zero variances; recommended to remove them as they are sparse and lowly abundant 

# remove bad taxa 
bad_taxa <- c("ASV_1071", "ASV_11590", "ASV_642", "ASV_231", "ASV_4564", "ASV_7355", "ASV_1920", "ASV_11514", "ASV_4168", "ASV_4105", "ASV_3231", "ASV_2424", "ASV_2964", "ASV_2205", "ASV_7156", "ASV_10120", "ASV_4270", "ASV_8273", "ASV_3557", "ASV_7055", "ASV_2530", "ASV_4099", "ASV_1677", "ASV_10767", "ASV_1252")

water_ch4_phy_bc <- water_ch4_cyclers_physeq %>%
  subset_taxa(., !(ASV %in% bad_taxa)) %>% 
  prune_taxa(taxa_sums(.) > 0,.)


# relevel solar_progress
water_ch4_cyclers_physeq@sam_data$solar_progress <- factor(water_ch4_cyclers_physeq@sam_data$solar_progress, levels = c("No FPV", "FPV"))

# run ancombc2 for water all methane cyclers
# water_ch4_asv_output <- ancombc2(data = water_ch4_phy_bc,
#                                  tax_level = "ASV", # Test for each phylum
#                                  fix_formula = "solar_progress", 
#                                  p_adj_method = "fdr", 
#                                  pseudo_sens = TRUE, # Run sensitivity test to make sure taxa isn't sensitive to psuedo-count choice
#                                  prv_cut = 0.05, # Prevalence filter of 1%
#                                  group = NULL, # Use Comp_Group_Hier as groups when doing pairwise comparisons
#                                  struc_zero = FALSE, # Do not detect structural zeroes
#                                  alpha = 0.05, # Significance threshold of 0.05
#                                  n_cl = 10, # Use 10 threads
#                                  verbose = FALSE, # Don't print verbose output
#                                  s0_perc = 0.05,
#                                  global = FALSE, # Run a global test (sorta like an ANOVA to first find if a given ASV is sig diff)
#                                  pairwise = FALSE) # Run pairwise tests between groups (sorta like a post-hoc test like Tukey)


# save(water_ch4_asv_output, file = "data/03_diff_abund/water_ch4_asv_output.RData")

load("data/03_diff_abund/water_ch4_asv_output.RData")


# plot ASV differential abundance
water_ch4_fsp <- 
  water_ch4_asv_output$res %>%
  select(taxon, starts_with("lfc"), starts_with("diff"), starts_with("passed_ss")) %>%
  pivot_longer(cols = !taxon, names_to = "metric", values_to = "value") %>%
  separate_wider_delim(cols = metric, delim = "_", names = c("variable", "Comparison"), too_many = "merge") %>%
  mutate(Comparison = str_remove(Comparison, "\\(Intercept\\)")) %>% 
  mutate(Comparison = str_remove(Comparison, "ss_")) %>%
  pivot_wider(id_cols = c("taxon","Comparison"), names_from = variable, values_from = value) %>%
  mutate(Comparison = str_remove(Comparison, "solar_progress"),
         Comparison = str_replace(Comparison, "_solar_progrss", ";")) %>%
  separate_wider_delim(Comparison, delim = ";", names = c("Ref1", "Ref2"), too_few = "align_start") %>%
  dplyr::filter(!is.na(Ref1) & Ref1 != "") %>%
  mutate(
    Ref2 = ifelse(is.na(Ref2), "No FPV", Ref2), # relevel with basegroup which is no solar 
    Comparison = paste0(Ref2, " : ", Ref1)) %>% 
  dplyr::filter(diff == 1, passed == 1, abs(lfc) > 1) %>% # 1.2 is good cut off to see effect, below we get ASVs with super low cell counts 
  select(ASV = taxon, Comparison, lfc, passed)


# join by tax table
clean_water_ch4 <- water_ch4_fsp %>% 
  left_join(., as.data.frame(water_ch4_cyclers_physeq@tax_table), 
            by = "ASV")

# plot log fold changes
clean_water_ch4 %>% 
  ggplot(aes(x = ASV, y = lfc, fill = Genus)) +
  geom_col() +
  #scale_fill_manual(values = phylum_colors) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom") + 
  ggtitle("Water Column ASV Log-fold Change in FPV Ponds") 

# create column for nice plotting of taxnomic name and asv
clean_asv24 <- clean_water_ch4 %>%
   mutate(
    label_tax = coalesce(Genus, paste0("f_", Family), paste0("o_", Order)),
    polished_tax = paste0(label_tax, " (", ASV, ")")) %>% 
  arrange(desc(lfc))

clean_asv24 <- clean_asv24 %>%
  mutate(
    polished_tax = ifelse(
      !is.na(Genus) & Genus != "",
      paste0(Genus, " (", ASV, ")"),
      ifelse(
        !is.na(Family) & Family != "",
        paste0("f_", Family, " (", ASV, ")"),
        ifelse(
          !is.na(Order) & Order != "",
          paste0("o_", Order, " (", ASV, ")"),
          ifelse(
            !is.na(Class) & Class != "",
            paste0("c_", Class, " (", ASV, ")"),
            paste0("p_", Phylum, " (", ASV, ")")
          )
        )
      )
    )
  ) %>% 
  arrange(desc(lfc))

# reorder factor levels
clean_asv24$polished_tax <- factor(clean_asv24$polished_tax, levels = as.list(clean_asv24$polished_tax))


# plot log fold changes
da2 <- clean_asv24 %>% 
  ggplot(aes(x = polished_tax, y = lfc, fill = Family)) +
  geom_bar(stat = "identity") +
  #scale_fill_manual(values = class_colors) +
  facet_wrap(Comparison ~.) +
  theme_classic() +
  coord_flip()+
  theme(
    legend.position = "bottom",
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank()) + 
  ggtitle("ASV Log-fold Change in Taxa 2024") 
da2

# plot differentially abundant ASVs overtime 

#1. tax glom at ASV level
water_ch4_asv_df_glom <- water_ch4_cyclers_physeq %>% 
  tax_glom(taxrank = "ASV") %>% 
  psmelt() %>% 
  mutate(
    solar_progress = recode(solar_progress, "Solar" = "FPV", "No FPV" = "Open"),
    Depth_Class = case_when(
      Depth_Class == "S" ~ "Surface Water",
      Depth_Class == "B" ~ "Bottom Water"),
    Depth_Class = factor(Depth_Class, levels = c("Surface Water", "Bottom Water")))

water_ch4_asv_df_glom$solar_progress <- factor(water_ch4_asv_df_glom$solar_progress,
                                               levels = c("FPV", "Open"))

# create list of differentially abundanct asvs, updated results
water_ch4_methanotrophs <- c("ASV_13", "ASV_141", "ASV_32", "ASV_44", "ASV_1479", "ASV_976", "ASV_1367", "ASV_828")
#water_ch4_methanotrophs <- c("ASV_13", "ASV_141", "ASV_32")

# now plot overtime
water_ch4_trophs <- water_ch4_asv_df_glom %>% 
  dplyr::filter(ASV %in% water_ch4_methanotrophs) %>% 
  dplyr::mutate(total_abundance = Abundance) %>%
  ggplot(aes(x = as.factor(JDate), y = total_abundance/1e6, color = solar_progress, shape = Pond)) +
  geom_line(aes(group = interaction(Pond, Depth_Class)), 
            alpha = 0.2) +
  geom_smooth(aes(group = solar_progress), se = FALSE) +
  geom_point(aes(shape = Pond), size = 2) +
  ggh4x::facet_grid2(Depth_Class~ASV) +
  scale_color_manual(values = solar_colors) +
  scale_shape_manual(values = pond_shapes) +
  labs(
    x = "Date Collected",
    y = "Total Cells per ml",
    title = "Differentially Abundant Methylococcales ASVs in FPV Ponds"
  ) +
  theme(legend.position = "bottom") +
  theme_bw()

# Show the plot 
water_ch4_trophs
```
When we look at the water column of just our methane cyclers, we see that there are only log fold change increases. It is no suprise that Methylococcales has 4 differentially abundant ASVs, but ASV 32 is a log fold change just shy of 3! I am kinda shocked that the Methanobacteriales ASV 1063 is differentially abundant in the water column of solar ponds...



```{r diff-abund-boxplots}
# Prepare the dataframe with only those 5 ASVs
diff_abund_df <- water_ch4_asv_df_glom %>% 
  # dplyr::filter(ASV %in% c("ASV_1063", "ASV_13", "ASV_141", "ASV_32", "ASV_44")) %>% 
  dplyr::filter(ASV %in% c("ASV_13", "ASV_141", "ASV_32", "ASV_44")) %>% 
  dplyr::group_by(
    JDate, Pond, Depth_Class, solar_progress,
    CH4_Cycler, Phylum, Class, Order, Family, Genus, ASV) %>%
  dplyr::summarise(
    total_abundance = sum(Abundance, na.rm = TRUE),
    .groups = "drop") %>%  
  # as.data.frame() %>%
  dplyr::mutate(Genus = ifelse(ASV== "ASV_13", Order, Genus),
                Genus = if_else(Genus == "Methanobacterium_B_963", 
                                "Methanobacterium_B", Genus),
                Genus = if_else(Genus == "Methylobacter_C_601751", 
                                "Methylobacter_C", Genus),
                facet_label = paste0("<i>", Genus, "</i><br>", ASV))

# Combine labels
#diff_abund_df$italicized_label <-
  #paste(diff_abund_df$italicized_label, diff_abund_df$ASV, sep = "\n")

# italicize facet labels
# diff_abund_df <- diff_abund_df %>%
#   mutate(
#     combined_label = paste0(
#       "italic(", Genus, ")*'\n' *'", ASV, "'"
#     )
#   )

# shapiro test
diff_abund_df %>%
  group_by(ASV, solar_progress) %>%
  summarise(
    shapiro_p = shapiro.test(total_abundance)$p.value,
    n = n()
  )

# Make Boxplots of the ASVs!
diffAbund_boxplots <- diff_abund_df %>%
  ggplot(aes(x = solar_progress, y = total_abundance,
             color = solar_progress)) + 
  geom_point(aes(shape = Pond),
             size = 2, alpha = 0.8, stroke = 0.8,
             position = position_jitterdodge(jitter.width = .5, dodge.width = .3)) +
  geom_boxplot(outlier.shape = NA, alpha = 0, color = "black", position = position_dodge(0.6)) + 
  labs(color = "Treatment", y = "Water Column<br>Abundance (Cells mL<sup>-1</sup>)") +
  ggh4x::facet_wrap2(~ facet_label, scales = "free_y", nrow = 2) +
  # facet_grid2(~ facet_label, scales = "free_y",
  #             strip = strip_themed(text_x = element_markdown(size = 9, face = "plain")))+
  # facet_wrap2(~ facet_label, scales = "free_y", nrow = 2,
  #             strip = strip_themed(text_x = element_markdown(size = 10, face = "plain")))+
  #facet_wrap2(~italicized_label, scales = "free_y", nrow = 2, labeller = label_parsed) + 
  scale_color_manual(values = solar_colors) +
  scale_shape_manual(values = pond_shapes) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale(), accuracy = 1)) +
  theme_classic() +
  # stat_compare_means(method = "wilcox.test", 
  #                    #comparisons = list(c("FPV", "Open")),
  #                    label = "p.format", 
  #                    #group.by = "combined_label",
  #                    size = 3,          
  #                    label.y.npc = 0.9,
  #                    fontface = "italic",
  #                    #label.y = c(8000, 100000, 500000, 400000, 400000),
  #                    label.x = c(1.75, 1.75, 1.75)) + 
  guides(
    color = "none",
    # color = guide_legend(
    #   order = 1,
    #   ncol= 2,
    #   title.position = "top",
    #   title.hjust = 0.5,
    #   override.aes = list(size = 2.5)),
    shape = guide_legend(
      nrow= 2,
      byrow = TRUE,
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5))) +
  theme_classic() +
  theme(legend.position = "bottom",
        legend.spacing = unit(0, "cm"),
        plot.title = element_text(hjust = 0.5),
        panel.background =  element_rect(color = 'black', size = 1),
        panel.grid = element_blank(),
        legend.background = element_rect(fill = "transparent", color = NA), # remove legend box
        legend.key = element_rect(fill = "transparent", color = NA),
        #legend.box.background = element_rect(fill='transparent', color = "transparent"), #transparent legend panel
        legend.box.just = "center",
        #strip.background = element_rect(colour = NA, fill = 'transparent'),
        plot.background = element_rect(fill = "transparent", color="transparent"),
        legend.key.size = unit(0.4, "cm"),
        legend.spacing.x = unit(0.2, "cm"),
      # legend.margin = margin(t = -5, unit = "pt"),
        strip.text = element_markdown(size = 8),
       axis.title.y = element_markdown(size = 8, colour = "black"),
       axis.title.x = element_blank(),
       axis.text.y = element_text(size = 8, colour = "black"),
       legend.title = element_text(size = 9, colour = "black"),
      legend.text = element_text(size = 8, colour = "black")); diffAbund_boxplots

# # get legend
# legend <- cowplot::get_plot_component(diffAbund_boxplots, "guide-box", return_all = TRUE)
# legend_full <- legend$grobs[[2]]

  
# Save the plot   
# ggsave(diffAbund_boxplots, width = 6, height = 4, dpi = 300,
#         filename = "figures/Fig_4/Fig_4.png")

##### ASV_13 Methylococcales #####
#1. Run stats
# 1a. calculate abundances
asv_13_sum <- diff_abund_df %>%
  dplyr::filter(ASV == "ASV_13") %>% 
  group_by(solar_progress, JDate) %>% # 
  dplyr::summarize(sd_meth = sd(total_abundance),
                   mean_meth = mean(total_abundance),
                   .groups = "drop")
asv_13_sum

# 1b. linear mixed model
asv_13_data <- diff_abund_df %>%
  dplyr::filter(ASV == "ASV_13")  
asv_13_data$Pond <- as.factor(asv_13_data$Pond)

asv_13_data_model <- lmerTest::lmer(total_abundance ~ solar_progress + JDate + (1|Pond), data = asv_13_data) # omitting jdate as term
summary(asv_13_data_model) #0.000119
emmeans::emmeans(asv_13_data_model, pairwise ~ solar_progress) #0.0135


# Make individual boxplots of the ASVs
ASV_13_methylococcales <- diff_abund_df %>%
  dplyr::filter(ASV == "ASV_13") %>% 
  ggplot(aes(x = solar_progress, y = total_abundance,
             color = solar_progress)) + 
  geom_point(aes(shape = Pond),
             size = 2, alpha = 0.8, stroke = 0.8,
             position = position_jitterdodge(jitter.width = .5, dodge.width = .3)) +
  geom_boxplot(outlier.shape = NA, alpha = 0, color = "black", position = position_dodge(0.6)) + 
  labs(color = "Treatment", y = "Water Column<br>Abundance (Cells mL<sup>-1</sup>)") +
  # facet_grid2(~ facet_label, scales = "free_y",
  #             strip = strip_themed(text_x = element_markdown(size = 9, face = "plain")))+
  # ggh4x::facet_wrap2(~ facet_label, scales = "free_y", nrow = 2,
  #                    strip = strip_vanilla(clip = "on")) +
  ggh4x::facet_grid2(. ~ facet_label) +
  #facet_wrap2(~italicized_label, scales = "free_y", nrow = 2, labeller = label_parsed) + 
  scale_color_manual(values = solar_colors) +
  scale_shape_manual(values = pond_shapes) +
  scale_y_continuous(
    breaks = c(0, 2e5, 4e5),
    labels = label_number(scale_cut = cut_short_scale(), accuracy = 1)) +
  # stat_compare_means(method = "wilcox.test", 
  #                    #comparisons = list(c("FPV", "Open")),
  #                    label = "p.format", 
  #                    #group.by = "combined_label",
  #                    size = 3,          
  #                    label.y.npc = 0.9,
  #                    fontface = "italic",
  #                    #label.y = c(8000, 100000, 500000, 400000, 400000),
  #                    label.x = c(1.75, 1.75, 1.75)) + 
  guides(
    # color = "none",
    color = guide_legend(
      #order = 1,
      ncol= 2,
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5)),
    shape = guide_legend(
      nrow= 2,
      byrow = TRUE,
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5))) +
  theme_classic() +
  theme(legend.position = "none",
        legend.spacing = unit(0, "cm"),
        plot.title = element_text(hjust = 0.5),
        panel.background =  element_rect(color = 'black', size = 1),
        panel.grid = element_blank(),
        legend.background = element_rect(fill = "transparent", color = NA), # remove legend box
        legend.key = element_rect(fill = "transparent", color = NA),
        legend.box.background = element_rect(fill='transparent', color = "transparent"), #transparent legend panel
        #legend.box.just = "center",
        #strip.background = element_rect(colour = NA, fill = 'transparent'),
        plot.background = element_rect(fill = "transparent", color="transparent"),
        legend.key.size = unit(0.4, "cm"),
        legend.spacing.x = unit(0.7, "cm"),
       legend.margin = margin(t = -5, unit = "pt"),
      strip.text = element_markdown(size = 8),
       axis.title.y = element_markdown(size = 8, colour = "black"),
       axis.title.x = element_blank(),
       axis.text.y = element_text(size = 8, colour = "black"),
       legend.title = element_text(size = 9, colour = "black"),
      legend.text = element_text(size = 8, colour = "black"))+
  annotate("text", x = 2, y = 3.9e5, label = "p = 0.014", size = 2.822,  fontface = "italic") 
ASV_13_methylococcales


##### ASV_32 Methylococcales #####
#1. Run stats
# 1a. calculate abundances
asv_32_sum <- diff_abund_df %>%
  dplyr::filter(ASV == "ASV_32") %>% 
  group_by(solar_progress, JDate) %>% # 
  dplyr::summarize(sd_meth = sd(total_abundance),
                   mean_meth = mean(total_abundance),
                   .groups = "drop")
asv_32_sum

# 1b. linear mixed model
asv_32_data <- diff_abund_df %>%
  dplyr::filter(ASV == "ASV_32")  
asv_32_data$Pond <- as.factor(asv_32_data$Pond)

asv_32_data_model <- lmerTest::lmer(total_abundance ~ solar_progress + JDate + (1|Pond), data = asv_32_data)
summary(asv_32_data_model) #0.0334
emmeans::emmeans(asv_32_data_model, pairwise ~ solar_progress) #0.0334


# Make individual boxplots of the ASVs
ASV_32_methyloparacoccus <- diff_abund_df %>%
  dplyr::filter(ASV == "ASV_32") %>% 
  ggplot(aes(x = solar_progress, y = total_abundance,
             color = solar_progress)) + 
  geom_point(aes(shape = Pond),
             size = 2, alpha = 0.8, stroke = 0.8,
             position = position_jitterdodge(jitter.width = .5, dodge.width = .3)) +
  geom_boxplot(outlier.shape = NA, alpha = 0, color = "black", position = position_dodge(0.6)) + 
  labs(color = "Treatment", y = "Water Column<br>Abundance (Cells mL<sup>-1</sup>)") +
  # facet_wrap2(~ facet_label, scales = "free_y", nrow = 2,
  #             strip = strip_themed(text_x = element_markdown(size = 9, face = "plain")))+
  # ggh4x::facet_wrap2(~ facet_label, scales = "free_y", nrow = 2, 
  #                    strip = strip_vanilla(clip = "on")) +
  ggh4x::facet_grid2(. ~ facet_label) +
  #facet_wrap2(~italicized_label, scales = "free_y", nrow = 2, labeller = label_parsed) + 
  scale_color_manual(values = solar_colors) +
  scale_shape_manual(values = pond_shapes) +
  scale_y_continuous(
    breaks = c(0, 1.5e5, 3e5),
    labels = label_number(scale_cut = cut_short_scale(), accuracy = 1)) +
  theme_classic() +
  # stat_compare_means(method = "wilcox.test", 
  #                    #comparisons = list(c("FPV", "Open")),
  #                    label = "p.format", 
  #                    #group.by = "combined_label",
  #                    size = 3,          
  #                    label.y.npc = 0.9,
  #                    fontface = "italic",
  #                    #label.y = c(8000, 100000, 500000, 400000, 400000),
  #                    label.x = c(1.75, 1.75, 1.75)) + 
  guides(
    #color = "none",
    color = guide_legend(
      #order = 1,
      ncol= 2,
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5)),
    shape = guide_legend(
      nrow= 2,
      byrow = TRUE,
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5))) +
  theme(legend.position = "none",
        legend.spacing = unit(0, "cm"),
        plot.title = element_text(hjust = 0.5),
        panel.background =  element_rect(color = 'black', size = 1),
        panel.grid = element_blank(),
        legend.background = element_rect(fill = "transparent", color = NA), # remove legend box
        legend.key = element_rect(fill = "transparent", color = NA),
        legend.box.background = element_rect(fill='transparent', color = "transparent"), #transparent legend panel
        #legend.box.just = "center",
        #strip.background = element_rect(colour = NA, fill = 'transparent'),
        plot.background = element_rect(fill = "transparent", color="transparent"),
        legend.key.size = unit(0.4, "cm"),
        legend.spacing.x = unit(0.7, "cm"),
        legend.margin = margin(t = -5, unit = "pt"),
        strip.text = element_markdown(size = 8),
       #axis.title.y = element_markdown(size = 8, colour = "black"),
       axis.title.x = element_blank(),
       axis.title.y = element_blank(),
       axis.text.y = element_text(size = 8, colour = "black"),
       legend.title = element_text(size = 9, colour = "black"),
      legend.text = element_text(size = 8, colour = "black"))+
  annotate("text", x = 2, y = 3e5, label = "p = 0.033", size = 2.822,  fontface = "italic") 
ASV_32_methyloparacoccus

##### ASV_44 Methylomonas #####
#1. Run stats
# 1a. calculate abundances
asv_44_sum <- diff_abund_df %>%
  dplyr::filter(ASV == "ASV_44") %>% 
  group_by(solar_progress, JDate) %>% # 
  dplyr::summarize(sd_meth = sd(total_abundance),
                   mean_meth = mean(total_abundance),
                   .groups = "drop")
asv_44_sum

# 1b. linear mixed model
asv_44_data <- diff_abund_df %>%
  dplyr::filter(ASV == "ASV_44")  
asv_44_data$Pond <- as.factor(asv_44_data$Pond)

asv_44_data_model <- lmerTest::lmer(total_abundance ~ solar_progress + JDate + (1|Pond), data = asv_44_data)
summary(asv_44_data_model) #0.1708 
emmeans::emmeans(asv_44_data_model, pairwise ~ solar_progress) #0.1708


# Make individual boxplots of the ASVs
ASV_44_methylomonas <- diff_abund_df %>%
  dplyr::filter(ASV == "ASV_44") %>% 
  ggplot(aes(x = solar_progress, y = total_abundance,
             color = solar_progress)) + 
  geom_point(aes(shape = Pond),
             size = 2, alpha = 0.8, stroke = 0.8,
             position = position_jitterdodge(jitter.width = .5, dodge.width = .3)) +
  geom_boxplot(outlier.shape = NA, alpha = 0, color = "black", position = position_dodge(0.6)) + 
  labs(color = "Treatment", y = "Water Column<br>Abundance (Cells mL<sup>-1</sup>)") +
  # ggh4x::facet_wrap2(~ facet_label, scales = "free_y", nrow = 2,
  #             strip = ggh4x::strip_themed(text_x = list(
  #               ggtext::element_markdown(size = 9, face = "plain"))))+
  ggh4x::facet_wrap2(~ facet_label, scales = "free_y", nrow = 2) +
  #facet_wrap2(~italicized_label, scales = "free_y", nrow = 2, labeller = label_parsed) + 
  scale_color_manual(values = solar_colors) +
  scale_shape_manual(values = pond_shapes) +
  scale_y_continuous(
    breaks = c(0, 2e5, 4e5),
    labels = label_number(scale_cut = cut_short_scale(), accuracy = 1)) +
  theme_classic() +
  # stat_compare_means(method = "wilcox.test", 
  #                    #comparisons = list(c("FPV", "Open")),
  #                    label = "p.format", 
  #                    #group.by = "combined_label",
  #                    size = 3,          
  #                    label.y.npc = 0.9,
  #                    fontface = "italic",
  #                    #label.y = c(8000, 100000, 500000, 400000, 400000),
  #                    label.x = c(1.75, 1.75, 1.75)) + 
  guides(
    color = "none",
    color = guide_legend(
      order = 1,
      ncol= 2,
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5)),
    shape = guide_legend(
      nrow= 2,
      byrow = TRUE,
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5))) +
  theme_classic() +
  theme(legend.position = "none",
        legend.spacing = unit(0, "cm"),
        plot.title = element_text(hjust = 0.5),
        panel.background =  element_rect(color = 'black', size = 1),
        panel.grid = element_blank(),
        legend.background = element_rect(fill = "transparent", color = NA), # remove legend box
        legend.key = element_rect(fill = "transparent", color = NA),
        legend.box.background = element_rect(fill='transparent', color = "transparent"), #transparent legend panel
        legend.box.just = "center",
        #strip.background = element_rect(colour = NA, fill = 'transparent'),
        plot.background = element_rect(fill = "transparent", color="transparent"),
        legend.key.size = unit(0.4, "cm"),
        legend.spacing.x = unit(0.2, "cm"),
      # legend.margin = margin(t = -5, unit = "pt"),
        strip.text = element_markdown(size = 8),
       #axis.title.y = element_markdown(size = 8, colour = "black"),
       axis.title.x = element_blank(),
       axis.title.y = element_blank(),
       axis.text.y = element_text(size = 8, colour = "black"),
       legend.title = element_text(size = 9, colour = "black"),
      legend.text = element_text(size = 8, colour = "black"))+
  annotate("text", x = 2, y = 4.5e5, label = "p = 0.17", size = 2.822,  fontface = "italic") 
ASV_44_methylomonas

##### ASV_141 Methylobacter_C #####
#1. Run stats
# 1a. calculate abundances
asv_141_sum <- diff_abund_df %>%
  dplyr::filter(ASV == "ASV_141") %>% 
  group_by(solar_progress, JDate) %>% # 
  dplyr::summarize(sd_meth = sd(total_abundance),
                   mean_meth = mean(total_abundance),
                   .groups = "drop")
asv_141_sum

# 1b. linear mixed model
asv_141_data <- diff_abund_df %>%
  dplyr::filter(ASV == "ASV_141")  
asv_141_data$Pond <- as.factor(asv_141_data$Pond)

asv_141_data_model <- lmerTest::lmer(total_abundance ~ solar_progress + JDate + (1|Pond), data = asv_141_data)
summary(asv_141_data_model) #0.000914 ***
emmeans::emmeans(asv_141_data_model, pairwise ~ solar_progress) #0.0238


# Make individual boxplots of the ASVs
ASV_141_methylobacterc <- diff_abund_df %>%
  dplyr::filter(ASV == "ASV_141") %>% 
  ggplot(aes(x = solar_progress, y = total_abundance,
             color = solar_progress)) + 
  geom_point(aes(shape = Pond),
             size = 2, alpha = 0.8, stroke = 0.8,
             position = position_jitterdodge(jitter.width = .5, dodge.width = .3)) +
  geom_boxplot(outlier.shape = NA, alpha = 0, color = "black", position = position_dodge(0.6)) + 
  labs(color = "Treatment", y = "Water Column<br>Abundance (Cells mL<sup>-1</sup>)") +
  # ggh4x::facet_grid2(~ facet_label, scales = "free_y", nrow = 2, 
  #                    strip = strip_vanilla(clip = "on")) +
  # facet_wrap2(~ facet_label, scales = "free_y", nrow = 2,
  #             strip = strip_themed(text_x = element_markdown(size = 9, face = "plain")))+
  #facet_wrap2(~italicized_label, scales = "free_y", nrow = 2, labeller = label_parsed) + 
  ggh4x::facet_grid2(. ~ facet_label) +
  scale_color_manual(values = solar_colors) +
  scale_shape_manual(values = pond_shapes) +
  scale_y_continuous(
    #breaks = c(0, 1.5e5, 3e5),
    labels = label_number(scale_cut = cut_short_scale(), accuracy = 1)) +
  theme_classic() +
  # stat_compare_means(method = "wilcox.test", 
  #                    #comparisons = list(c("FPV", "Open")),
  #                    label = "p.format", 
  #                    #group.by = "combined_label",
  #                    size = 3,          
  #                    label.y.npc = 0.9,
  #                    fontface = "italic",
  #                    #label.y = c(8000, 100000, 500000, 400000, 400000),
  #                    label.x = c(1.75, 1.75, 1.75)) + 
  guides(
    #color = "none",
    color = guide_legend(
      #order = 1,
      ncol= 2,
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5)),
    shape = guide_legend(
      nrow= 2,
      byrow = TRUE,
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5))) +
  theme_classic() +
  theme(legend.position = "none",
        legend.spacing = unit(0, "cm"),
        plot.title = element_text(hjust = 0.5),
        panel.background =  element_rect(color = 'black', size = 1),
        panel.grid = element_blank(),
        legend.background = element_rect(fill = "transparent", color = NA), # remove legend box
        legend.key = element_rect(fill = "transparent", color = NA),
        legend.box.background = element_rect(fill='transparent', color = "transparent"), #transparent legend panel
        #legend.box.just = "center",
        #strip.background = element_rect(colour = NA, fill = 'transparent'),
        plot.background = element_rect(fill = "transparent", color="transparent"),
        legend.key.size = unit(0.4, "cm"),
        legend.spacing.x = unit(0.7, "cm"),
       legend.margin = margin(t = -5, unit = "pt"),
        strip.text = element_markdown(size = 8),
       #axis.title.y = element_markdown(size = 8, colour = "black"),
       axis.title.x = element_blank(),
      axis.title.y = element_blank(),
       axis.text.y = element_text(size = 8, colour = "black"),
       legend.title = element_text(size = 9, colour = "black"),
      legend.text = element_text(size = 8, colour = "black"))+
    annotate("text", x = 2, y = 1.3e5, label = "p = 0.024", size = 2.822,  fontface = "italic") 
ASV_141_methylobacterc

# # extract legend
# legend <- cowplot::get_plot_component(ASV_13_methylococcales, "guide-box", return_all = TRUE) # can also do right
# legend_only <- legend[[3]]


fig4 <-  ASV_13_methylococcales +
  ASV_32_methyloparacoccus +
  ASV_141_methylobacterc +
  #plot_spacer() +
  plot_layout(ncol = 3, guides = "collect", widths = c(1,1,1)) & 
  theme(
    legend.position = "bottom")
    #legend.justification = c(0,.2))
    #legend.box.margin = margin(2, 2, 2, 2.5))

fig4





p13  <- ASV_13_methylococcales + theme(legend.position = "bottom")
p32  <- ASV_32_methyloparacoccus + theme(legend.position = "bottom")
p141 <- ASV_141_methylobacterc + theme(legend.position = "bottom")

fig4 <- (p13 + p32 + p141) +
  plot_layout(ncol = 3, guides = "collect") &
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.box.just = "center",
    legend.direction = "horizontal",
    legend.title.position = "top",
    legend.title.align = 0.5,
    legend.title = element_text(margin = margin(b = 0, unit = "pt")))

  # theme(legend.position = "bottom", legend.box.just = "center", legend.box = "horizontal")
fig4



fig4 <- fig4 &
  theme(
    legend.key.height = unit(0.35, "cm"),
    legend.key.width  = unit(0.35, "cm"),
    legend.spacing.x  = unit(0.4, "cm")
  )
fig4 




## plot all together
fig4 <- ASV_13_methylococcales +
  ASV_32_methyloparacoccus +
  ASV_141_methylobacterc +
  #plot_spacer() +
  plot_layout(ncol = 3, guides = "collect", widths = c(1,1,1)) &
  theme(
    legend.position = "bottom",
    legend.spacing.x = unit(1, "cm"),
    legend.margin = margin(t = 8, unit = "pt"),
   # legend.justification = c(-30,.2),
   # legend.box.just = "center",
    legend.box.margin = margin(0, 0, 0, 0))

fig4


# Save the plot   
ggsave(fig4, width = 6.5, height = 4.5, dpi = 300,
        filename = "figures/Fig_4/Fig_4.png")


#or
library(ggplot2)
library(patchwork)

p13  <- ASV_13_methylococcales + guides(color = guide_none(), shape = guide_none())
p32  <- ASV_32_methyloparacoccus + guides(color = guide_none(), shape = guide_none())
p141 <- ASV_141_methylobacterc + guides(color = guide_none(), shape = guide_none())

fig4 <- (p13 + p32 + p141) +
  plot_layout(ncol = 3, guides = "collect") &
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    #legend.box.just = "center",
    legend.box.margin = margin(2, 2, 2, 2),

    # make both guide boxes have identical sizing behavior
    legend.title.position = "top",
    legend.title.align = 0.5,
    legend.key.height = unit(0.35, "cm"),
    legend.key.width  = unit(0.35, "cm"),
    legend.spacing.y  = unit(0, "cm")
  ) &
  guides(
    color = guide_legend(title = "Treatment", nrow = 1, byrow = TRUE, title.position = "top", title.hjust = 0.5),
    shape = guide_legend(title = "Pond",      nrow = 2, byrow = TRUE, title.position = "top", title.hjust = 0.5)
  )

fig4


library(cowplot)
library(patchwork)

# 1) Make one plot that DEFINITELY contains BOTH legends (color + shape),
#    and format it how you want the legend to look.
leg_plot <- ASV_141_methylobacterc +
  theme(legend.position = "right",
        legend.spacing = unit(.4, "cm"),
        #legend.key.size = unit(0.4, "cm"),
        legend.box.just = "center",
        legend.spacing.x = unit(1, "cm"));leg_plot 
       #legend.margin = margin(t = -5, unit = "pt"))
#+
  # guides(
  #   color = guide_legend(title = "Treatment", nrow = 2, byrow = TRUE, title.position = "top", title.hjust = 0.5),
  #   shape = guide_legend(title = "Pond",      nrow = 2, byrow = TRUE, title.position = "top", title.hjust = 0.5)


leg <- cowplot::get_legend(leg_plot)

# 2) Remove legends from the plots that go into the 2x2 grid
p13  <- ASV_13_methylococcales     + theme(legend.position = "none")
p32  <- ASV_32_methyloparacoccus   + theme(legend.position = "none")
p141 <- ASV_141_methylobacterc     + theme(legend.position = "none")

# 3) Put the legend in the empty cell (bottom-right)
fig4 <- (p13 + p32) +
        (p141 + cowplot::ggdraw(leg)) +
  plot_layout(widths = c(1,0.5));fig4


library(patchwork)

p13  <- ASV_13_methylococcales     + theme(legend.position = "bottom")
p32  <- ASV_32_methyloparacoccus   + theme(legend.position = "bottom")
p141 <- ASV_141_methylobacterc     + theme(legend.position = "bottom")

fig4 <- (p13 + p32) /
        (p141 + guide_area()) +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "right",        # collected legend will be drawn in guide_area
    legend.box.just = "center",
    plot.margin = margin(0, 0, 0, 0)
  )

fig4

```



### Sediment
First we are running this analysis with all methane cyclers.
```{r diff-abund-sediment}
# filter out for ASVs with zero variances; recommended to remove them as they are sparse and lowly abundant 

# remove bad taxa 
bad_taxa <- c("ASV_3924", "ASV_4747", "ASV_6182", "ASV_2152", "ASV_5977", "ASV_11602", "ASV_568", "ASV_10610", "ASV_8789", "ASV_15179", "ASV_5923", "ASV_8468", "ASV_3463", "ASV_2539", "ASV_2756", "ASV_4541", "ASV_7355", "ASV_2182", "ASV_5033", "ASV_2551", "ASV_4211", "ASV_3467", "ASV_1479", "ASV_1677", "ASV_7654", "ASV_13612", "ASV_6715", "ASV_9006", "ASV_8702", "ASV_9210")


scaled_sed_ch4_physeq_bc <-sed_ch4_cyclers_physeq %>%
  subset_taxa(., !(ASV %in% bad_taxa)) %>% 
  prune_taxa(taxa_sums(.) > 0,.)


# relevel solar_progress 
scaled_sed_ch4_physeq_bc@sam_data$solar_progress <- factor(scaled_sed_ch4_physeq_bc@sam_data$solar_progress, levels = c("No FPV", "FPV"))

# # run ancombc2 for all sediment methane cyclers
# sed_ch4_asv_output <- ancombc2(data = scaled_sed_ch4_physeq_bc,
#                                  tax_level = "ASV", # Test for each phylum
#                                  fix_formula = "solar_progress", # Use Comp_Group_Hier to estimate diff. abundance
#                                  p_adj_method = "fdr", # Adjust with Holm-Bonferroni correction; recommended by authors
#                                  pseudo_sens = TRUE, # Run sensitivity test to make sure taxa isn't sensitive to psuedo-count choice
#                                  prv_cut = 0.05, # Prevalence filter of 10%
#                                  group = NULL, # Use Comp_Group_Hier as groups when doing pairwise comparisons
#                                  struc_zero = FALSE, # Do not detect structural zeroes
#                                  alpha = 0.05, # Significance threshold of 0.05
#                                  n_cl = 5, # Use 5 threads
#                                  s0_perc = 0.05,
#                                  verbose = FALSE, # Don't print verbose output
#                                  global = FALSE, # Run a global test (sorta like an ANOVA to first find if a given ASV is sig diff)
#                                  pairwise = FALSE) # Run pairwise tests between groups (sorta like a post-hoc test like Tukey)


# save(sed_ch4_asv_output, file = "data/03_diff_abund/sed_ch4_asv_output.RData")

load("data/03_diff_abund/sed_ch4_asv_output.RData")


# plot ASV differential abundance
sed_ch4_fsp <- sed_ch4_asv_output$res %>%
  select(taxon, starts_with("lfc"), starts_with("diff"), starts_with("passed_ss")) %>%
  pivot_longer(cols = !taxon, names_to = "metric", values_to = "value") %>%
  separate_wider_delim(cols = metric, delim = "_", names = c("variable", "Comparison"), too_many = "merge") %>%
  mutate(Comparison = str_remove(Comparison, "\\(Intercept\\)")) %>% 
  mutate(Comparison = str_remove(Comparison, "ss_")) %>%
  pivot_wider(id_cols = c("taxon","Comparison"), names_from = variable, values_from = value) %>%
  mutate(Comparison = str_remove(Comparison, "solar_progress"),
         Comparison = str_replace(Comparison, "_solar_progress", ";")) %>%
  separate_wider_delim(Comparison, delim = ";", names = c("Ref1", "Ref2"), too_few = "align_start") %>%
  filter(!is.na(Ref1) & Ref1 != "") %>%
  mutate(
    Ref2 = ifelse(is.na(Ref2), "Open", Ref2), # relevel with basegroup which is no solar 
    Comparison = paste0(Ref2, " : ", Ref1)) %>% 
  dplyr::filter(diff == 1, passed == 1, abs(lfc) > 1) %>% # play around with log fold change
  select(ASV = taxon, Comparison, lfc, passed)

# join by tax table
clean_sed_ch4 <-  sed_ch4_fsp %>% left_join(., as.data.frame(sed_ch4_cyclers_physeq@tax_table), by = "ASV")

# plot log fold changes
clean_sed_ch4 %>% 
  ggplot(aes(x = ASV, y = lfc, fill = Family)) +
  geom_col() +
  #scale_fill_manual(values = phylum_colors) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom") + 
  ggtitle("Sediment CH4 Cycler ASV Log-fold Change in FPV Ponds") 

# plot differentially abundant ASVs overtime 


# methanogen = ASV_4603; Methanosarcinales_A_2632 order

# get metadata from water physeq 
# metadata <- scaled_sed_ch4_physeq %>%
#   sample_data() %>%
#   data.frame()

sed_ch4_asv_df_glom <- sed_ch4_cyclers_physeq %>% # this phyloseq object has been transformed to relative abundance already
  tax_glom(taxrank = "ASV") %>% 
  psmelt() %>% 
  mutate(
    solar_progress = recode(solar_progress, "Solar" = "FPV", "No FPV" = "Open"))


# plot Methanogen overtime
sed_ch4_asv4603 <- sed_ch4_asv_df_glom %>% 
  dplyr::filter(ASV == "ASV_4603") %>%
  dplyr::group_by(
    JDate, Pond, solar_progress) %>%
  dplyr::summarise(
    rel_abundance = sum(Abundance, na.rm = TRUE),
    .groups = "drop") %>%  
  ggplot(aes(x = JDate, y = rel_abundance, color = solar_progress)) +
  geom_line(aes(group = interaction(Pond)), 
            alpha = 0.2) +
  geom_smooth(aes(group = solar_progress), se = FALSE) +
  geom_point(aes(shape = Pond), size = 2) +
  scale_color_manual(values = solar_colors) +
  scale_shape_manual(values = pond_shapes) +
  labs(
    x = "Date Collected",
    y = "Relative Abundance (%)",
    title = "Dif Abund Sed Methanosarcinales_A_2632 (ASV_4603) in FPV Ponds"
  ) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
        legend.position = "bottom") +
  scale_x_discrete(guide = guide_axis(angle = 60)) +
  theme_bw()
sed_ch4_asv4603

```
There is a differentially abundant ASV Methanosarcinales_A_2632 that is increased in solar ponds! When we plot the abundance overtime this ASV is barely in the community but it is higher in solar ponds. Doesnt really feel worth it to report considering its a minor contribution to community.

```{r plot-sed-ASV-diffAbund}
sed_difAbund_plot <- sed_ch4_asv_df_glom %>%
  dplyr::filter(ASV == "ASV_4603") %>%
  dplyr::group_by(
    JDate, Pond, Depth_Class, solar_progress,
    CH4_Cycler, Phylum, Class, Order, Family, Genus, ASV) %>%
  dplyr::summarise(
    rel_abundance = sum(Abundance, na.rm = TRUE),
    .groups = "drop") %>%  
  ggplot(aes(x = solar_progress, y = rel_abundance, color = solar_progress)) +
  geom_point(aes(shape = Pond),
             size = 2, alpha = 0.8, stroke = 0.8,
             position = position_jitterdodge(jitter.width = .5, dodge.width = .3)) +
  geom_boxplot(outlier.shape = NA, alpha = 0, color = "black", 
               position = position_dodge(0.6)) + 
  labs(color = "Treatment",
       y = "Relative Abundance (%)",
       title = "Methanoperedens_A \nASV_4603") +
  scale_color_manual(values = solar_colors) +
  scale_shape_manual(values = pond_shapes) +
  theme_classic() +
  ### ADD PVALUES 
  stat_compare_means(method = "wilcox.test", 
                     #comparisons = list(c("FPV", "Open")),
                     label = "p.format", # or "p.format" or "p.value"
                     #group.by = "combined_label",
                     size = 3,               # ⬅️ Font size
                     label.y.npc = 0.7,
                     #label.y = c(8000, 100000, 500000, 400000, 400000),
                     label.x = 1.75) +    # ⬅️ Manually set y position)  +
  guides(
    color = guide_legend(ncol = 2,override.aes = list(size = 2)),
    shape = guide_legend(ncol = 2, override.aes = list(size = 3))) +
  theme(legend.position = "none", #c(0.75, 0.7),
        axis.title.x = element_blank(),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 7),
        legend.key.size = unit(0.4, "cm"),
        legend.spacing.x = unit(0.2, "cm"),
        legend.margin = margin(t = -5, unit = "pt"),
        strip.text = element_text(size = 10),
        plot.title = element_text(size = 10)); sed_difAbund_plot


# save the plot   
# ggsave(sed_difAbund_plot, width = 3, height = 2, dpi = 300,
#         filename = "figures/Fig_S2/Fig_S2.png")

```

# Supplemental Figures
# Figure S3
Now we will split this up by water and sediment methane cyclers.

## Water Methanogens
Cannot run methanogens on water column, data is too sparse
```{r s3a-water-pcoa-methanogen, eval = FALSE}
# first lets filter our sed_ch4_cycler_physeq for methanogen or methanotroph
water_methanogens_physeq <- subset_taxa(water_ch4_cyclers_physeq, CH4_Cycler == "Methanogen") %>% 
  prune_taxa(taxa_sums(.) > 0, .) # 66 ASVs

# it seems that there are so few methanogen asvs that i cant even calculate a distance matrix without removing those 0s. lets see who those are
samps <- sample_sums(water_methanogens_physeq)
which(samps == 0) 

# further indication of how strongly methanotrophs shape water column 
```

## Water Methanotrophs
```{r s1-sed-pcoa-methanotrophs}
# first lets filter our physeq for methanotroph
water_methanotrophs_physeq <- subset_taxa(water_ch4_cyclers_physeq, CH4_Cycler == "Methanotroph") %>% 
  prune_taxa(taxa_sums(.) > 0, .) # 189 ASVs

# 2. methanotrophs

# Calculate Bray-Curtis Dissimilarity 
water_methanotroph_BC_pcoa <- 
  ordinate(
    physeq = water_methanotrophs_physeq,
    method = "PCoA",
    distance = "bray", 
    binary = FALSE
  )


## NEW PLOT 
#### Grab the data for the plot 
water_methanotroph_df <- 
  plot_ordination(
  physeq = water_methanotrophs_physeq,
  ordination = water_methanotroph_BC_pcoa,
  color = "solar_progress",
  shape = "Pond",
  justDF = TRUE)

# now lets mutate the columns
water_methanotroph_df <- water_methanotroph_df %>% 
dplyr::mutate(
    solar_progress = recode(solar_progress, "FPV" = "FPV", "No FPV" = "Open")) # solar progress


### Now, plot Figure S3B: WATER METHANOTROPHS 
figS3B_water_methanotroph_pcoa <- 
  ggplot(data = water_methanotroph_df, 
       aes(x = Axis.1, 
           y = Axis.2,
           color = solar_progress,
           shape = Pond)) + 
  geom_point(size = 3, alpha = 0.8, stroke = 0.8) +
  scale_shape_manual(values = pond_shapes) + 
  scale_color_manual(values = solar_colors) +
  labs(color = "Treatment",
       shape = "Pond",
       x = "Axis.1 [28.7%]",
       y = "Axis.2 [11.6%]",
       title = "Water Methanotrophs") + 
  guides(
    color = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5)),
    shape = guide_legend(
      nrow = 2,
      byrow = TRUE,
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5))
  ) +
  theme_classic() +
  theme(legend.position = "bottom",
        #legend.spacing = unit(0, "cm"),
        legend.box.background = element_blank(),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        plot.title = element_text(hjust = 0.5, size = 9, colour = "black"),
        axis.text.x = element_text(size = 8, colour = "black"),
        axis.text.y = element_text(size = 8, colour = "black"),
        axis.title.x = element_text(size = 8, colour = "black"),
        axis.title.y = element_text(size = 8, colour = "black"))


# Show the plot
figS3B_water_methanotroph_pcoa


# PCoA of sediments color by treatment shaped by pond
s3b_water_troph <- plot_ordination(
  physeq = water_methanotrophs_physeq,
  ordination = water_methanotroph_BC_pcoa,
  color = "solar_progress",
  shape = "Pond",
  title = "Sediment Methanotrophs") +
  geom_point(size = 5, alpha = 0.5, aes(color = solar_progress, fill = solar_progress, shape = Pond)) +
  scale_color_manual(values = solar_colors) +
  scale_fill_manual(values = solar_colors) +
  scale_shape_manual(values = pond_shapes) +
  guides(color = "none",
         fill = "none",
         shape = "none")+
  theme_classic()
s3b_water_troph

```

## Procrustes and Mantels Tests - Water Column CH4 Cyclers
When we plot our water column only methanotrophs and compare it to our main figure water column methanogens + methanotrophs, the plots are nearly identical. This indicates that methanotrophs are responsible for structuring the water column as indicated by abundances in figure 2. 

But to confirm if water column methanotrophs truly structure the water column we will perform Procrustes analysis and Mantel's test
```{r procrustes-mantel}
###### Procrustes ######
# run procrustes on our water methanotroph pcoas

# we will fit our methanotroph only phyloseq onto our methanogen+methanotroph water phyloseq
mt_only_sites <- plot_ordination(
  physeq      = water_methanotrophs_physeq,
  ordination  = water_methanotroph_BC_pcoa,
  justDF      = TRUE) %>%
  dplyr::select(Axis.1, Axis.2) %>%
  tibble::rownames_to_column("SampleID")

all_sites <- plot_ordination(
  physeq      = water_ch4_cyclers_physeq,
  ordination  = water_BC_pcoa,
  justDF      = TRUE) %>%
  dplyr::select(Axis.1, Axis.2) %>%
  tibble::rownames_to_column("SampleID")

# ensure that all SampleID names match
shared_ids <- intersect(mt_only_sites$SampleID, all_sites$SampleID)

# arrange SampleID names in exact order for comparison
mt_only_sites2 <- mt_only_sites %>%
  filter(SampleID %in% shared_ids) %>%
  arrange(SampleID)

all_sites2 <- all_sites %>%
  filter(SampleID %in% shared_ids) %>%
  arrange(SampleID)

# sanity check to make sure everything looks right!
identical(mt_only_sites2$SampleID, all_sites2$SampleID)


# run procrustes test
pro <- procrustes(
  X = as.matrix(all_sites2[, c("Axis.1", "Axis.2")]),  # target for fitting 
  Y = as.matrix(mt_only_sites2[, c("Axis.1", "Axis.2")]),
  symmetric = FALSE)
pro

# now plot procrustes test
plot(pro, kind = 1, main = "Procrustes: Methanotrophs fitted to All CH4 Cyclers")

plot(pro, kind = 2)   # residuals; very low meaning good fit

# Protest which is a permutational test of the significance of the procrustes result
  # based on the correlation from a symmetric Procrustes analysis
  # the order of X and Y makes no difference to the test of significance
proc_test <- protest(
  X = as.matrix(all_sites2[, c("Axis.1", "Axis.2")]),
  Y = as.matrix(mt_only_sites2[, c("Axis.1", "Axis.2")]),
  permutations = 9999)
proc_test

# procrustest sum of squares (m12 squared): 0.001626 --> 0 is perfect overlap, points are nearly identical 
  # This statistic measures the total mismatch between the two ordinations after optimal rotation and scaling

# Correlation in a symmetric Procrustes rotation: 0.9992 --> insanely high correlation meaning water column is predominately driven by methanotrophs NOT methanogens which makes sense



###### Mantel's Test ######
# Calculate Bray–Curtis distances
dist_all <- phyloseq::distance(water_ch4_cyclers_physeq, method = "bray")

dist_mt_only <- phyloseq::distance(water_methanotrophs_physeq, method = "bray")

# Convert to matrices
dist_all_mat <- as.matrix(dist_all)
dist_mt_only_mat  <- as.matrix(dist_mt_only)

# Find shared samples
shared_ids <- intersect(rownames(dist_all_mat),
                        rownames(dist_mt_only_mat))

# Ensure all samples match and are in the same order
dist_all_mat <- dist_all_mat[shared_ids, shared_ids]
dist_mt_only_mat  <- dist_mt_only_mat[shared_ids, shared_ids]

# Intuition check to make sure everything matches
identical(rownames(dist_all_mat), rownames(dist_mt_only_mat))

# Run Mantel test
mantel_res <- mantel(xdis = dist_all_mat,
                     ydis = dist_mt_only_mat,
                     method = "spearman", # try out spearman first which is recommended for ecological distances
                     permutations = 9999)

mantel_res # mantel statistic r = 0.9959 indicating perfect correspondence and is statistically significant 

# mantel statistic r (0 = no relationship; 1 = perfect correspondence)
```

## Sediment Methanogens
We want to see who is structuring the community within the sediments. In water column it is clearer that methanotrophs are, but what about in sediment communities?
```{r s3-sed-pcoa-methanogen}
# first lets filter our sed_ch4_cycler_physeq for methanogen or methanotroph
sed_methanogens_physeq <- subset_taxa(sed_ch4_cyclers_physeq, CH4_Cycler == "Methanogen") %>% 
  prune_taxa(taxa_sums(.) > 0, .) # 150 ASVs


# 1. methanogens

# Calculate Bray-Curtis Dissimilarity 
sed_methanogens_BC_pcoa <- 
  ordinate(
    physeq = sed_methanogens_physeq,
    method = "PCoA",
    distance = "bray", 
    binary = FALSE
  )

## NEW PLOT 
#### Grab the data for the plot 
sed_methanogen_ord_df <- 
  plot_ordination(
  physeq = sed_methanogens_physeq,
  ordination = sed_methanogens_BC_pcoa,
  color = "solar_progress",
  shape = "Pond",
  justDF = TRUE)

# now lets mutate the columns
sed_methanogen_ord_df <- sed_methanogen_ord_df %>% 
dplyr::mutate(
    solar_progress = recode(solar_progress, "FPV" = "FPV", "No FPV" = "Open")) # solar progress


### Now, plot Figure S3: SEDIMENT METHANOGENS 
figS3_sed_methanogens_pcoa <- 
  ggplot(data = sed_methanogen_ord_df, 
       aes(x = Axis.1, 
           y = Axis.2,
           color = solar_progress,
           shape = Pond)) + 
  geom_point(size = 3, alpha = 0.8, stroke = 0.8) +
  scale_shape_manual(values = pond_shapes) + 
  scale_color_manual(values = solar_colors) +
  labs(color = "Treatment",
       shape = "Pond",
       x = "Axis.1 [33.3%]",
       y = "Axis.2 [17.7%]",
       title = "Sediment Methanogens") + 
  guides(
    color = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5)),
    shape = guide_legend(
      nrow = 2,
      byrow = TRUE,
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5))
  ) +
  theme_classic() +
  theme(legend.position = "bottom",
        #legend.spacing = unit(0, "cm"),
        legend.box.background = element_blank(),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        plot.title = element_text(hjust = 0.5, size = 9, colour = "black"),
        axis.text.x = element_text(size = 8, colour = "black"),
        axis.text.y = element_text(size = 8, colour = "black"),
        axis.title.x = element_text(size = 8, colour = "black"),
        axis.title.y = element_text(size = 8, colour = "black"))

# Show the plot
figS3_sed_methanogens_pcoa

figS1A_sed_methanogens_pcoaold <-
  plot_ordination(
  physeq = sed_methanogens_physeq,
  ordination = sed_methanogens_BC_pcoa,
  color = "solar_progress",
  shape = "Pond",
  title = "Sediment Methane Cyclers") +
  geom_point(size = 5, alpha = 0.5,
             aes(color = solar_progress, fill = solar_progress, shape = Pond)) +
  scale_color_manual(values = solar_colors) +
  scale_fill_manual(values = solar_colors) +
  scale_shape_manual(values = pond_shapes) +
  guides(color = "none",
         fill = "none",
         shape = "none") +
  theme_classic()
  # theme(
  # legend.position = c(0.82, 0.01),  # inside bottom-left
  # legend.justification = c(0, 0),  # anchor the legend's top-left corner there
  # legend.spacing = unit(0.1, "cm"),
  # legend.background = element_rect(color = NA, fill = NA),
  # legend.box.background = element_rect(size = 0.1, linetype = "solid", color = "black"),
  # legend.text = element_text(size = 6),
  # legend.margin = margin(2, 2, 2, 2))
figS1A_sed_methanogens_pcoaold
```


## Sediment Methanotrophs

```{r s3-sed-pcoa-methanotrophs}
# first lets filter our sed_ch4_cycler_physeq for methanogen or methanotroph
sed_methanotrophs_physeq <- subset_taxa(sed_ch4_cyclers_physeq, CH4_Cycler == "Methanotroph") %>% 
  prune_taxa(taxa_sums(.) > 0, .) # 197 ASVs

# 2. methanotrophs

# Calculate Bray-Curtis Dissimilarity 
sed_methanotroph_BC_pcoa <- 
  ordinate(
    physeq = sed_methanotrophs_physeq,
    method = "PCoA",
    distance = "bray", 
    binary = FALSE
  )


## NEW PLOT 
#### Grab the data for the plot 
sed_methanotroph_ord_df <- 
  plot_ordination(
  physeq = sed_methanotrophs_physeq,
  ordination = sed_methanotroph_BC_pcoa,
  color = "solar_progress",
  shape = "Pond",
  justDF = TRUE)

# now lets mutate the columns
sed_methanotroph_ord_df <- sed_methanotroph_ord_df %>% 
dplyr::mutate(
    solar_progress = recode(solar_progress, "FPV" = "FPV", "No FPV" = "Open")) # solar progress


### Now, plot Figure S1B: SEDIMENT METHANOTROPHS 
figS3_sed_methanotroph_pcoa <- 
  ggplot(data = sed_methanotroph_ord_df, 
       aes(x = Axis.1, 
           y = Axis.2,
           color = solar_progress,
           shape = Pond)) + 
  geom_point(size = 3, alpha = 0.8, stroke = 0.8) +
  scale_shape_manual(values = pond_shapes) + 
  scale_color_manual(values = solar_colors) +
  labs(color = "Treatment",
       shape = "Pond",
       x = "Axis.1 [36.1%]",
       y = "Axis.2 [15.4%]",
       title = "Sediment Methanotrophs") + 
  guides(
    color = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5)),
    shape = guide_legend(
      nrow = 2,
      byrow = TRUE,
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(size = 2.5))
  ) +
  theme_classic() +
  theme(legend.position = "bottom",
        #legend.spacing = unit(0, "cm"),
        legend.box.background = element_blank(),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        plot.title = element_text(hjust = 0.5, size = 9, colour = "black"),
        axis.text.x = element_text(size = 8, colour = "black"),
        axis.text.y = element_text(size = 8, colour = "black"),
        axis.title.x = element_text(size = 8, colour = "black"),
        axis.title.y = element_text(size = 8, colour = "black"))


# Show the plot
figS3_sed_methanotroph_pcoa


# PCoA of sediments color by treatment shaped by pond
s2b_sed_troph <- plot_ordination(
  physeq = sed_methanotrophs_physeq,
  ordination = sed_methanotroph_BC_pcoa,
  color = "solar_progress",
  shape = "Pond",
  title = "Sediment Methanotrophs") +
  geom_point(size = 5, alpha = 0.5, aes(color = solar_progress, fill = solar_progress, shape = Pond)) +
  scale_color_manual(values = solar_colors) +
  scale_fill_manual(values = solar_colors) +
  scale_shape_manual(values = pond_shapes) +
  guides(color = "none",
         fill = "none",
         shape = "none")+
  theme_classic()
s2b_sed_troph
```


## Save Fig S3

```{r plot-FigS3}
# extract legend 
leg <- cowplot::get_legend(
  figS3_sed_methanogens_pcoa +
    scale_color_manual(values = solar_colors, breaks = c("FPV", "Open")) +
    guides(
      color = guide_legend(title.hjust = 0.5, nrow = 1, byrow = TRUE,override.aes = list(size = 2.5)),  # horizontal Treatment
      shape = guide_legend(title.hjust = 0.5, nrow = 2, byrow = TRUE, override.aes = list(size = 2.5))   # optional
    ) +
    theme(#legend.key.size = unit(0.4, "cm"), 
          #legend.spacing.x = unit(0.7, "cm"),
          #legend.margin = margin(t = -5, unit = "pt"),
          legend.position = "right", 
          legend.box = "vertical",
          legend.box.just = "center"))

leg_plot <- wrap_elements(leg) 


# edit individual figures
pA <- figS3B_water_methanotroph_pcoa +
  labs(tag = "A.") +
  theme(legend.position = "none",
        plot.title = element_text(margin = margin(b = 0)))

pC <- figS3_sed_methanotroph_pcoa +
  labs(tag = "B.") +
  theme(legend.position = "none",
        plot.title = element_text(margin = margin(b = 0)))

pD <- figS3_sed_methanogens_pcoa +
  labs(tag = "C.") +
  theme(legend.position = "none",
        plot.title = element_text(margin = margin(b = 0)))

# plot together
plot_figS3 <-
  (pA + leg_plot) / (pC + pD) +
  plot_layout(widths = c(1, 1)) &
  theme(
    plot.tag = element_text(size = 8, colour = "black"))
plot_figS3



# Now, actually save the plot   
ggsave(plot_figS3, width = 6.5, height = 4.5, dpi = 300,
        filename = "figures/Fig_S3/Fig_S3.png")
```

Sediment samples are still distinct from other and separate along first axis

## Fig S3: PERMANOVA 

PERMANOVA (Permutational Multivariate Analysis of Variance) is a non-parametric, permutation-based test used to compare groups of objects based on a distance matrix. The goal is to test the null hypothesis that the centroids and dispersion of groups are equivalent in the space defined by the dissimilarity measure. 

### Water Methanotrophs

### Sediment Methanogens

Here we are performing a PERMANOVA on the sediment methanogen and methanotrophs
```{r permanova-sed-methanogens}
#1. methanogen
# calculate Bray-Curtis PERMANOVA using phyloseq distance
sed_gen_bray <- phyloseq::distance(sed_methanogens_physeq, method = "bray", binary = FALSE)


# pull out metadata 
sed_methanogens_metadata <- sed_methanogens_physeq %>%
  sample_data() %>%
  data.frame()

# Permutational Multivariate Analysis of Variance Using Distance Matrices
# aka PERMANOVA using the adonis2 function from vegan 


#1. Test the individual terms for significance
# Testing if the centroids of solar progress are different: significant p = 0.001 ***
adonis2(sed_gen_bray ~ solar_progress, 
        data = sed_methanogens_metadata, by = "terms")


# Does pond matter? significant p = 0.001 ***
adonis2(sed_gen_bray ~ Pond, 
        data = sed_methanogens_metadata, by = "terms")

# Does date matter? significant p = 0.001 ***
adonis2(sed_gen_bray ~ as.factor(JDate), 
        data = sed_methanogens_metadata, by = "terms")

#2. Test the terms together
# Now lets see the effect of each pond by date_collected and solar progress
sed_methanogens_permanova <- 
  adonis2(sed_gen_bray ~ solar_progress * Pond * JDate, 
          data = sed_methanogens_metadata, by = "terms");

# Show the results! 
sed_methanogens_permanova
```

### Sediment Methanotrophs

```{r permanova-sed-methanotrophs}
#1. methanotrophs
# calculate Bray-Curtis PERMANOVA using phyloseq distance
sed_troph_bray <- phyloseq::distance(sed_methanotrophs_physeq, 
                     method = "bray", binary = FALSE)

# pull out metadata 
sed_methanotrophs_metadata <- sed_methanotrophs_physeq %>%
  sample_data() %>%
  data.frame()

# Permutational Multivariate Analysis of Variance Using Distance Matrices
# aka PERMANOVA using the adonis2 function from vegan 


#1. Test the individual terms for significance
# Testing if the centroids of solar progress are different: significant p = 0.002 **
adonis2(sed_troph_bray ~ solar_progress, 
        data = sed_methanotrophs_metadata, by = "terms")

# Does pond matter? significant p = 0.001 ***
adonis2(sed_troph_bray ~ Pond, 
        data = sed_methanotrophs_metadata, by = "terms")

# Does date matter? significant p = 0.001 ***
adonis2(sed_troph_bray ~ as.factor(JDate), 
        data = sed_methanotrophs_metadata, by = "terms")

#2. Test the terms together
# Now lets see the effect of each pond by date_collected and solar progress
sed_methanotrophs_permanova <- 
  adonis2(sed_troph_bray ~ solar_progress * Pond * JDate, 
        data = sed_methanotrophs_metadata, by = "terms")

# Show the results! 
sed_methanotrophs_permanova
```
**Water Methanotrophs**

**Sed Methanogens**
With our PERMANOVA we find that treatment (solar_progress), day of year sampled (JDate - Julian date), and Pond is significant 

Treatment explains 11.8% of the variance and has the largest effect size (F = 10.5) but pond explains the most variation, 28.2%, and contributes to the community but weaker than treatment (F = 6.22). JDate explains 9.0% of variation but is the second most important term for its weight contributing to structuring thet community. 

Solar progress and time explains 2.5% of the variation but is not a strong contributer to the community. Pond and time explais 10% of variation but has a smaller effect on community structure.

Together this explains 62% of the variation.

**Sed Methanotrophs**

With our PERMANOVA we see that Pond explains the most variance (20.4%) but does not have a strong effect on community structure (F = 3.9). There is a temporal effect along the first axis due to time that explains 13.1% of data and is a strong factor for shaping the community (F = 10.0). Treatment is important for explaining 10.1% of variance and is teh second most important factor for shaping the community which we kinda see along the second axis (F = 7.7)

The interactions of pond and date explain 9.7% of the variance but is not an important factor for shaping our community (F = 1.9). 

The interaction of solar treatment and pond is > 0.05 (p = 0.095) indicating that their interaction is not strong and important for shaping the community. while it does answer 2.2% of variation, it has the smallest effect size (F = 1.7) 

Together these variables explain 56% of the data

## Fig S3: Betadisper

### Water Methanotrophs

### Sediment Methanogens 
```{r betadisper-sed-methanogens}
# 1. methanogens 
# Homogeneity of Disperson test with beta dispr

## Bray-Curtis
betadispr_sed_methanogens_pond <- 
  betadisper(sed_gen_bray, sed_methanogens_metadata$Pond)

betadispr_sed_methanogens_solar <- 
  betadisper(sed_gen_bray, sed_methanogens_metadata$solar_progress)

betadispr_sed_methanogens_JDate <- 
  betadisper(sed_gen_bray, sed_methanogens_metadata$JDate)

# permutest() performs a non-parametric permutation test, which is robust and valid for the kind of data used in beta diversity analysis (e.g., dissimilarity matrices).
permutest(betadispr_sed_methanogens_pond) # not significant p = 0.659

permutest(betadispr_sed_methanogens_solar) # not significant p = 0.067

permutest(betadispr_sed_methanogens_JDate) # not significant p = 0.44
```

### Sediment Methanotrophs
```{r betadisper-sed-methanotrophs}
# 2. methanotrophs 
# Homogeneity of Disperson test with beta dispr

## Bray-Curtis
betadispr_sed_methanotrophs_pond <- 
  betadisper(sed_troph_bray, sed_methanotrophs_metadata$Pond)

betadispr_sed_methanotrophs_solar <- 
  betadisper(sed_troph_bray, sed_methanotrophs_metadata$solar_progress)

betadispr_sed_methanotrophs_JDate <-
  betadisper(sed_troph_bray, sed_methanotrophs_metadata$JDate)


# permutest() performs a non-parametric permutation test, which is robust and valid for the kind of data used in beta diversity analysis (e.g., dissimilarity matrices).
permutest(betadispr_sed_methanotrophs_pond) # not significant p = 0.515

permutest(betadispr_sed_methanotrophs_solar) # not significant p = 0.682

permutest(betadispr_sed_methanotrophs_JDate) # significant p = 0.011 *
```
**Sediment Methanogens**
With betadispr we find the PERMANOVA results are are valid as pond, treatment, and date are not significant but significant in the PERMANOVA. Thus our PERMANOVA result is reliable and the differences between groups are due to location/centroids of groups rather than differences in variation within groups 

**Sediment Methanotrophs**
With betadispr we find the PERMANOVA results are are valid as pond and treatment are not significant but significant in the PERMANOVA. Thus our PERMANOVA result is reliable and the differences between groups are due to location/centroids of groups rather than differences in variation within groups 

However, date is statistically significant in PERMANOVA and in the betadispr indicating that theres variability in within the sampling dates so there are likely differences in community composition and probably heterogeneity over time. 


# Table S6 - Taxonomy Table
reformat later
```{r table-s6-taxtable}
# water
taxonomy_table <- tax_table(water_ch4_cyclers_physeq) %>% 
  as.data.frame() %>%
  tibble::rownames_to_column("OTU")

order_genus_summary <- taxonomy_table %>%
  group_by(Kingdom, Phylum, Class, Order, Family) %>%
  summarize(
    Genera = paste(sort(unique(Genus[!is.na(Genus)])), collapse = ", ")
  ) %>%
  arrange(Phylum, Class, Order, Family)
order_genus_summary

# sediment

taxonomy_table_s <- tax_table(sed_ch4_cyclers_physeq) %>% 
  as.data.frame() %>%
  tibble::rownames_to_column("OTU")

order_genus_summary_s <- taxonomy_table_s %>%
  group_by(Kingdom, Phylum, Class, Order, Family) %>%
  summarize(
    Genera = paste(sort(unique(Genus[!is.na(Genus)])), collapse = ", ")
  ) %>%
  arrange(Phylum, Class, Order, Family)
order_genus_summary_s

```




# Bonus code
# Plot DO vs Methanotroph Abundance Dynamics
1. Plot DO dynamics over 2024 sampling season.
2. Plot methanotroph abundances overtime.
  - first we will plot all methanotrophs (gammaproteobacterial and alphaproteobacterial) abundances overtime
  - filter for gammaproteobacterial and alphaproteobacterial methanotroph abundances
      - gammaproteobacterial (historically named type I methanotrophs) dominate the system but curious about alphaproteobacterial (historically type II) abundance and NC10 that are anaerobic methane oxidizers
        - note the designation type I and type II are just historical names at this point!

3. Then we will plot all water column methanotrophs against DO concentrations. Could separate this further by gamma and alphaproteobacterial methanotrophs but wont be doing that now.
```{r do-abundance-ot}
# load metadata with DO concentrations, could also plot as DO % saturation
load("data/00_load_data/chem_metadata_23_24.RData")

# load in water_ch4_cycler_df data frame before we manipulated it when we calculated normality
load("data/01_phyloseq/water_ch4_cyclers_df.RData")

# set levels
water_ch4_cyclers_df$solar_progress <- factor(
  water_ch4_cyclers_df$solar_progress,
  levels = c("FPV", "Open"))

# factor depth
water_ch4_cyclers_df$Depth_Class <- factor(
  water_ch4_cyclers_df$Depth_Class,
  levels = c("Surface Water", "Bottom Water"))

# rename df to preserver abundance data whcih we will calculate below 
water_ch4_cyclers_df_dynamics <- water_ch4_cyclers_df

# now add interaction column to our df for temporal dynamics old code im not ready to delete
# water_ch4_cyclers_df_dynamics <- water_ch4_cyclers_df %>%
#   mutate(group = interaction(CH4_Cycler, Depth_Class, sep = " ")) %>% 
#   group_by(Pond, solar_progress, Depth_Class, group, JDate, CH4_Cycler, DNA_ID) %>%
#   summarise(
#     total_abundance = sum(Abundance, na.rm = TRUE), # total abundance across all samples
#     .groups = "drop"
#   )

# join with chem metadata with abundance data for temporal dynamics
do_abundance <- left_join(water_ch4_cyclers_df_dynamics, chem_metadata_23_24, by = "DNA_ID")

# 1. plot DO mg/L overtime; can alternatively plot with DO % saturation
do_mgl <- do_abundance %>% 
  group_by(JDate.x, Pond.x, solar_progress.x, Depth_Class.x) %>% 
  summarise(avg_do = mean(HDO_mg.l, na.rm = TRUE), .groups = "drop") %>% 
  ggplot(aes(x = JDate.x,
             y = avg_do,
             color = solar_progress.x,
             fill = solar_progress.x,
             group = Pond.x,
             shape = Pond.x)) +
  geom_line(alpha = 0.1) +
  geom_smooth(aes(group = solar_progress.x), se = FALSE, linewidth = 1.4) +
  geom_point(alpha = 0.7) +
  scale_shape_manual(values = pond_shapes)+
  ggh4x::facet_wrap2(~Depth_Class.x)
do_mgl

do_mgl_simple <- do_abundance %>% 
  group_by(JDate.x, Pond.x, solar_progress.x, Depth_Class.x) %>% 
  summarise(avg_do = mean(HDO_mg.l, na.rm = TRUE), .groups = "drop") %>% 
  ggplot(aes(x = JDate.x,
             y = avg_do,
             color = solar_progress.x,
             fill = solar_progress.x,
             group = Pond.x)) +
 # geom_line(alpha = 0.1) +
  geom_smooth(aes(group = solar_progress.x), se = FALSE, linewidth = 1.4) +
  geom_point(alpha = 0.3) +
  #scale_shape_manual(values = pond_shapes)+
  ggh4x::facet_wrap2(~Depth_Class.x) +
  labs(
    y = "Dissolved Oxygen (mg/L)",
    x = "Day of Year") +
  guides(
    color = guide_legend(title = "Treatment"),
    fill = "none") +
  theme_classic2()
do_mgl_simple



# 2. plot all methanotroph abundances overtime 
methanotroph_abundance <- do_abundance %>% 
  dplyr::filter(CH4_Cycler == "Methanotroph") %>% 
  group_by(JDate.x, Pond.x, solar_progress.x, Depth_Class.x, HDO_mg.l) %>% 
  summarise(
    total_abundance = sum(Abundance, na.rm = TRUE), # total across all samples
    .groups = "drop") %>% 
 # summarise(avg_do = mean(HDO_mg.l, na.rm = TRUE), .groups = "drop") %>% 
  ggplot(aes(x = JDate.x,
             y = total_abundance,
             color = solar_progress.x,
             fill = solar_progress.x)) +
  geom_line() +
  geom_point() +
  labs(title = "All Methanotrophs") +
  #geom_smooth(aes(group = solar_progress.x), se = FALSE, linewidth = 1.4) +
  ggh4x::facet_wrap2(~Depth_Class.x)
methanotroph_abundance

# 2 simple; still plot all methanotrophs, just make it simpler and not shape by pond
methanotroph_all_abund_simple <- do_abundance %>% 
  dplyr::filter(CH4_Cycler == "Methanotroph") %>% 
  group_by(JDate.x, Pond.x, solar_progress.x, Depth_Class.x) %>% 
  summarise(
    total_abundance = sum(Abundance, na.rm = TRUE), # total across all samples
    .groups = "drop") %>% 
 # summarise(avg_do = mean(HDO_mg.l, na.rm = TRUE), .groups = "drop") %>% 
  ggplot(aes(x = JDate.x,
             y = total_abundance,
             color = solar_progress.x,
             fill = solar_progress.x,
             group = Pond.x)) +
  #geom_line() +
  geom_smooth(aes(group = solar_progress.x), se = FALSE, linewidth = 1.4) +
  geom_point(alpha = 0.3) +
  labs(#title = "All Methanotroph Abundances",
       y = "Absolute Abundance (cells per ml)",
       x = "Day of Year") +
  guides(
    color = guide_legend(title = "Treatment"),
    fill = "none") +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale(), accuracy = 1)) +
  ggh4x::facet_wrap2(~Depth_Class.x) +
  theme_classic2()
methanotroph_all_abund_simple

# 2a. plot only gammaproteobacterial methanotroph abundances overtime 
gamma_mob_abundance <- do_abundance %>% 
  dplyr::filter(CH4_Cycler == "Methanotroph" & Class == "Gammaproteobacteria") %>% 
  group_by(JDate.x, Pond.x, solar_progress.x, Depth_Class.x) %>% 
  summarise(
    total_abundance = sum(Abundance, na.rm = TRUE), # total across all samples
    .groups = "drop") %>% 
  ggplot(aes(x = JDate.x,
             y = total_abundance,
             color = solar_progress.x,
             fill = solar_progress.x,
             group = Pond.x,
             shape = Pond.x)) +
  geom_smooth(aes(group = solar_progress.x), se = FALSE, linewidth = 1.4) +
  geom_point(alpha = 0.3) +
  labs(title = "Gammaproteobacterial Methanotroph Abundances",
       y = "Absolute Abundance (cells per ml)",
       x = "Day of Year") +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale(), accuracy = 1)) +
  ggh4x::facet_wrap2(~Depth_Class.x) +
  scale_shape_manual(values = pond_shapes) +
  theme_classic2()
gamma_mob_abundance 

# 2a simple. plot only gammaproteobacterial methanotroph abundances overtime 
gamma_mob_abundance_simple <- do_abundance %>% 
  dplyr::filter(CH4_Cycler == "Methanotroph" & Class == "Gammaproteobacteria") %>% 
  group_by(JDate.x, Pond.x, solar_progress.x, Depth_Class.x) %>% 
  summarise(
    total_abundance = sum(Abundance, na.rm = TRUE), # total across all samples
    .groups = "drop") %>% 
  ggplot(aes(x = JDate.x,
             y = total_abundance,
             color = solar_progress.x,
             fill = solar_progress.x,
             group = Pond.x)) +
  geom_smooth(aes(group = solar_progress.x), se = FALSE, linewidth = 1.4) +
  geom_point(alpha = 0.3) +
  labs(title = "Gammaproteobacterial Methanotroph Abundances",
       y = "Absolute Abundance (cells per ml)",
       x = "Day of Year") +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale(), accuracy = 1)) +
  ggh4x::facet_wrap2(~Depth_Class.x) +
  #scale_shape_manual(values = pond_shapes) +
  theme_classic2()
gamma_mob_abundance_simple



# 2b. plot only alphaproteobacterial methanotroph abundances overtime 
alpha_mob_abundance <- do_abundance %>% 
  dplyr::filter(CH4_Cycler == "Methanotroph" & Class == "Alphaproteobacteria") %>% 
  group_by(JDate.x, Pond.x, solar_progress.x, Depth_Class.x) %>% 
  summarise(
    total_abundance = sum(Abundance, na.rm = TRUE), # total across all samples
    .groups = "drop") %>% 
  ggplot(aes(x = JDate.x,
             y = total_abundance,
             color = solar_progress.x,
             fill = solar_progress.x,
             shape = Pond.x,
             group = Pond.x)) +
  geom_smooth(aes(group = solar_progress.x), se = FALSE, linewidth = 1.4) +
  geom_point(alpha = 0.3) +
  labs(title = "Alphaproteobacterial Methanotroph Abundances",
       y = "Absolute Abundance (cells per ml)",
       x = "Day of Year") +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale(), accuracy = 1)) +
  scale_shape_manual(values = pond_shapes) +
  ggh4x::facet_wrap2(~Depth_Class.x) +
  theme_classic2()
alpha_mob_abundance #interesting that Open ponds have higher alphaproteobacterial methanotrophs; very little in FPV ponds; abundances are less than 80k for alphaproteo methanotrophs in Open ponds though but compared to FPV which have ~20k

# 2b simple. plot only alphaproteobacterial methanotroph abundances overtime; simply and not shaped by pond
alpha_mob_abundance_simple <- do_abundance %>% 
  dplyr::filter(CH4_Cycler == "Methanotroph" & Class == "Alphaproteobacteria") %>% 
  group_by(JDate.x, Pond.x, solar_progress.x, Depth_Class.x) %>% 
  summarise(
    total_abundance = sum(Abundance, na.rm = TRUE), # total across all samples
    .groups = "drop") %>% 
  ggplot(aes(x = JDate.x,
             y = total_abundance,
             color = solar_progress.x,
             fill = solar_progress.x,
             group = Pond.x)) +
  geom_smooth(aes(group = solar_progress.x), se = FALSE, linewidth = 1.4) +
  geom_point(alpha = 0.3) +
  labs(title = "Alphaproteobacterial Methanotroph Abundances",
       y = "Absolute Abundance (cells per ml)",
       x = "Day of Year") +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale(), accuracy = 1)) +
  ggh4x::facet_wrap2(~Depth_Class.x) +
  theme_classic2()
alpha_mob_abundance_simple


# 2c. plot only Methylomirabilia methanotroph abundances overtime (NC10 phylum); these are anaerobic bacterial methanotrophs that couples denitrification to methane oxidation! (AOM = anaerobic oxidation of methane)
methylomirabilia_aom_abundance <- do_abundance %>% 
  dplyr::filter(CH4_Cycler == "Methanotroph" & Class == "Methylomirabilia") %>% 
  group_by(JDate.x, Pond.x, solar_progress.x, Depth_Class.x) %>% 
  summarise(
    total_abundance = sum(Abundance, na.rm = TRUE), # total across all samples
    .groups = "drop") %>% 
  ggplot(aes(x = JDate.x,
             y = total_abundance,
             color = solar_progress.x,
             fill = solar_progress.x,
             shape = Pond.x,
             group = Pond.x)) +
  geom_smooth(aes(group = solar_progress.x), se = FALSE, linewidth = 1.4) +
  geom_point(alpha = 0.3) +
  labs(title = "Methylomirabilia (NC10 Phylum) Methanotroph Abundances",
       y = "Absolute Abundance (cells per ml)",
       x = "Day of Year") +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale(), accuracy = 1)) +
  scale_shape_manual(values = pond_shapes) +
  ggh4x::facet_wrap2(~Depth_Class.x) +
  theme_classic2()
methylomirabilia_aom_abundance #super lowly abundant in both FPV and Open ponds. virtually nonexistent in FPV ponds at both depths. virtually nonexistent in surface for Open ponds too; lowly abundant in Open bottom waters ~2k but completely crashes by last time point

# 2c simple. plot only NC10 phylum (Methylomirabilia class) methanotroph abundances overtime; simply and not shaped by pond
methylomirabilia_aom_abundance_simple <- do_abundance %>% 
  dplyr::filter(CH4_Cycler == "Methanotroph" & Class == "Methylomirabilia") %>% 
  group_by(JDate.x, Pond.x, solar_progress.x, Depth_Class.x) %>% 
  summarise(
    total_abundance = sum(Abundance, na.rm = TRUE), # total across all samples
    .groups = "drop") %>% 
  ggplot(aes(x = JDate.x,
             y = total_abundance,
             color = solar_progress.x,
             fill = solar_progress.x,
             group = Pond.x)) +
  geom_smooth(aes(group = solar_progress.x), se = FALSE, linewidth = 1.4) +
  geom_point(alpha = 0.3) +
  labs(title = "Methylomirabilia (NC10 Phylum) Methanotroph Abundances",
       y = "Absolute Abundance (cells per ml)",
       x = "Day of Year") +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale(), accuracy = 1)) +
  ggh4x::facet_wrap2(~Depth_Class.x) +
  theme_classic2()
methylomirabilia_aom_abundance_simple



# plot DO vs methanotroph abundance
## note: we are plotting all methanotrophs (gammaproteobacterial and alphaproteobacterial); should the specific gammmaproteobacterial methanotrophs become an interest then we must reprocess the dataframe to filter specifically for gammaproteos/ type I methanotrophs

abundvs_do_mgl <- do_abundance %>% 
  dplyr::filter(CH4_Cycler == "Methanotroph") %>% 
  group_by(JDate.x, Pond.x, solar_progress.x, Depth_Class.x, HDO_mg.l) %>% 
  summarise(
    total_abundance = sum(Abundance, na.rm = TRUE), # total across all samples
    .groups = "drop") %>% 
 # summarise(avg_do = mean(HDO_mg.l, na.rm = TRUE), .groups = "drop") %>% 
  ggplot(aes(x = HDO_mg.l,
             y = total_abundance,
             color = solar_progress.x,
             fill = solar_progress.x,
             group = Pond.x)) +
 # geom_line() +
  geom_point(alpha = 0.3) +
  geom_smooth(aes(group = solar_progress.x), se = FALSE, linewidth = 1.4) +
  ggh4x::facet_wrap2(~Depth_Class.x)
abundvs_do_mgl

# lets plot methanotroph abundance with dissolved oxygen concentration overtime with a double y axis - not the best plot!

# first calculate mean abundance and DO concentration between FPV and Open ponds overtime
abund_do_mean <- do_abundance %>% 
  dplyr::filter(CH4_Cycler == "Methanotroph") %>% 
  group_by(JDate.x, solar_progress.x, Depth_Class.x) %>% 
  summarise(
    mean_abundance = mean(Abundance, na.rm = TRUE),
    mean_do        = mean(HDO_mg.l, na.rm = TRUE),
    .groups = "drop")

# then we must perform a linear transformation by rescaling our DO concentration to methanotroph abundance
scale_factor <- max(abund_do_mean$mean_abundance, na.rm = TRUE) /
                max(abund_do_mean$mean_do, na.rm = TRUE)

# now lets plot!
ggplot(abund_do_mean, aes(x = JDate.x)) +
  geom_line(aes(
    y = mean_abundance, color = solar_progress.x, group = solar_progress.x),
    linewidth = 1.2) +
  geom_point(aes(
    y = mean_abundance, color = solar_progress.x),
    size = 2) +
  geom_line(aes(
      y = mean_do * scale_factor, color = solar_progress.x,
      group = solar_progress.x, linetype = "Dissolved oxygen"), 
      linewidth = 1, alpha = 0.7) +
  scale_y_continuous(
    name = expression("Methanotroph abundance (cells mL"^-1*")"),
    sec.axis = sec_axis(~ . / scale_factor, name = "Dissolved oxygen (mg L⁻¹)")) +
  facet_grid(~Depth_Class.x, scales = "free_y") +
  labs(x = "Day of year", color = "Treatment", linetype = NULL) +
  scale_linetype_manual(values = c("Dissolved oxygen" = "dotted")) +
  guides(
    color    = guide_legend(order = 1),
    linetype = guide_legend(order = 2)
  ) +
  theme_classic2()
# dont really love this plot but it is cool to see!


# put together for revision response
do_meth_abund <- do_mgl_simple + methanotroph_all_abund_simple +
  plot_layout(nrow = 2, ncol = 1,
              guides = "collect") 
do_meth_abund

ggsave(do_meth_abund, width = 6.5, height = 6.5, dpi = 300,
        filename = "figures/bonus/do_meth_abund.png")
```
From our rough plots we see that gammaproteobacterial methanotrophs dominate ponds, this makes sense from our figure 2. 

When we plot only alphaproteobacterial methanotrophs, control ponds seem to have a greater abundance in both depths although much lower abundances (<80k). July has the highest surfacw water time point in one Open
although max cell densities reach over 60,000 in surface waters while bottom waters have average of 40k cells 

Open ponds also have more Methylomirabilia (NC10 phylum) in bottom waters but pretty lowly abundant (~3k cells)

## Top Sediment Methanogen ASVs Over Time 

```{r methanogen-over-time}
# Create a dataframe 
sed_methanogens_df <- 
  sed_methanogens_physeq %>%
  speedyseq::psmelt() %>% # melt into dataframe
  dplyr::select(OTU, Sample, Abundance, DNA_ID, Date_Collected, Deployment_Depth_m, Pond, solar_progress, JDate, Kingdom:CH4_Cycler) %>%
  dplyr::select(-ASVseqs) %>%
  mutate(solar_progress = recode(solar_progress, "No FPV" = "Open"))

## Summary Stats of the ASVs for some perusing 
sed_methanogen_asv_stats <- 
  sed_methanogens_df %>%
  group_by(ASV) %>%
  summarize(mean = mean(Abundance), median = median(Abundance), min = min(Abundance), max = max(Abundance)) 

## Check by FPV and Open 
sed_methanogen_asv_stats_FPV <- 
  sed_methanogens_df %>%
  group_by(ASV, solar_progress) %>%
  summarize(mean = mean(Abundance), median = median(Abundance), min = min(Abundance), max = max(Abundance)) 

# Just create a vector with the ASV names 
sed_methanogen_asvs <- 
  sed_methanogen_asv_stats %>%
  dplyr::filter(mean > 1e-03) %>% # PULL ASVs with a mean of 0.0015 or 0.15% abundance or higher 
  pull(ASV) 

length(sed_methanogen_asvs) 

# Make a new df with only these ASVs 
sed_methanogen_abund_asvs_df <- 
  sed_methanogens_df %>%
  dplyr::filter(ASV %in% sed_methanogen_asvs)

# TIME TO PLOT 
top21_sed_methanogen_ASVS_plot <- 
  sed_methanogen_abund_asvs_df %>%
  mutate(ASV_Genus = paste0(ASV, "\n", Genus)) %>%
  group_by(JDate, Pond, solar_progress, Abundance, ASV) %>%
  ggplot(aes(x = JDate, y = Abundance, color = solar_progress)) + 
  facet_wrap(~ASV_Genus, scales = "free_y") +
  geom_line(aes(group = interaction(Pond, ASV)), 
            alpha = 0.2) +
  labs(x = "Day of Year", y = "Relative Abundance", 
       title = "Top 21 Sediment Methanogen ASVs") + 
  geom_point(aes(shape = Pond), size = 2) +
  scale_color_manual(values = solar_colors) +
  scale_shape_manual(values = pond_shapes) + 
  theme(legend.position = "bottom"); top21_sed_methanogen_ASVS_plot

# Save the plot   
ggsave(top21_sed_methanogen_ASVS_plot, 
       width = 12, height = 10, dpi = 300,
       filename = "figures/bonus/sed_methogen_ASV_time.png")
```


# Reproducibility
```{r reproducibility}
# Reproducibility
devtools::session_info()
```

